---
title: "{{< var course.short >}} - Course Syllabus"
filters:
    - list-table
---

```{r}
#| echo: false
#| message: false
library(tidyverse)
class_date <- function(n){
    read_csv("key_dates.csv") |>
        filter(`Course Element` == "Class Session") |>
        filter(`Item Number` == n) |>
        pull(Date) |>
        as.Date() |>
        format("%Y-%m-%d")
}
```

```{r}
#| echo: false
#| message: false
library(tidyverse)
holiday_date <- function(n, details=FALSE){
    h <- read_csv("key_dates.csv") |>
            filter(`Course Element` == "Holiday") |>
            filter(`Item Number` == n)
    
    if(details){
        h |> pull(Details)
    } else {
        h |> pull(Date) |> 
             as.Date()  |>
             format("%Y-%m-%d")
    }
}
```

```{r}
#| echo: false
#| message: false
library(tidyverse)
test_date <- function(n, details=FALSE){
    t <- read_csv("key_dates.csv") |>
            filter(`Course Element` == "Test") |>
            filter(`Item Number` == n)
    
    if(details){
        t |> pull(Details)
    } else {
        t |> pull(Date) |>
             as.Date() |>
             format("%Y-%m-%d")
    }
}
```


```{r}
#| echo: false
#| message: false
library(tidyverse)
report_date <- function(n, due=TRUE){
    r <- read_csv("key_dates.csv") |>
            filter(`Course Element` == "Research Report") |>
            filter(`Item Number` == n)
    
    if(due){
        r <- r |> filter(`Responsible Party` == "Students")
    } else {
        r <- r |> filter(`Responsible Party` == "Instructor")
    }
    
    r |> pull(Date) |>
         as.Date() |>
         format("%Y-%m-%d")
}
```

### Instructor

-   [Professor Michael Weylandt](https://michael-weylandt.com)
-   Department of Information Systems & Statistics
-   Zicklin School of Business
-   Baruch College, CUNY

### Course Meetings

#### Lecture / Lab Sessions

-   Tuesdays 6:05pm - 9:00pm
    -   [Baruch Main Campus (1 Bernard Baruch
         Way)](https://www.baruch.cuny.edu/campus-map-and-directions/)
    -   {{< var course.location >}}
    
#### Office Hours

-   In-Person
    -   [Baruch Main Campus (1 Bernard Baruch Way)](https://www.baruch.cuny.edu/campus-map-and-directions/)
    -   Newman Vertical Campus (NVC) 11-246
    -   Tuesdays 4:30-5:30pm
    -   Subject to periodic cancellation (communicated via Brightspace)
-   Virtual:
    -   Thursdays 4:30pm-5:30pm
    -   Zoom link provided via Brightspace

### Grading

{{< var course.short >}} is a(n advanced) master's-level course in Statistical
Machine Learning. As such, by the end of the course, students should be able to
capably apply standard machine learning techniques to important scientific and
business applications, to extend and adapt standard machine learning techniques
in novel fashions, to implement complex estimation and prediction algorithms,
and to critically engage with modern machine learning literature. The grading
in this course reflects this diversity of objectives.

Specifically, the grading in this course draws on three separate modes of
assessment: i) in-class 'midterms' which assess students' fluency with the 
theoretical and mathematical aspects of machine learning; ii) three 'research
reports' designed to build students' skills in the computational and 
methodological aspects of machine learning; and iii) a course prediction
competition, designed to build students' skills in applications of machine
learning to important scientific problems. All three elements are equally 
weighted as follows: 

- 33.3% Mid-Semester Tests (Best two of three; 100 points each, 200 points total)
  - Test 1: `r test_date(01)` - `r test_date(01, details=TRUE)`
  - Test 2: `r test_date(02)` - `r test_date(02, details=TRUE)`
  - Test 3: `r test_date(03)` - `r test_date(03, details=TRUE)`
- 33.3% Research Reports (Best two of three; 100 points each, 200 points total)
  - Report 1: `r report_date(01)`
  - Report 2: `r report_date(02)`
  - Report 3: `r report_date(03)`
- 33.3% Prediction Competition (200 points total)

Final course grades will be curved in accordance with relevant program,
departmental, school, and college policies.[^1]

[^1]: Theoretically, this may result in scores equivalent to an `A` in an
un-curved course receiving a lower grade in this course. In practice, the
instructor will design course assessments to induce a range of scores and does
not anticipate "down-curving" happening.

#### Weekly Quizzes

In lieu of homework, I will provide a short list of *practice problems* and
**suggested readings** following each lecture. At the start of the following
lecture, a short three question ($\approx$ 15 minute) quiz will be administered.
The quiz questions will not be **verbatim** from the practice problems, but if
you can answer the practice problems quickly and fluently, the quiz should pose
little difficulty.

These in-class weekly quizzes will generate *extra credit* applied to your
*final aggregate score*. Each quiz will receive a score out of 3 added
directly to your final score. Because the final aggregate score is out of
600, perfect scores on all 9 weekly quizzes can raise your final aggregate
score (*pre-curve*) **up to 5%**.

To take part in the weekly quizzes, please come to class with both i) a black
or blue pen; and ii) a red pen (for peer grading) each week.

Make-up opportunities for the weekly quizzes will only be allowed in
*exceptional* and *unforeseeable* circumstances.

#### Regrading Policy

If you feel an assignment has been improperly graded, please contact the instructor by private message on the course discussion board within 48 hours of the graded assignment being returned. Note that the instructor will regrade the assignment *de novo*, so your grade may be adjusted upwards or downwards.

#### Late Work Policy

Students are given 4 **late days** that can be used at any time during the
semester, without instructor pre-approval or permission, on any of the submitted
elements of this course (competition final report, research reports). When late
days are used, work is accepted without penalty; late work submitted beyond the
four late days will not be accepted. Late days are used on a "first come, first
serve" basis, so if 2 late days are used on the first two research reports, any
late submission on the final research report will not be accepted. 

Outside of the use of Late Days, no late work will be accepted. Late Days cannot
be applied to in-class activities such as tests or presentations. 

Individual exceptions will be made i) with *prior written* approval of
instructor; or ii) *ex post* with written note from the [Office of the
Dean of Students](https://studentaffairs.baruch.cuny.edu/dean-of-students/).

### Tentative Course Schedule

::: {.list-table width="1,1,6,1,1,2,1" aligns="c,c,l,l,l,l,l" header-rows=1 .hover}

   * - Week 
     - Lecture Date
     - Topics
     - Pre-Reading
     - Post-Reading
     - Additional Notes
     - Due Dates
     
   * - 1
     - `r class_date (01)`
     - Course Overview & Introduction to ML: 
       - Taxonomy of Learning
       - In-Sample and Out-of-Sample Accuracy
       - Complexity and Overfitting
       - Nearest Neighbor Methods
     - 
       
   * - 2
     - `r class_date (02)`
     - Regression I:
       - Review of OLS
       - Linear Algebra Review
       - Introduction to Convex Optimization
       
   * - 3
     - `r class_date (03)`
     - Regression II:
       - Regularization
       - Ridge Regression
       - Sparsity and the Lasso
       - Supervised Model Selection

   * - 
     - `r holiday_date(1)`
     <td colspan="5" style="text-align:center"> ***`r holiday_date(1, details=TRUE)`*** </td>

       
   * - 4
     - `r class_date (04)`
     - Regression III: 
       - Non-Linear Methods
       - Kernel Trick
       - Splines
       
   * - 5
     - `r class_date (05)`
     <td colspan="5" style="text-align:center"> **Mid-Term Test I: Regression** 
       <br> Introduction to Classification </td> 

   * - 6
     - `r class_date (06)`
     - Classification I:
       - Basics of Classification
       - Mixture Methods & Generative Methods

   * - 7
     - `r class_date (07)`
     - Classification II:
       - Discriminative Methods
       - Multi-Class Classification
       - Fairness in Machine Learning

   * - 8
     - `r class_date (08)`
     - Ensemble Learning & Resampling Methods
     
   * - 9
     - `r class_date (09)`
     - Tree-Based Methods
     
   * - 10
     - `r class_date (10)`
      <td colspan="5" style="text-align:center"> **Mid-Term Test II: Classification, Ensemble Learning, Tree-Based Methods** <br> Introduction to Unsupervised Learning </td> 

   * - 
     - `r holiday_date(2)`
     <td colspan="5" style="text-align:center"> ***`r holiday_date(2, details=TRUE)`*** </td>

   * - 11
     - `r class_date (11)`
     - Unsupervised Learning I: 
       - Dimension Reduction
       - Principal Components Analysis
       - Covariance Estimation
     

   * - 12
     - `r class_date (12)`
     - Unsupervised Learning II: 
       - Clustering
       - Density Estimation
       - Outlier Detection
       - Manifold Learning
       
   * - 13
     - `r class_date (13)`
     <td colspan="5" style="text-align:center"> **Mid-Term Test III: Unsupervised Learning**      <br> Introduction to Generative Models </td> 
     
   * - 14
     - `r class_date(14)`
     <td colspan="5" style="text-align:center"> **Course Project Presentations** <br> **Course Wrap-Up** </td>
::: 


**`All syllabus and course schedule provisions subject to change with suitable advance notice.`**


### Workload Expectations

The following _approximate_ breakdown of expected course workload is
intended to help you properly prepare for and schedule the out-of-class
work associated with this course. Note that, persuant to relevant
[Federal](https://www.ecfr.gov/current/title-34/part-600#p-600.2(Credit%20hour))
and [State](https://govt.westlaw.com/nycrr/Document/Ieca5c8abc22111dd97adcd755bda2840)
regulations, a 3-credit course taken over a 15 week semester should
require approximately 6 hours of out-of-class work from an _average_ 
student, or 90 hours total over the course of the semester.[^avg][^gc] This
time should be roughly allocated as: 

- Weekly Pre-Reading (9 hours - 1 hour per lecture)
- Weekly Post-Reading and Review (18 hours - 2 hours per lecture)
- Test Preparation (15 hours total - 5 hours per test)
- Research Reports (27 hours total - 9 hours each)
- Prediction Competition (21 hours total)
  - Ongoing Submissions (20 hours - approximately 2 per week)
  - Final Presentation Preparation (1 hour)
  
Note that, for the course prediction competition, you will need to make regular progress throughout the semester. If you attempt to "back-load" your work, you will do poorly. 

[^avg]: For this course, an _average_ student is a student who enters the
course with: 

    i) Fluency with statistical and numerical software at the level of
       (at least) STA 9750
    ii) Fluency with univariate and multivariate regression at the level of
        (at least) STA 9700
    iii) Familiarity with probability and linear algebra

    and is earning a B-range grade. If you have less background or are 
    aiming for a higher grade, you should expect to commit proportionally
    more time to this course. 

[^gc]: The CUNY Graduate Center has a useful [summary of these
expectations](https://gc.catalog.cuny.edu/general-information/credithourpolicy).
Baruch courses follow the same standards. 
