---
title: "{{< var course.short >}} - Course Syllabus"
filters:
    - list-table
---

```{r}
#| echo: false
#| message: false
#| warning: false
library(tidyverse)
class_date <- function(n, details=FALSE){
    c <- read_csv("key_dates.csv") |>
        filter(`Course Element` == "Class Session") |>
        filter(`Item Number` == n)
    
    if(details){
        c |> pull(Details)
    } else {
        c |> pull(Date) |>
             as.Date() |>
             format("%Y-%m-%d")
    }
}
```



```{r}
#| echo: false
#| message: false
library(tidyverse)
holiday_date <- function(n, details=FALSE){
    h <- read_csv("key_dates.csv") |>
            filter(`Course Element` == "Holiday") |>
            filter(`Item Number` == n)
    
    if(details){
        h |> pull(Details)
    } else {
        h |> pull(Date) |> 
             as.Date()  |>
             format("%Y-%m-%d")
    }
}
```

```{r}
#| echo: false
#| message: false
library(tidyverse)
test_date <- function(n, details=FALSE){
    t <- read_csv("key_dates.csv") |>
            filter(`Course Element` == "Test") |>
            filter(`Item Number` == n)
    
    if(details){
        t |> pull(Details)
    } else {
        t |> pull(Date) |>
             as.Date() |>
             format("%Y-%m-%d")
    }
}
```


```{r}
#| echo: false
#| message: false
library(tidyverse)
presentation_date <- function(){
    read_csv("key_dates.csv") |>
        filter(`Course Element` == "Prediction Competition") |>
        filter(str_detect(Details, "Presentations")) |>
        pull(Date) |>
        as.Date() |>
        format("%Y-%m-%d")
}
```

```{r}
#| echo: false
#| message: false
library(tidyverse)
voe_date <- function(){
    read_csv("key_dates.csv") |>
      filter(`Course Element` == "Syllabus Quiz") |>
      pull(Date) |>
      as.Date() |>
      format("%Y-%m-%d")
}
```

**`All syllabus and course schedule provisions subject to change with suitable advance notice.`**

### Instructor

-   [Professor Michael Weylandt](https://michael-weylandt.com)
-   Department of Information Systems & Statistics
-   Zicklin School of Business
-   Baruch College, CUNY

### Course Meetings

#### Lecture / Lab Sessions

-   Wednesdays 6:05pm - 9:00pm
    -   [Baruch Main Campus (1 Bernard Baruch
         Way)](https://www.baruch.cuny.edu/campus-map-and-directions/)
    -   {{< var course.location >}}
    
#### Office Hours

-   In-Person
    -   [Baruch Main Campus (1 Bernard Baruch Way)](https://www.baruch.cuny.edu/campus-map-and-directions/)
    -   Newman Vertical Campus (NVC) 11-246
    -   Wednesdays 5:00pm-5:45pm
    -   Subject to periodic cancellation (communicated via Brightspace)
-   Virtual:
    -   Thursdays 5:00pm-5:45pm
    -   Zoom link provided via Brightspace

### Grading

{{< var course.short >}} is an advanced master's-level course in Statistical
Machine Learning. As such, by the end of the course, students should be able to
capably apply standard machine learning techniques to important scientific and
business applications, to extend and adapt standard machine learning techniques
in novel fashions, to implement complex estimation and prediction algorithms,
and to critically engage with modern machine learning literature. The grading
in this course reflects this diversity of objectives.

Specifically, the grading in this course draws on two separate modes of
assessment: 

  i) in-class 'tests' which assess students' fluency with the 
     theoretical and mathematical aspects of machine learning; and
  ii) a course prediction competition, designed to build students' skills in
       applications of machine learning to important scientific problems. 
       
Both elements are equally weighted as follows: 

- 50% Ninety-Minute Mid-Semester Tests (Best two of three; 100 points each, 200 points total)
  - Test 1: `r test_date(01)` - `r test_date(01, details=TRUE)`
  - Test 2: `r test_date(02)` - `r test_date(02, details=TRUE)`
  - Test 3: `r test_date(03)` - `r test_date(03, details=TRUE)`
- 50% [Prediction Competition](./competition.html) (200 points total)

Final course grades will be curved in accordance with relevant program,
departmental, school, and college policies.[^1]

[^1]: Theoretically, this may result in scores equivalent to an `A` in an
un-curved course receiving a lower grade in this course. In practice, the
instructor will design course assessments to induce a range of scores and does
not anticipate "down-curving" happening.

#### Weekly Quizzes

In lieu of homework, I will provide a set of **suggested readings** following
each lecture. At the start of the following lecture, a short three question
($\approx$ 15 minute) quiz will be administered. The quiz questions will not be
**verbatim** from the reading, but if you can answer the textbook questions in
the suggested sections quickly and fluently, the quiz should pose little
difficulty.

These in-class weekly quizzes will generate *extra credit* applied to your
*final aggregate score*. Each quiz will receive a score out of 3 added
directly to your final score. Because the final aggregate score is out of
400, perfect scores on all 9 weekly quizzes can raise your final aggregate
score (*pre-curve*) **up to 7%**.

To take part in the weekly quizzes, please come to class with both i) a black
or blue pen; and ii) a red pen (for peer grading) each week.

Make-up opportunities for the weekly quizzes will only be allowed in
*exceptional* and *unforeseeable* circumstances.

#### Verification of Enrollment {#voe}

CUNY mandates a [Verification of
Enrollment (VoE)](https://enrollmentmanagement.baruch.cuny.edu/registrar/verification-of-enrollment/)
activity be completed for students to remain enrolled in a course. 
For {{< var course.short >}}, this VoE activity takes the form of an ungraded, 
but mandatory, Brightspace quiz. This quiz covers the contents of this syllabus.

Students who do not successfully complete this quiz by the required date 
(`r voe_date()`) may be involuntarily disenrolled from this course. 

You must complete the VoE Syllabus Quiz with a perfect score before you will
be able to submit any assignments on Brightspace. The Quiz is set up to allow
unlimited attempts. 

#### Regrading Policy

If you feel an assignment has been improperly graded, please contact the
instructor by private message on the course discussion board within 48 hours
of the graded assignment being returned. Note that the instructor will regrade
the assignment *de novo*, so your grade may be adjusted upwards or downwards.

#### Late Work Policy

Late work will not be accepted in this course. 

Additionally, note that: 

- Work **must be submitted on Brightspace** unless *specifically* otherwise
  stated in the written course description or approved *in writing* by the
  instructor *at least 48 hours* before the deadline. (If we discuss alterations
  in office hours or after class, you **must** follow up by email and get my 
  pre-approval in writing.) Email submissions are not allowed.[^email]
- Technology issues are not an approved excuse for late submissions; in 
  particular, Brightspace issues are not an approved excuse for late submission.
  Brightspace support is provided by
  [BCTC](https://bctc.baruch.cuny.edu/students/) and the [BCTC Help 
  Desk](https://bctc.baruch.cuny.edu/instructional-technology/services/service-now/). 
  The Help Desk is not staffed 24 hours per day, so you are encouraged to submit
  early and submit often to avoid issues. 

Individual exceptions will be made i) with *prior written* approval of
instructor; or ii) *ex post* with written note from the [Office of the
Dean of Students](https://studentaffairs.baruch.cuny.edu/dean-of-students/).

[^email]: Though ubiquitous, email is a remarkably 'flaky' service, providing
the sender no way to guarantee their message arrives untampered and providing
the recipient no way to guarantee the providence of a message received.
(This is not quite true: there are [tools for more secure 
email](https://gnupg.org) but they are somewhat more difficult to use and
are not supported at CUNY.)
Brightspace is integrated with CUNY's Identity Verification Services and
allows students to guarantee correct submission. Note that Brightspace does
not, by default, send students an email confirming submission, but I believe
this is an option that can be enabled on the student's end. 

#### Attendance Policy

Attendance is not required for this course and absences do not need
to be excused. Make-up opportunities for Quizzes and Tests missed while
absent **will not provided** except with i) with *prior written* approval
of instructor; or ii) *ex post* with written note from the [Office of
the Dean of Students](https://studentaffairs.baruch.cuny.edu/dean-of-students/).

### Pre- and Post-Reading Suggestions

Students learn material most effectively when exposed to it on multiple
occasions, ideally using alternative presentations strategies and formats.[^2]
To this end, suggested *pre-reading* and *post-reading* is provided for each
week of the course. Students are encouraged to *pre-read* the recommended text,
which typically presents that week's material in a less technical / more
intuitive manner, before each week's course session. Similarly, students are
encouraged to review the *post-reading* for each week after lecture to see
additional examples of topics covered. 

[^2]: Haoyu Chen and Jiongjiong Yang. "Multiple Exposures Enhance Both Item 
Memory and Contextual Memory over Time". *Frontiers in Psychology* **11**. 
November 2020. 
[DOI:10.3389/fpsyg.2020.565169](https://doi.org/10.3389/fpsyg.2020.565169)

While lectures will focus primarily on 'big picture' and 'major themes', the
recommended reading, especially the post-reading, provides additional coverage
of relevant technical detail. Note that pre- and post-reading are *entirely
optional* and will not be directly assessed. I have included (overlapping)
presentations from various text books to encourage students to find the style
and presentation that works best for them. 

### Tentative Course Schedule

::: {.list-table width="1,1,6,1,1,1" aligns="c,c,l,l,l,l" header-rows=1 .hover}

   * - Week 
     - Lecture Date
     - Topics
     - Pre-Reading
     - Post-Reading
     - Additional Notes
     
   * - 1
     - `r class_date (01)`
     - Course Overview & `r class_date (01, details=TRUE)`: 
       - Taxonomy of Learning
       - In-Sample and Out-of-Sample Accuracy
       - Complexity and Overfitting
       - Nearest Neighbor Methods
     - None
     - ISL §2.1, 3.5
     
       HR §1-3
   * - 2
     - `r class_date (02)`
     - `r class_date (02, details=TRUE)`: 
       - Review of OLS
       - Linear Algebra Review
       - Singular Value Decomposition
     - ISL §3
     
       DFO §2.1-2.6, §3.1-3.4, §4.2, §4.4-4.5 , §5.1-5.2, §5.7-5.8, 
     - DFO §9
     
       HR §4, 14
   * - 3
     - `r class_date (03)`
     - `r class_date (03, details=TRUE)`: 
       - Accuracy, Loss, Error, and Risk
       - Hold-Out and Cross Validation
       - Metrical Determinism and Statistical Evaluation
       - Bias and Variance
       - Tradeoffs *vs.* Decompositions
       - Metrics for Classification
     - ISL §2.2, 5
     
       DFO §8.1-8.3
     - HR §6
     - [VOE Syllabus Quiz](#voe) due on Brightspace **`r voe_date()`**.  

   * - 4
     - `r class_date (04)`
     - `r class_date (04, details=TRUE)`: 
       - Introduction to Convex Optimization
       - Simulation as a Means of Method Evaluation
       - Limitations of Simulation
     - DFO §7.1-7.3
       
     - HR §5, 8
     
       BV §2-4

   * - 5
     - `r class_date (05)`
       <td colspan="3" style="text-align:center"> 
         **`r test_date(1, details=TRUE)`** <br> 
         `r class_date (5, details=TRUE)`
       </td>

   * - 6
     - `r class_date (06)`
     - `r class_date (06, details=TRUE)`: 
       - Ridge Regression & $\ell_2$ Regularization
       - Sparsity
       - $\ell_1$-Penalization
       - $\ell_2$ vs $\ell_1$
       - Generalizations of the Lasso
     - ISL §6
     
     - SLS §2, 4.1-4.3, 5.1-5.4
     
       BV §6-7.1

   * - 7
     - `r class_date (07)`
     - `r class_date (07, details=TRUE)`: 
       - Basics of Classification
       - Mixture Methods & Generative Methods
       - Multi-Class Classification
       - Fairness in Machine Learning
     - ISL §4
     
       DFO §11
     - SLS §3
     
       BHN §1, 3, 4
       
   * - 8
     - `r class_date (08)`
     - `r class_date (08, details=TRUE)`:
       - Discriminative Methods
       - Non-Linear Methods & Feature Expansion
       - Kernel Trick
       - Splines
     - ISL §9, 7
     
       DFO §12
     - SLS §4.4-4.5
     
       ["Generalized Additive Models" *ARSIA* 12](https://doi.org/10.1146/annurev-statistics-112723-034249)
       
       ["GAMs in R" by Noam Ross](https://noamross.github.io/gams-in-r-course/)
   * - 9
     - `r class_date (09)`
     - `r class_date (09, details=TRUE)`:
       - Decision Trees
       - Model Stacking
       - Bagging
       - Boosting
       - Mini-Patching
       - Random Forests
     - ISL §8
     - SF §1-3, 5
   * -
     <td colspan="5" style="text-align:center"> 
       ***`r holiday_date(1, details=TRUE)`*** 
     </td>

   * -
     <td colspan="5" style="text-align:center"> 
       ***`r holiday_date(2, details=TRUE)`*** 
     </td>
     
   * - 10
     - `r class_date (10)`
       <td colspan="4" style="text-align:center"> 
         **`r test_date(2, details=TRUE)`** <br> 
         `r class_date (10, details=TRUE)`
       </td>

   * - 11
     - `r class_date (11)`
     - `r class_date (11, details=TRUE)`:
       - Hierarchical Clustering
       - $K$-Means and EM Clustering
       - Outlier Detection
       - Density Based Methods
       
     - ISL §12.1-12.3
     - SLS §8.5

   * - 12
     - `r class_date (12)`
     - `r class_date (12, details=TRUE)`:
       - Principal Components Analysis
       - NMF
       - Sparse PCA
       - Manifold Learning
       
     - ISL §12.2
     
       DFO §3.5, §3.8, §4.2-4.6, §10
     
     - SLS §7, 8.4
       
   * - 13
     - `r class_date (13)`
     - `r class_date (13, details=TRUE)`.
       - $t$-SNE
       - UMAP
       - Variational Auto-Encoders
     - TBD
     - TBD
     
   * - 14
     - `r class_date(14)`
       <td colspan="3" style="text-align:center"> 
         **`r test_date(3, details=TRUE)`** <br> 
         `r class_date (14, details=TRUE)` 
       </td>
      
   * - Final Exam Period
     - `r presentation_date()`
       <td colspan="3" style="text-align:center"> 
         **Course Project Presentations** <br> 
         *[Tentative pending Registrar's Exam Schedule]*
       </td>
::: 




### Workload Expectations

The following _approximate_ breakdown of expected course workload is
intended to help you properly prepare for and schedule the out-of-class
work associated with this course. Note that, persuant to relevant
[Federal](https://www.ecfr.gov/current/title-34/part-600#p-600.2(Credit%20hour))
and [State](https://govt.westlaw.com/nycrr/Document/Ieca5c8abc22111dd97adcd755bda2840)
regulations, a 3-credit course taken over a 15 week semester should
require approximately 6 hours of out-of-class work from an _average_ 
student, or 90 hours total over the course of the semester.[^avg][^gc] This
time should be roughly allocated as: 

- Weekly Pre-Reading (10 hours - 1 hour per lecture on non-test days)
- Weekly Post-Reading and Review (26 hours - 2 hours per lecture)
- Test Preparation (30 hours total - 10 hours per test)
- Prediction Competition (24 hours total)
  - Ongoing Submissions (18 hours - approximately 2 hours per week for 9 weeks)
  - Final Presentation (3 hours)
  - Final Report (3 hours)
  
Note that, for the course prediction competition, you will need to make
regular progress throughout the semester. If you attempt to "back-load" your
work, you will do poorly. 

[^avg]: For this course, an _average_ student is a student who enters the
course with: 

    i) Fluency with statistical and numerical software at the level of
       (at least) STA 9750; 
    ii) Fluency with univariate and multivariate regression at the level of
        (at least) STA 9700; and
    iii) Familiarity with probability and linear algebra; 

    <br>and is earning a B-range grade. If you have less background or are 
    aiming for a higher grade, you should expect to commit proportionally
    more time to this course. 
    
    If you lack the prerequisite background listed above or
    simply wish to review it before the semester begins in
    earnest, please reach out to the instructor and I will be
    more than happy to provide supplementary readings. 

[^gc]: The CUNY Graduate Center has a useful [summary of these
expectations](https://gc.catalog.cuny.edu/general-information/credithourpolicy).
Baruch courses follow the same standards. See also [CUNY Central
Policy](https://www.cuny.edu/wp-content/uploads/sites/4/page-assets/about/administration/offices/registrar/resources/State-and-Federal-Credit-Hour-Definitions-08022019-2-1.pdf).

### Coding Requirements

{{< var course.short >}} is, at its heart, a **machine learning** course and as
such use of a machine, *i.e.* **coding**, is required, even though there is no
formal coding prerequisite for this course. Per the [External Resources 
Policy](./resources.html#externalresources), you are allowed (and encouraged) to
use freely available coding assistance technologies, including generative tools
like GitHub CoPilot.[^copilot] 

You may use whatever programming language you prefer to complete the course
assignments, subject to instructor approval: `R`, `python`, `julia`, and `matlab`
are pre-approved for all students by default. (I will approve most other
languages as well, provided they are not *too* obscure.) 

Tools like [`quarto`](https://quarto.org/) or [`Jupyter 
Notebooks`](https://jupyter.org/) will be useful for completing the 
[Course Competition](./competition.html) and associated final report. 
If you have not used these previously, many useful free resources can be found
online, including my [`STA9750`](https://michael-weylandt.com/STA9750/) course
materials.

#### `PyTorch`, `TensorFlow`, and Similar

An exciting aspect of modern ML is the widespread availability of flexible
and powerful model fitting frameworks like
[`TensorFlow`](https://www.tensorflow.org/) and [`PyTorch`](https://pytorch.org/).
Use of these frameworks is not required for this course, though students 
interested in ML focused careers would benefit from familiarizing themselves 
with them. 

For the [Prediction Competition](./competition.html), you may use these
frameworks or indeed any freely available (F/OSS) statistical or ML software;
instructor pre-approval is not required. 

[^copilot]: As a student, you have free access to [GitHub
CoPilot](https://github.com/features/copilot) once you create a student GitHub
account and register for the [Student Developer
Pack](https://education.github.com/pack). 

**`All syllabus and course schedule provisions subject to change with suitable advance notice.`**
