---
session: "`r session <- 12; session`"
topic: "`r library(tidyverse); topic <- read_csv('../key_dates.csv', name_repair='universal') |> filter(Course.Element == 'Class Session', Item.Number==session) |> pull(Details); topic`"
date: "Last Updated `r Sys.Date()`"
---

{{< include _latex_defines.qmd >}}

Today, we begin our study of *dimension reduction*, the task of finding
a simplified representation of your data. Before we begin formalizing
DR, it is worthwhile to take a second to think about _why_ we may want
to perform DR as this will help us design suitable loss functions. 

For now, let's take DR 'literally' and assume that we're trying to 
find a _lower-dimensional_ representation: that is, given data points
$\bx_i \in \R^P$, we want to find $\bx_i \in \R^p$ for some $p \ll P$.[^man] Typically, $P$ may be $\Ocal(1000-10000)$ and $p$
will be $\Ocal(1-10)$. To identify the goals of DR, let's think
about the problems associated with large $P$: 

- Difficult data exploration and visualization
  - Our standard tools, like a scatter plot or a 'pairs plot', simply
    aren't feasible for large dimensions. Even if we lean into 
    tools like `ggplot`, we really can only plot 5-6 dimensions at a
    time ($x$, $y$, size, color, shape, 2 facets)
- Interpretation
  - We can't really make sense of data with several thousand features.
    There will be significant correlation among features (even if
    just 'by chance') and finding 'important' features as opposed to
    'noise' features becomes essentially impossible (even ignoring the
    definition issues)
- Supervised Learning
  - Building supervised models with $P$ features is harder than building
    models with $p$ features, particuarly if we are prioritizing
    interpretability. While we have developed techniques that prioritize
    interpretation in some methods (lasso and variants for linear
    regression), these are somewhat special purpose and can't be applied
    to arbitrary learning methods. 
    
[^man]: Later, we will use a more generalized notion of 'dimension',
leading us to the task of _manifold learning_.

Clearly, in each of these, we're going to have an easier time if we only
have $p$ features. And we could get $p$ features quite easily by simply
'throwing away' most of our variables... But this has a rather obvious
problem: if we throw away features willy-nilly, we might throw away the
important 'signal' components of our data. So we want to build a model
for automatically mapping from $\R^P$ to $\R^p$ that keeps the 'signal'
(as best as we can estimate it). 

Of course, this requires us to make the concept of 'signal'
mathematically precise. We might consider goals like the following: 

- 'Distance preservation': if $\bx, \by \in \R^P$ are far apart, we 
  want $f(\bx), f(\by) \in \R^p$ to be far aapart as well. Similarly,
  if two points $\bx, \by \in \R^P$ are close, we want their DR'd points
  $f(\bx), f(\by)$ to be close in the reduced space.
- 'Angle preservation': if $\bx, \by \in \R^p$ are at an angle of 
  approximately $\theta$, we want the angle between $f(\bx)$ and
  $f(\by)$ to be similar. In particular, if $\bx, \by$ are orthogonal, 
  we want $f(\bx), f(\by)$ to be orthogonal as well. 
  
Mathematically, it will suffice to maintain inner products, as inner
products generate distances and angles: that is, we want

$$\langle \bx, \by \rangle_{\R^P} \approx \langle f(\bx), f(\by) \rangle_{\R^p}$$

If you prefer a more statistical (as opposed to geometric) motivation, 
you can think of our desiderata as maintaining variances (distances) and
correlations (angles) among aspects of our data: clearly, it suffices
to maintain covariances (inner products). 

## PCA: First Steps

So, with that goal in mind, let's start building our first dimension
reduction technique. Consistent with our treatment of unsupervised
learning, we'll begin by assuming $f$ is a linear operation and living
in the world of MSE. As we build our method, let's start by taking $p=1$
(reducing down to a single dimension) but we'll relax this in a bit. 

Our reduced dimension space can be parameterized by a single vector 
$\bv \in \R^P$, with the reduced space being defined by $$\textsf{Span}(\bv) = \{\lambda \bv: \lambda \in \R\}.$$ That is,
our reduced space is the line generated by $\bv$: 

```{r}
#| echo: false
v <- c(1, 1) / sqrt(2)

plot(c(3, -3, 3, -3), c(3, 3, -3, -3), type="n", 
     asp=1, xlab=expression(x[1]), ylab=expression(x[2]))
arrows(0, 0, v[1], v[1], lwd=3)
lines(x=c(-3, 3), y=c(-3, 3), col="red4", lwd=1, lty=2)
```

Here the black arrow is $\bv$ and the dashed red line is $\textsf{Span}(\bv)$. For a given space, the choice of $\bv$ is clearly
not unique: if we streteched or even flipped $\bv$,
$\textsf{Span}(\bv)$ would remain unchanged, so we might as well
assume $\|\bv\|=1$ without loss of generality. 

Once we have $\bv$, how can we actually build the mapping $f(\cdot): \R^P \to \textsf{Span}(\bv)$? Well, if our goal is to minimize MSE, it
makes sense to find the point in $\textsf{Span}(\bv)$ that lies closest
to our original point. It's hopefully not too hard to convince yourself
that this point is given by the projection of $\bx$ onto $\textsf{Span}(\bv)$, which is given by 

$$\bv \frac{\langle \bx, \bv \rangle}{\langle \bv, \bv\rangle} = \langle \bx, \bv\rangle \bv = \bx^{\top}\bv \bv$$

Visually: 

```{r}
#| echo: false
v <- c(1, 1) / sqrt(2)

plot(c(3, -3, 3, -3), c(3, 3, -3, -3), type="n", 
     asp=1, xlab=expression(x[1]), ylab=expression(x[2]))
arrows(0, 0, v[1], v[1], lwd=3)
lines(x=c(-3, 3), y=c(-3, 3), col="red4", lwd=1, lty=2)

x <- c(2, 2.75)
proj_coef <- sum(x * v)
x_proj <- v * proj_coef

points(x[1], x[2], col="blue4", cex=1.2, pch=16)

points(x_proj[1], x_proj[2], col="blue4", cex=1, pch=18)
segments(x[1], x[2], x_proj[1], x_proj[2], col="blue4", lty=2)
```

Here, we have projected the original point (big blue) to the nearest
point on $\textsf{Span}(\bv)$ (red dashed line), resulting in the DR'd 
smaller blue point. The 'loss' of this projection is measured by the
Euclidean distance between the two points, here captured by the dashed
blue line. 

We note here that the relevant distance here is the perpendicular
(straight line) distance from the original point to the target space.
We _do not_ make the $y$-axis special, unlike supervised learning where
there is a 'special' response variable. 

With this in hand, we are now ready to define the loss function we
use to select $v$: 

$$\argmin_{\bv: \|\bv\|==1} \left\|\bx - \textsf{proj}_{\textsf{Span}(\bv)}(\bx)\right\|_2^2 = \argmin_{\bv: \|\bv\|==1} \left\|\bx - \bv \langle \bx, bv\rangle\right\|_2^2 = \argmin_{\bv: \|\bv\|==1} \left\|\bx - \bx^{\top}\bv \bv\right\|_2^2 $$

With many training points, our problem generalizes to the sum of total
errors, giving us: 

$$ \argmin_{\bv: \|\bv\|==1} \sum_i \left\|\bx_i - \bx^{\top}_i\bv \bv\right\|_2^2 = \argmin_{\bv: \|\bv\|==1} \left\|\bX - \bX\bv\bv^{\top}\right\|_F^2 = \argmin_{\bv: \|\bv\|==1} \left\|\bX(\bI - \bv\bv^{\top})\right\|_F^2$$

Looking at this, we see a certain similarity to OLS: we are using $\bX$ to predict the projected $\bX$ in a minimum MSE sense. But,
unlike standard supervised learning where the response is fixed, here
we get to _select_ the response (indirectly by selecting $\bv$) that
makes the problem as easy as possible (*i.e.*, we get to pick the 
response so that the final residuals are as small as possible). 

In this sense DR is a _self-supervised_ problem: we are finding a
projection that is the most predictable and looses the least amount of
variance in $\bX$. This actually ties unsupervised learning to
supervised learning nicely: in supervised learning, we want to find
reliable accurate predictions. In unsupervised learning, we don't have
a sense of prediction and are instead settling for finding a 'repeatable' pattern. But, if the pattern we find corresponds to a
reliable and accurate prediction, it's not unreasonable to assume
it will generalize out-of-sample in some sense.

So how can we actually solve this problem? It's not as clean as OLS
since $\bv$ gets 'squared' inside the loss (giving a fourth-degree
polynomial in $\bv$), but it will turn out to be quite solvable
anyways. If we expand our our loss, we get the following: 

$$\begin{align*}
 \left\|\bX - \bX\bv\bv^{\top}\right\|_F^2 &= \left\|\bX\right\|_F^2 - 2 \langle \bX, \bX\bv\bv^{\top}\rangle + \left\|\bX\bv\bv^{\top}\right\|_F^2\\
 &= \left\|\bX\right\|_F^2 - 2 \langle \bX, \bX\bv\bv^{\top}\rangle + \langle\bX\bv\bv^{\top}, \bX\bv\bv^{\top}\rangle\\
 &= \left\|\bX\right\|_F^2 - 2 \Tr(\bX^{\top}\bX\bv\bv^{\top})+ \Tr((\bX\bv\bv^{\top})^{\top}(\bX\bv\bv^{\top})) \\
 &= \left\|\bX\right\|_F^2 - 2 \Tr(\bv^{\top}\bX^{\top}\bX\bv)+ \Tr(\bv\bv^{\top}\bX^{\top}\bX\bv\bv^{\top}) \\
 &= \left\|\bX\right\|_F^2 - 2 \Tr(\bv^{\top}\bX^{\top}\bX\bv)+ \Tr(\bv^{\top}\bX^{\top}\bX\bv\underbrace{\bv^{\top}\bv}_{=\|\bv\|^2=1}) \\
 &= \left\|\bX\right\|_F^2 - 2 \Tr(\bv^{\top}\bX^{\top}\bX\bv)+ \Tr(\bv^{\top}\bX^{\top}\bX\bv) \\
 &= \left\|\bX\right\|_F^2 - \Tr(\bv^{\top}\bX^{\top}\bX\bv) \\
 &= \left\|\bX\right\|_F^2 - \bv^{\top}\bX^{\top}\bX\bv
\end{align*}$$
Because $\|\bX\|_F^2$ doesn't change as a function of $\bv$, this means our problem 
can be reduced to 

$$\argmin_{\bv: \|\bv\|=1} - \bv^{\top}\bX^{\top}\bX\bv = \argmax_{\bv: \|\bv\|=1} \bv^{\top}\bX^{\top}\bX\bv$$

We see here that the $\|\bv\|=1$ constraint is important: if this constraint was
missing, we could simply make the objective arbitrarily large by increasing $\bv$
without limit. 

This gives us our second interpretation of our DR problem: 

1) We want to minimize the loss in projecting $\bX$ into $\textsf{Span}(\bv)$
2) We want to _maximize_ the amount of $\bX^{\top}\bX$ captured in the $\bv$ 
   direction. 

At this point, it is perhaps with noting that the DR problem we have developed
(in whatever form you prefer) is *Principal Components Analysis* (PCA). 

## Solving the PCA Problem with Linear Matrix Decompositions

So how can we solve the PCA problem?

$$\argmax_{\bv: \|\bv\|=1} \bv^{\top}\bX^{\top}\bX\bv$$

While you might be tempted to use the tools of convex optimization here, this
isn't actually a convex problem. In particular, we're looking to _maximize_, while
convex optimization normally lets us minimize. So we need to return to first principles: 

We might first consider using a gradient _ascent_ method, modifying it to maintain
our $\|\bv\|=1$ constraint. In particular, if we differentiate $\bv^{\top}\bX^{\top}\bX\bv$
with respect to $\bv$, we get $2\bX^{\top}\bX\bv$, so an iteration like

$$ \bv^{(k+1)} \leftarrow \bv^{(k)} + c 2\bX^{\top}\bX\bv^{(k)} = (\bI + 2c\bX^{\top}\bX)\bv^{(k)}$$

might make sense. Rather than selecting the step size $c$ directly, we can 
instead assume that we want to as far as we can without violating the $\|\bv\|=1$ 
constraint and take our update to be: 

$$ \bv^{(k+1)} \leftarrow \frac{(\bI + 2c\bX^{\top}\bX)\bv^{(k)}}{\|(\bI + 2c\bX^{\top}\bX)\bv^{(k)}\|}$$
Note that the division here ensures $\|\bv^{(k+1)}\}=1$ for all $k$. Letting $c$
be as large as possible, we can ignore that $\bI$ term and get the simpler update: 

$$ \bv^{(k+1)} \leftarrow \frac{\bX^{\top}\bX\bv^{(k)}}{\|\bX^{\top}\bX\bv^{(k)}\|}$$
Because we get a result of the form $\bv^{(k+1)} = (\bX^{\top}\bX)^{k}\bv^{(0)}$,
this algorithm is sometimes known as the _power method_. We do not yet quite
have the tools to prove that it converges to the right answer, but it usually
does and is rather robust (if sometimes slow). The power method has the advantage
of only requiring us to multiply by $\bX^{\top}\bX$, so if $\bX$ is easy to 
multiply by (*e.g.*, if it is mainly 0s), our algorithm with automatically inherit
any optimizations we can apply in the multiplication step. 

This is an example of an _iterative_ algorithm for PCA. It turns out that direct
methods are also available using (yet another) characterization of PCA. 
In particular, you might recognize $\bv$ as an _eigenvalue_ of $\bX^{\top}\bX$. 
If this isn't immediately apparent, read on: note that, at convergence, $\bv$
has to be a _fixed point_ of the power iteration. That is, 

$$\bv = \frac{\bX^{\top}{\bX}\bv}{\|\bX^{\top}{\bX}\bv\|}$$
(If this equation weren't satisfied, the power method would encourage us to take
another step.) If we let $\lambda = \|\bX^{\top}\bX\bv\|$, then this gives
us the _stationary condition_:

$$\lambda \bv = \bX^{\top}{\bX}\bv$$

Note you can also get this by setting our gradient equal to zero in the original
derivation to the power method, hence the (calculus-sounding) 'stationary condition'
name.

This type of equation: 

$$\textsf{scalar} * \text{vector} = \textsf{matrix} * \textsf{same vector}$$

is incredibly important in mathematics, where it is used to define the _eigenvectors_
of a matrix. In particular, any vector that satisfies an equation of the above
form is known as an _eigenvector_ and the associated scalar is known as an
_eigenvector_. Hence, PCA consists of finding the (leading) eigenvector-eigenvalue
pair of the matrix $\bX^{\top}\bX$.[^leading] 


[^leading]: A matrix of size $p \times p$ can have up to $p$ eigenvalue-eigenvector
pairs. Because we are _maximizing_ $\bv^{\top}\bX^{\top}\bX\bv$, we want the maximal
(or leading) eigenvector-eigenvalue pair, defined as the one for which the 
(absolute value of) the eigenvalue is the largest. 

Eigenvectors are very well-studied and so efficient algorithms exist for finding
them in almost every programming environment. In `R`, the `eigen()` function has
a high-performance general purpose eigenvalue/eigenvector solver. Let's see it
in action and compare it to the power method. 

```{r}
n <- 500
p <- 250
X <- matrix(rnorm(n * p), nrow=n)
XtX <- crossprod(X)

## Power method
normalize <- function(v) v / sqrt(sum(v^2))
v <- matrix(rnorm(p), ncol=1); v <- normalize(v)

system.time({
    repeat {
        v_old <- v
        v <- normalize(XtX %*% v)
        
        if(sum(abs(v - v_old)) < 1e-10) break
    }
})

v_pm <- v
lambda_pm <- mean((XtX %*% v_pm) / v_pm)

## eigen()
system.time({eigen_decomp <- eigen(XtX, symmetric=TRUE)})
```

So here the optimized algorithm is about 10x faster. But it's actually much more
impressive than that. If you look at `eigen_decomp`, you'll see it actually computed
all 250 eigenpairs, not just the one we are interested in, so it's really more
like 2500 times faster. 

To wit, 

```{r}
all.equal(eigen_decomp$value[1], 
          lambda_pm)
```

```{r}
all.equal(eigen_decomp$vectors[,1,drop=FALSE], +v_pm)
all.equal(eigen_decomp$vectors[,1,drop=FALSE], -v_pm)
```

Here we compare against both $\pm \bv$ since the eigenvalue is only defined up
to sign (recall from our picture above where we can 'flip' the seed direction
but the span is unchanged) but `R` isn't smart enough to know the difference
so we have to help it out a bit. 

For small to medium problems, it is _usually_ a good idea to use built-in
algorithms like `eigen` or (as we will see below) `svd`. For _very large_ problems,
it may be better to use the power method (or similar) as it only computes the
first eigenvector. As we will see below, the power method is also quite easy to 
extend if we want to introduce sparsity or similar forms of regularization. 

## Additional Principal Components

So far, we have done _extreme_ dimension reduction all the way from $\R^P$ down
to $\R^1$. We have found that the eigenvectors of $\bX^{\top}\bX$ give us a way
to do this _optimally_, but that's still a huge (and unavoidable) loss of information.
What if we want to reduce only to $\R^2$ or $\R^3$ (or generally $\R^p$ for 
$P \gg p > 1$)? We expect that this additional dimension (or 'additional degree
of freedom') will help us, but by how much? And how can we best take advantage
of it? 

To go down this path, we need to look at one form of our decomposition from above: 

$$\argmin_{\bv: \|\bv\|=1} \|\bX(\bI - \bv\bv^{\top})\|_F^2$$

This quantity -- $\bR = \bX(\bI - \bv\bv^{\top})$ -- plays the role of _residuals_ for
PCA; it is all of the data 'left-over' after we project onto $\bv$. By definition,
there will be no $\bv$ left in this residual, as we can see: 

$$\bR\bv = \bX(\bI - \bv\bv^{\top})\bv = \bX(\bv - \bv\bv^{\top}\bv) = \bX(\bv - \bv) = \bzero$$

This suggests to us that, if we want to use our new second dimension smartly, it
will not have anything to do with $\bv$. In particular, we might expect it to be
_orthogonal_ to $\bv$ (and it turns out - we would be right!). 

We can start with a rather naive approach: what if we just get "the best of the rest"
and perform PCA again on $\bR$. The resulting vector $\bv_2$ will capture something
in $\bR$ (hopefully a large part of it) and, because $\bR$ has only variance
orthogonal to $\bv_1$, it will add something new to our decomposition. 

Formally, this approach is known as _deflation_: we perform PCA first on $\bX$, 
then on $\bR_1 = \bX(\bI - \bv_1\bv_1^{\top})$, then on $\bR_2 = \bR_1(\bI - \bv_2\bv_2^{\top})$, 
*etc.* The coefficients in the resulting space are given by 

$$(\bX\bv_1, \bX\bv_2, \dots)$$

We can see this in action visually: 

```{r}
#| echo: false
conflicted::conflict_prefer("layout", "plotly")
```

```{r}
#| warning: false
#| message: false
library(plotly)

S <- matrix(c(16, -8, -2, 
              -8, 22, 10, 
              -2, 10, 25), nrow=3, byrow=TRUE)

library(MASS)

X <- mvrnorm(5000, Sigma=S, mu=c(0, 0, 0))

plot_ly(x=X[,1], y=X[,2], z=X[,3], 
        type="scatter3d", mode="markers") |>
    layout(title="Original 3D Data Set")
```

Here, we have generated $n=5000$ points in $\R^3$. Because our feature are 
not uncorrelated, the resulting point cloud is not quite spherical. 
If we perform PCA on this data, the one dimensional projection of our results
is rather uninteresting: 

```{r}
pca_power_method <- function(X, thresh=1e-8){
    XtX <- crossprod(X)
    p <- NCOL(XtX)
    normalize <- function(x) x / sqrt(sum(x^2))
    
    v <- normalize(matrix(rnorm(p), ncol=1))
    
    repeat {
        v_old <- v
        v <- normalize(XtX %*% v)
        
        # To fix sign irregularity, force v[1] > 0
        if(v[1] < 0) v <- -v
        
        if(sum(abs(v - v_old)) < thresh){
            break
        }
    }
    
    list(v=v, lambda=sqrt(sum((XtX %*% v)^2)))
}

PC1 <- pca_power_method(X)

X_pc1 <- X %*% PC1$v

plot(X_pc1)
```

As we might expect for this IID data, there's no meaningful structure here. 
The residuals on the other hand ($\bR_1$) are somewhat interesting: 

```{r}
eye <- function(n) diag(1, n, n)
R1 <- X %*% (eye(3) - tcrossprod(PC1$v))

plot_ly(x=R1[,1], y=R1[,2], z=R1[,3], 
        type="scatter3d", mode="markers") |>
    layout(title="Residuals after removing 1 PC")
```

If you rotate this data in just the right way, you will see that it is _flat_
along one dimension. And if you look even closer, you will see that the flat 
dimension is _exactly_ $\bv_1$ - that is, the direction we first found in PCA.
This makes sense - we "pulled out" $\bv_1$ in forming $\bR_1$, so there is no
$\bv_1$ left in our residuals. 

At this point, we can benefit from introducing another important linear algebra
concept: _rank_. The _rank_ of a matrix (*i.e.*, of a point cloud) is the 
number of dimensions it 'actually' occupies. Our original data was rank-3, since
it used all three dimensions of our data space, but the residuals $\bR_1$ are 
only rank-2, since we removed one dimension. You can think of rank as the 
"intrinsic dimension" associated with data. 

If we repeat this process again, we go down to a rank-1 set of residuals: 

```{r}
v1 <- matrix(c(-1/3, 2/3, 2/3), ncol=1)

X_projection1 <- X %*% v1

X_residuals <- X - X %*% v1 %*% t(v1)

plot_ly(x=X_residuals[,1], 
        y=X_residuals[,2], 
        z=X_residuals[,3], 
        type="scatter3d", mode="markers")

v2 <- matrix(c(2/3, -1/3, 2/3), ncol=1)

eye <- function(n) diag(1, n, n)

X_residuals_projection1 <- X_residuals %*% v2

X_resid_resid <- X_residuals - X_residuals %*% v2 %*% t(v2)

# Project again
# X_residuals %*% (eye(NCOL(X)) - v1 %*% t(v1) - v2 %*% t(v2))

plot_ly(x=X_resid_resid[,1], 
        y=X_resid_resid[,2], 
        z=X_resid_resid[,3], 
        type="scatter3d", mode="markers")
```

And, while it's a bit unnatural, if we were to go one step further, we would
be left with residuals that were all 0, so our residuals would live on a 
'zero dimensional' point at the origin. 

This 'dimension-reduction' is often cited as a key advantage of PCA, but it's
worth thinking about what it does and does not mean for us: 

- 
- 
- 
- 


## PCA Patterns in Observations

So far, we've focused on the patterns in among the features identified by 
PCA, but just as often, we are interested in patterns in 'observation space'. 
We find an interesting pattern, but then we want to ask whether it is trending
upwards or downwards or if any points are very high or very low on than pattern.

We've seen above that we can do this by multiplying each row of $\bX$ and seeing
by our PC $\bv$. This gave us the `X_pc1` variable above:

```{r}
plot(X_pc1)
```

This was a situation where the rows were IID, so the 'fuzzy caterpillar' was 
not surprising. But let's consider a pattern where there is structure to the
rows, *e.g.*, a multivariate time series. 

Here I'm going to use the `OpenMeteo` website to pull temperature data from
NYC, Boston, Philadelphia, Baltimore, and Washington DC for the past 6 months:

```{r}
library(tidyverse)
library(httr2)
eps <- 0.2
cities <- tribble(
    ~city, ~lat, ~lon, 
    "NYC", 40.73, -73.93,
    "BOS", 42.36, -71.06,
    "PHL", 39.96, -75.17,
    "BAL", 39.29, -76.61,
    "WDC", 38.89, -77.03
) |> 
    mutate(lat_p = lat + eps, 
           lat_n = lat - eps, 
           lon_p = lon + eps, 
           lon_n = lon - eps)

resp <- request("https://api.open-meteo.com") |>
    req_url_path("v1/forecast") |>
    req_url_query(latitude=paste(cities$lat, collapse=","), 
                  longitude=paste(cities$lon, collapse=","),
                  hourly="temperature_2m", 
                  past_days=60) |>
    req_perform()

temperature_df <- resp |> 
    resp_body_json() |> 
    map(\(r) data.frame(time = r |> pluck("hourly", "time") |> list_c(), 
                        temperature_2m = r |> pluck("hourly", "temperature_2m") |> list_c(), 
                        latitude = r |> pluck("latitude"), 
                        longitude = r |> pluck("longitude"))) |> 
    list_rbind() |> 
    mutate(time=ymd_hm(time)) |> 
    left_join(cities, join_by(between(latitude, lat_n, lat_p), 
                              between(longitude, lon_n, lon_p))) |> 
    select(time, temperature_2m, city) |>
    pivot_wider(names_from=city, 
                values_from=temperature_2m) |>
    # Remove baseline temperature
    mutate(across(-time, \(x) x - mean(x))) 

glimpse(temperature_df)
```

which gives us data like the following (except with `r NROW(temperature_df)` rows):

```{r}
#| echo: false
library(gt)
temperature_df |> 
  slice_head(n=5) |>
  gt(id="tbl_temperatures", 
     rowname_col="time") 
```

If we plot this, we can see a clear common temporal pattern with
the day-night cycle:

```{r}
#| cache: true
library(GGally)
ggpairs(temperature_df)
```

So let's now apply PCA to the temperature columns in this
data set: 

```{r}
#| cache: true
temperature_mat <- temperature_df |> 
  select(-time) |> 
  as.matrix()

temperature_pc1 <- pca_power_method(temperature_mat)

temperature_pc1 |> 
    pluck("v") |> 
    as.data.frame() |> 
    rownames_to_column("city") |> 
    inner_join(cities, join_by(city)) |> 
    arrange(desc(lat)) |> 
    mutate(city = factor(city, levels=city)) |> 
    ggplot(aes(x=city, y=V1)) + 
    geom_col(fill="#8B000033", color="red4") + 
    xlab("City") + 
    ylab("Principal Component 1") + 
    theme_bw()
```

So we see that our PC takes a bit from each city, but is most heavily loaded
on the southernmost cities in this sample. 

If we return and look at $\bX \bv_1$ or equivalently $\bx_i^{\top}\bv_1$ for
each observation $i$, we get the following

```{r}
#| cache: true
temperature_mat %*% temperature_pc1$v |> 
  data.frame() |> 
  rename(pc_row=1) |> 
  mutate(time = temperature_df$time) |> 
  ggplot(aes(x=time, y=pc_row)) + 
  geom_point() + 
  xlab("Time") + 
  ylab("X * v")
```

So there is a time structure! Note that this isn't _quite_ the time series we'd
get from just an average of the five temperatures--it's a weighted average via
the columns above--but it's close. 

Let's go ahead and add some cities from the other side of the world to get
a more interesting pattern: 

```{r}
#| cache: true
library(tidyverse)
library(httr2)
eps <- 0.2
cities <- tribble(
    ~city, ~lat, ~lon, 
    "NYC", 40.73, -73.93,
    "BOS", 42.36, -71.06,
    "PHL", 39.96, -75.17,
    "SYD", -33.87, 151.21,
    "MEL", -37.81, 144.97,
    "BNE", -27.47, 153.03
) |> 
    mutate(lat_p = lat + eps, 
           lat_n = lat - eps, 
           lon_p = lon + eps, 
           lon_n = lon - eps) 

resp <- request("https://api.open-meteo.com") |>
    req_url_path("v1/forecast") |>
    req_url_query(latitude=paste(cities$lat, collapse=","), 
                  longitude=paste(cities$lon, collapse=","),
                  hourly="temperature_2m", 
                  past_days=60) |>
    req_perform()

temperature_df2 <- resp |> 
    resp_body_json() |> 
    map(\(r) data.frame(time = r |> pluck("hourly", "time") |> list_c(), 
                        temperature_2m = r |> pluck("hourly", "temperature_2m") |> list_c(), 
                        latitude = r |> pluck("latitude"), 
                        longitude = r |> pluck("longitude"))) |> 
    list_rbind() |> 
    mutate(time=ymd_hm(time)) |> 
    left_join(cities, join_by(between(latitude, lat_n, lat_p), 
                              between(longitude, lon_n, lon_p))) |> 
    select(time, temperature_2m, city) |> 
    pivot_wider(names_from=city, 
                values_from=temperature_2m) |>
    mutate(across(-time, \(x) x - mean(x))) # Remove baseline temperature


temperature_mat2 <- temperature_df2 |> 
    select(-time) |> 
    as.matrix()

temperature2_pc1 <- pca_power_method(temperature_mat2)

temperature2_pc1 |> 
    pluck("v") |> 
    as.data.frame() |> 
    rownames_to_column("city") |> 
    inner_join(cities, join_by(city)) |> 
    arrange(desc(lat)) |> 
    mutate(city = factor(city, levels=city)) |> 
    ggplot(aes(x=city, y=V1)) + 
    geom_col(fill="#8B000033", color="red4") + 
    xlab("City") + 
    ylab("Principal Component 1") + 
    theme_bw()
```

Here, we've augmented our US cities with the Australian cities of Sydney
(`SYD`), Brisbane (`BNE`) and Melbourne (`MEL`). These cities are on
the other side of the world from the US cities we've looked at and their
temperatures move contrary to the US cities (both their seasonal cycles
and their day-night cycles) so this isn't too surprising. 

::: {.callout-tip title="Pause for Reflection"}

Modify the example above to include some Northern + Eastern Hemisphere
cities like Beijing and Kyiv and some Southern + Western cities like
Sao Paolo and Buenos Aires. What patterns do you find? Does this make sense?
What does this tell you about weather patterns? What happens if you take
daily average temperatures instead of hourly data?

:::

As before, the 'time series' that goes with this data is interesting: 

```{r}
temperature_mat2 %*% temperature2_pc1$v |> 
  data.frame() |> 
  rename(pc_row=1) |> 
  mutate(time = temperature_df2$time) |> 
  ggplot(aes(x=time, y=pc_row)) + 
  geom_point() + 
  xlab("Time") + 
  ylab("X * v")
```

It's actually not that different from what we saw above, in part because the
time patterns are correlated. If we were to put these together, we would expect
Sydney to move _opposite_ this pattern (zigging when this pattern zags) since it
has a negative loading. 

## A Singular Value Approach to PCA

Above, to get the time course, we computed $\bX\bv$, but this is a bit hokey.

It's easier to use the _SVD_-formulation of PCA, which we'll now introduce.
Given a matrix $\bX \in \R^{n \times p}$, we want to break it up into a sum of
simple components: for our purposes, these 'simple' components can be expressed
as rank-1 matrices. That is, our 'simple' components are basically 
one-dimensional bits (line segments) in a higher-dimensional space. So we're
looking for something like: 

$$\bX = \bX_1 + \bX_2 + \dots $$

If $\bX$ is rank-$k$, we should expect that we could use a sum of only $k$ 
rank-1 terms to exactly equal $\bX$. Determining the rank of a matrix is not
trivial, but an upper bound is $\text{rank}(\bX) \leq \min\{n, p\}$. That is, 
the rank of a matrix is bounded above by (and, it turns out, is typically equal 
to) its smaller dimension. So, if $\bX \in \R^{3 \times 5}$, we know the
rank of $\bX$ - and hence the number of sum terms we will need - is at most
3. This shouldn't surprise you - if we needed four degrees of freedom to 
reproduce 3 points, we would clearly be doing something inefficiently. 

So we have our goal: 

$$\bX = \bX_1 + \bX_2 + \dots + \bX_{\text{rank}(\bX)}$$

But these terms $\bX_i$ are not so easy to work with. As noted above, the idea
of forcing each $\bX_i$ to be rank-1 is not easy, but we have a trick! Any 
rank-1 matrix can be written as the (outer) product of two vectors: 

$$\bX_i = \bu_i\bv_i^{\top}$$

where $\bu_i \in \R^n$ and $\bv_i \in \R^p$. The fact that this gives a rank-1
term shouldn't be too surprising: by going through two vectors, we essentially
create a 'bottleneck' that forces things down to one dimension. 

Obviously, this sort of $\bu_i\bv_i^{\top}$ is not a _unique_ representation of
a given rank-1 matrix. If $\bA$ were rank-1, we could write it as: 

$$\bA = \bu \bv^{\top} = (\bu/2)(2\bv)^{\top} = (\alpha\bu)(\bv/\alpha)^{\top} \text{ for any } \alpha \neq 0$$

To avoid this, let's force $\bu$ and $\bv$ to be unit vectors and to put any 
scale factors outside the product: so, we would have something like

$$\bA = d\, \bu\bv^{\top}$$

where $d$ is a scalar and $\bu \in \R^n, \bv \in \R^p$ are unit vectors. We 
still have a bit of sign ambiguity ($d\bu = (-d)(-\bu)$) but we can fix
this by setting $d > 0$.[^sign]

So now, we know we want to break up our original data matrix $\bX$ as: 

$$\bX = d_1\bu_1\bv_1^{\top} + d_2\bu_2\bv_2^{\top} + \dots d_k\bu_k\bv_k^{\top}$$

where $k = \text{rank}(\bX)$. This seems so much harder! But we have a trick
up our sleeves. 

Think back to when we introduced the _singular value decomposition_ of a matrix
$\bX$. Recall that we had

$$\bX = \bU\bD\bV^{\top}$$


where

- $\bD$ is a diagonal matrix with all positive elements on the diagonal
- $\bU$ and $\bV$ are an orthogonal matrices

At the time, I said not to be too concerned about the exact dimensions of each
term, but we can now be precise: 

$$\bX = \bU\bD\bV^{\top}$$

for a rank-$k$ matrix where

- $\bD \in \R^{k \times k}$ is a diagonal matrix with all positive values on the
  diagonal
- $\bU \in \R^{n \times k}$ is an orthogonal matrix satisfying $\bU^{\top}\bU = \bI_{k \times k}$
- $\bV \in \R^{p \times k}$ is an orthogonal matrix satisfying $\bV^{\top}\bV = \bI_{k \times k}$

With this precise definition in place, let's start to expand this sum out. Because
$\bD$ is diagonal, we see that: 

$$\begin{align*}
\bX &= \bU\bD\bV^{\top} \\
    &= \begin{pmatrix} \bu_1 \\ \bu_2 \\ \dots \\ \bu_k \end{pmatrix} \begin{pmatrix} d_1 \\  0 & d_2 \\ && \ddots \\ &&& d_k \end{pmatrix} \begin{pmatrix} \bv_1 \\ \bv_2 \\ \dots \\ \bv_k \end{pmatrix}^{\top} \\
    &= \begin{pmatrix} d_1\bu_1 \\ d_2\bu_2 \\ \dots \\ d_k\bu_k \end{pmatrix} \begin{pmatrix} \bv_1 \\ \bv_2 \\ \dots \\ \bv_k \end{pmatrix}^{\top} \\
    &= d_1\bu_1\bv_1^{\top} + d_2\bu_2\bv_2^{\top} + \dots + d_k\bu_k\bv_k^{\top}
\end{align*}$$

So the SVD gives us _exactly_ the PCA decomposition we were looking for.
If you look inside most PCA software (*e.g.*,
[`sklearn.PCA`](https://github.com/scikit-learn/scikit-learn/blob/main/sklearn/decomposition/_pca.py#L175) in Python or
[`stats::prcomp`](https://github.com/wch/r-source/blob/trunk/src/library/stats/R/prcomp.R#L40) in `R`), you will see that it's all just really a wrapper around an SVD. 

In fact, let's just go ahead and use an SVD to do PCA 'by hand': 

```{r}
temp_svd <- svd(temperature_mat2)
temp_u1 <- temp_svd$u[,1]
temp_v1 <- temp_svd$v[,1]

library(patchwork)
p1 <- data.frame(u = temp_u1, time=temperature_df2$time) |>
  ggplot(aes(x=time, y=u)) + 
  geom_line() + 
  geom_point(alpha=0.2) + 
  xlab("Time Axis") + 
  ylab("U-Factor from PC1") + 
  theme_bw()

p2 <- data.frame(v = temp_v1, city=cities$city) |>
  ggplot(aes(x=city, y=v)) + 
  geom_col() + 
  geom_point() + 
  xlab("City") + 
  ylab("V-Factor from PC1") + 
  theme_bw()

p1 / p2
```

This is (up to scaling - via that $d$ term) what we saw before!

So the `svd()` gets us everything we want, but it's a bit magic. We can also
'build-our-own' SVD by returning to our approximation problem. To get the first
PC, we can write the problem of approximating $\bX$ as: 

$$\argmin_{d, \bu, \bv} \|\bX - d\bu\bv^{\top}\|_2^2$$

Here, we want to pick $d, \bu, \bv$ to approximate $\bX$ as well as possible. 
After a bit of algebra, we can show this becomes

$$\argmax_{\bu, \bv} \bu^{\top}\bX\bv$$

::: {.callout-tip title="Deriving the SVD Approximation Problem" collapse="true"}

We begin with 

$$\argmin_{d, \bu, \bv} \|\bX - d\bu\bv^{\top}\|_2^2$$

If we expand out the binomial, this becomes: 

$$\begin{align*}
\|\bX - d\bu\bv^{\top}\|_2^2 &= \Tr\left[(\bX - d\bu\bv^{\top})^{\top}(\bX - d\bu\bv^{\top})\right] \\
&= \Tr\left[(\bX^{\top} - d\bv\bu^{\top})(\bX - d\bu\bv^{\top})\right] \\
&= \Tr\left[\bX^{\top}\bX- d\bv\bu^{\top}\bX - d\bX^{\top}\bu\bv^{\top} + d^2\bv\bu^{\top}\bu\bv^{\top}\right] \\
&= \|\bX\|_2^2 - d\Tr(\bv\bu^{\top}\bX) - d\Tr(\bX^{\top}\bu\bv^{\top}) + d^2\Tr(\bv\bu^{\top}\bu\bv^{\top}) \\
&= \|\bX\|_2^2 - d\Tr(\bu^{\top}\bX\bv) - d\Tr(\bv^{\top}\bX^{\top}\bu) + d^2\Tr(\bv\|\bu\|_2^2\bv^{\top}) \\
&= \|\bX\|_2^2 - d\Tr(\bu^{\top}\bX\bv) - d\Tr(\bu^{\top}\bX\bv) + d^2\Tr(\bv^{\top}\bv) \\
&= \|\bX\|_2^2 - 2d\Tr(\bu^{\top}\bX\bv) + d^2 \\
&= \|\bX\|_2^2 - 2d(\bu^{\top}\bX\bv) + d^2
\end{align*}$$

Holding off on finding $\bu, \bv$ for now, let's find an optimal $d$. As always,
let's take the derivative with respect to $d$ and set it to zero: 

$$ 0 = -2(\bu^{\top}\bX\bv) + 2d \implies d = \bu^{\top}\bX\bv$$

Putting this into above, we have an objective of: 

$$\begin{align*}
\argmin_{\bu, \bv} \|\bX\|_2^2 - 2d(\bu^{\top}\bX\bv) + d^2 &= \argmin_{\bu, \bv} \|\bX\|_2^2 - (\bu^{\top}\bX\bv)^2
\end{align*}$$

From here, we note that $\bX$ is unchanging - it's the data we're trying 
to approximate - so we really just have

$$ \argmin_{\bu, \bv} - (\bu^{\top}\bX\bv)^2$$

which we can rewrite as 

$$ \argmax_{\bu, \bv} \bu^{\top}\bX\bv $$

:::

So what does this problem really mean? We're looking for $\bu, \bv$ that 
maximize this term. Informally, you can think of this as finding the $\bu, \bv$
that are best aligned with $\bX$. A bit more rigorously, if we expand $\bX$ out
with its SVD: 

$$\begin{align*}
\argmax_{\bu\bv} \bu^{\top}\bX\bv &= \argmax_{\bu\bv} \bu^{\top}\left(\sum d_i \bu_i\bv_i^{\top}\right)\bv \\
&= \argmax_{\bu\bv} \sum d_i (\bu^{\top}\bu_i)(\bv_i^{\top}\bv
\end{align*}$$

So our objective is some weighted combination of the $d_i$ values, based on how
aligned our $\bu$ is with each 'real' $\bu_i$. Since the $d_i$ are ordered, the
best solution is the one we get when $\bu = \bu_1$ and $\bv = \bv_1$, which is
exactly what we want!

So now that we've convinced ourselves that

$$\argmax_{\substack{\bu \in \R^n, \bv \in \R^p \\ \|\bu\| = \|\bv\|=1}} \bu^{\top}\bX\bv$$ 

is the problem we want to solve, how do we actually do that? We have two 
variables and handling them both at the same time is tricky, so let's just do
one at a time. If we fix $\bv$, it's easy to see that the best $\bu$ is the
one that is just aligned with $\bX\bv$; similarly, if we fix $\bu$, the best
$\bv$ is just one aligned with $\bu^{\top}\bX$. We can put these together into 
an power method algorithm that looks similar to the one we developed above 
for just $\bv$: 

- Initialize $\bv^{(0)}$ randomly 
- Repeat the following until convergence:
  $$ \bu^{(k)} \leftarrow \frac{\bX\bv^{(k)}}{\|\bX\bv^{(k)}\|}$$
  $$ \bv^{(k+1)} \leftarrow \frac{\bX^{\top}\bu^{(k)}}{\|\bX^{\top}\bu^{(k)}\|}$$

In code, this might look something like: 

```{r}
pca_power_svd <- function(X, thresh=1e-8){
    n <- NROW(X); p <- NCOL(X)
    
    normalize <- function(x) x / sqrt(sum(x^2))
    
    v <- normalize(matrix(rnorm(p), ncol=1))
    
    repeat {
        u <- normalize(X %*% v)
      
        v_old <- v
        v <- normalize(t(X) %*% u)
        
        # To fix sign irregularity, force v[1] > 0
        if(v[1] < 0) v <- -v
        
        if(sum(abs(v - v_old)) < thresh){
            break
        }
    }
    
    list(u=u, v=v, d=t(u) %*% X %*% v)
}
```

which gives us: 

```{r}
pca_svd_temp <- pca_power_svd(temperature_mat2)

p1 <- data.frame(u=pca_svd_temp$u, time=temperature_df2$time) |>
  ggplot(aes(x=time, y=u)) + 
  geom_line() + 
  geom_point(alpha=0.2) + 
  xlab("Time Axis") + 
  ylab("U-Factor from PC1") + 
  theme_bw()

p2 <- data.frame(v=pca_svd_temp$v, city=cities$city) |>
  ggplot(aes(x=city, y=v)) + 
  geom_col() + 
  geom_point() + 
  xlab("City") + 
  ylab("V-Factor from PC1") + 
  theme_bw()

p1 / p2
```

which matches what we got doing it with the built-in `svd()`.

[^sign]: You might still notice we have a $\bu\bv^{\top} = (-\bu)(-\bv)^{\top}$
ambiguity. This actually can't be resolved _uniquely_ and is a bit beyond our
course, but most software will typically do something meaningless like "make the
first value of $\bu$ positive" or "make the largest value of $\bu$ positive."




## Making Sense of PCA

The temperature results above are interesting, but not the easiest to interpret. 
Australia being 'opposite' the US makes sense, but if we want to go deeper in 
our analysis, what should we do? I like the strategy of using PCs to build an
'estimated' $\bX$, based on the 'sum of simple' components above. In the 
easiest case, we can make sense the first PC by looking into 

$$\hat{\bX}_1 = d_1 \bu_1 \bv_1^{\top} \in \R^{n \times p}$$





## Regularized PCA

## Other Applications of PCA-Type Decompositions

