---
session: "`r session <- 7; session`"
topic: "`r library(tidyverse); topic <- read_csv('../key_dates.csv', name_repair='universal') |> filter(Course.Element == 'Class Session', Item.Number==session) |> pull(Details); topic`"
date: "Last Updated `r Sys.Date()`"
---

{{< include _latex_defines.qmd >}}

Having (temporarily) finished our discussion of regression, we now turn our 
focus to *classification*. 

## A Taxonomy of Classification

Recall from before that we defined _regression_ as the subset of supervised
learning where the response ($\by$) was a continuous variable, or at least one
that is appropriate to model as such.[^effective_cont] By contrast, 
*classification* considers $\by$ to take values in a discrete set. Though this
restriction may feel limiting, it is actually quite powerful. We consider several
different forms of classification, based on the *support* of $\by$: 

[^effective_cont]: Consider, *e.g.*, predicting someone's annual tax liability.
*Technically*, this is *discrete* since tax liabilities are rounded to the nearest
dollar, but functionally it may as well be continuous. 

- *Binary* classification: in this simplest case, $\by$ takes one of two values,
  which we conventionally code as $\{0, 1\}$ or $\{-1, +1\}$, depending on which
  makes the math easier. Note that the actual response need not be these numbers;
  they are simply a mathematical convenience. 
  
  Examples of binary classification include: 
  
  - Does an incoming student graduate within the next 6 years or no?
  - Does a patient have a given disease? 
  - Is there a dog present in an image? 
  - Is this applicant qualified for a job? 
  
  Note that, in at least two of these, the underlying question is (arguably)
  not quite so binary, but we may choose to treat it as such. In particular,
  when assessing onset of a disease, the actual impairment to the patient is 
  continuous, but it is common in medical practice to "binarize" the outcome. 
  Any of you with aging family members know that cognitive decline occurs long
  before it's 'officially' Alzheimer's. Similarly, qualification for a job
  is perhaps best understood as a continuous (or even purely comparative)
  quantity "Is A more qualified than B?" but we chooes to binarize it when
  deciding to whom a job offer should be made. 
  
  Recalling your probability theory, there is (essentially) only one distribution
  on $\{0, 1\}$, the Bernoulli, so this scenario typically has the cleanest and
  simplest mathematics. 
  
- *Multiclass* classification: in this case, $\by$ takes values in a set of 
  finite and known, but potentially large, set of outcomes, which we will
  call $\Ycal$. Before about 20 years ago, multiclass problems took 
  $|\Ycal|$ to be not much larger than 2, *e.g.*, which blood type does
  a sample match or what language is a text sample written in, but much of the 
  modern practical success of machine learning comes from the ability to treat 
  *very large* outcome sets $\Ycal$. 
  
  For instance, modern computer vision (CV) has moved beyond far simple binary
  classification: 
  
  {{< video https://www.youtube.com/watch?v=ACmydtFDTGs >}}
  
  The famed[`ImageNet`](https://en.wikipedia.org/wiki/ImageNet) benchmark data
  set has over 20,000 classes (arranged with additional structure) and modern
  data sets include even more. This can be contrasted with the older 
  [`MNIST`](https://en.wikipedia.org/wiki/MNIST_database) data set which 
  collects only ten handwritten digits.
  
  Beyond CV, large multiclass classification has also revolutionized text modeling: 
  when a chatbot generates text, it selects the next word from a large but finite
  set of English words. 
  
  Just like binary classification is rooted in the Bernoulli distribution, 
  multiclass classification is rooted in the [categorical
  distribution](https://en.wikipedia.org/wiki/Categorical_distribution). When
  applied to a large class of outcomes, the categorical distribution becomes a
  [multinomial](https://en.wikipedia.org/wiki/Multinomial_distribution), 
  *cf* Bernoulli to binomial, so multiclass classification is also frequently
  called *multinomial* classification. 
  
While our main focus is on *binary* and *multiclass* problems, it is worth noting
some additional "modes" of classification. 

- *Ordinal* classification: in this case, $\Ycal$ is an _ordered_ set. The most
  common form of ordinal classification is the ubiquitous 
  [*Likert scale*](https://en.wikipedia.org/wiki/Likert_scale) which
  you have certainly encountered any time you have filled out a customer survey:
  
  - *Strongly Agree*
  - *Agree*
  - *Neutral*
  - *Disagree*
  - *Strongly Disagree*
  
  In this case, the categories clearly have an order and a ranking, but we can't
  quite treat them as 'numerical' in the usual sense: *Strongly Agree* is not
  just *Agree* + *Disagree*. 
  
  Proper analysis of *ordinal* data is tricky and often poorly done. 
  Complexities include the inappropriateness of statisticians' most beloved tool,
  averaging[^set], and a fundamental 'subjectiveness.' 
  Another common ordinal classification problem is the assignment of grades
  (A > B > C ...). In this case, students and professors may agree that 
  one piece of work may be better than another, but a student may feel their
  assignment deserves an A while the professor may judge it B work. It is
  quite difficult to infer each individual's personal threshold for each category
  without extensive analysis. 
  
  We will not discuss ordinal data much in this course as it requires specialized
  methods. Typically ordinal data is modeled by treating the response as a rounded
  version of some underlying continuous variable: *e.g.*, when grading hurricanes
  the classes (Category 1, Category 2, ...) are derived from the underlying
  continuous variable of wind speed. This 'hidden regression' structure makes
  analysis of ordinal classification much closer to regression than binary
  classification. 

[^set]: Even though it doesn't make sense to average ordinal data, University
administrators are *notorious* for averaging Likert data to evaluate faculty. 
This is a nigh-universal grievance of numerically-minded faculty, but getting
universities to change their established procedures to something statistically
reasonable is only slightly less impossible than drawing blood from a turnip.
  
- *Probabilistic* classification: we noted above that binary classification can
  be thought of as modeling a Bernoulli random variable. If we shift our focus
  to modeling the Bernoulli parameter (the probability $p \in [0, 1]$) than
  the outcome ($y \in \{0, 1\}$), we have the paradigm of *probabilistic*
  classification. 
  
  As a general rule, *statistically*-grounded classification methods have
  a natural probabilistic output built-in while methods coming from the CS 
  tradition are not inherently probabilistic. Because probabilistic estimates
  are often quite useful, there exist several [useful
  methods](https://en.wikipedia.org/wiki/Platt_scaling) to 'bolt-on' 
  probabilistic outputs to pure binary classifiers. 
  
## Metrics of Classification Accuracy

In the regression context, we had a relatively small number of loss functions: 

- Mean Squared Error (MSE): mathematically convenient, easy interpretation, 
  natural for OLS, optimizes for mean
- Mean Absolute Error (MAE): mathematically a bit tricky, most practical in many
  circumstances, gives rise to *robust* regression, optimizes for median
- Mean Absolute Percent Error (MAPE)
- Checkmark/Pinball Loss: gives rise to *quantile* estimation
- Huber loss: interpolates MSE and MAE

By contrast, in the classification context, we have *many* loss functions to 
work with. To get into them, it's worth thinking back to the formalism of
*statistical hypothesis testing*. Recall that, in hypothesis testing, we have
a 2-by-2 table of possible outcomes: 

+--------------+----------------+-----------------+
|              | **Truth**                        |
+==============+:==============:+:===============:+
| **Decision** | **Null**       | **Alternative** |
+==============+================+=================+
| **Retain**   | True           | False Negative  |
| **Null**     | Negative       | (Type II Error) |
+--------------+----------------+-----------------+
| **Reject**   | False          | True Positive   |
| **Null**     | Positive       |                 |
|              | (Type I Error) |                 |
+--------------+----------------+-----------------+

: Hypothesis Testing Table

Suppose that we have a large number ($n$) of binary classification points. If we
let $\hat{y}_i$ be our prediction and $y^*_i$ be the actual outcome, we can
construct a similar table: 

+----------------+-----------------+--------------------------+
|                | **Truth** (Second index)                   |
+================+:===============:+:========================:+
| **Prediction** | $y_i^* = 0$     | $y_i^* = 1$              |
| (First Index)  |                 |                          |
+================+=================+==========================+
| $\hat{y}_i=0$  | $n_{00}$ True   | $n_{01}$ False Negative  |
|                | Negatives       | (Type II Errors)         |
+----------------+-----------------+--------------------------+
| $\hat{y}_i=1$  | $n_{10}$ False  | $n_{11}$ True Positives  |
|                | Positives       |                          |
|                | (Type I Errors) |                          |
+----------------+-----------------+--------------------------+

: Classification Outcome Table


From this structure, we get several useful ideas: 

- A *confusion matrix* is a 2x2 (or $K$-by-$K$ for $K$-multiclass problems)
  table comparing the right answer (here "Truth" on the columns) with our guess
  (here "Decision" on the rows). A good classifier will have large values 
  "on diagonal" ($n_{00} + n_{11}$) and small values "off diagonal" 
  ($n_{01} + n_{10}$)
- The notion of *true/false* *positive/negatives*. These can be a bit tricky
  to remember, but the convention is: 
  
  - The *noun* (positive/negative) captures the *prediction*
  - The *adjective* (true/false) assesses the prediction
  
  So a "false positive" means that we guessed the positive (+1) class, but that
  we were wrong and the truth was the negative (-1 or 0) class.
  
From this simple count table, we can create _many_ different ratios. I find the
statistical terminology a bit more intuitive than the ML terminology: 

- True Positive Rate: Of the actual positives ($y^*=1$), what fraction are
  predicted correctly (truly) positives ($\hat{y} = 1$)? 
  $\text{TPR} = n_{11}/(n_{01} + n_{11})$
- False Negative Rate: Of the actual positives ($y^*=1$), what fraction are
  predicted incorrectly (falsely) as negatives ($\hat{y} = 0$)? 
  $\text{FNR} = n_{01}/(n_{01} + n_{11})$
- True Negative Rate: Of the actual negatives ($y^*=0$), what fraction are
  predicted correctly (truly) as negatives ($\hat{y} = 0$)?  
  $\text{FNR} = n_{00}/(n_{00} + n_{10})$
- False Positive Rate: Of the actual negatives ($y^*=0$), what fraction are
  predicted incorrectly (falsely) as positives ($\hat{y} = 1$)? 
  $\text{FPR} = n_{10} / (n_{00} + n_{10})$
  
These clearly satisfy some useful relationships: 

$$\text{TPR} = 1 - \text{FNR} \Leftrightarrow \text{FNR} = 1 - \text{TPR}$$

and 

$$\text{FPR} = 1 - \text{TNR} \Leftrightarrow \text{TNR} = 1 - \text{FPR}$$

In this scenario, we recognize the *false positive rate* as the *size* (or 
*level*) associated with a statistical test (5% for a 95% confidence test) and
the true positive rate as the *power*. Again, the statistical definitions can
help us navigate the confusing terminology here: the confidence of a test,
and therefore its false positive rate, are defined _assuming the null_ (*i.e.*,
in a "true" negative scenario). 

The "other direction" of rates -- where we condition on the prediction 
instead of the truth -- are less common, except for the *false discovery rate*: 

$$\text{FDR} = \frac{n_{10}}{n_{10} + n_{11}}$$

The FDR is commonly used in scientific settings and it answers this question: 
if a scientist makes $K$ discoveries, what fraction _of those claimed 
discoveries_ are correct? 

::: {.callout-warning} 

Note that the convention used to define *rates* can be quite confusing
and is perhaps opposite the convention used to define *counts*. 

To compute the number of *true positives*, we look at predicted 
positives and count the number of "truths", while to compute
the *true positive rate*, our denominator is the number of 
_actual_ positives and we count the fraction of correct predictions. 

To me, the FDR is actually the most clearly named rate; I try to remember
the definition of FDR and 'work backwards' to FPR since it must be something
different than FDR.

:::

In ML context, you will commonly hear reference to *precision* and *recall*. 
With our definitions above, we have

- Precision = $1 - \text{FDR}$
- Recall = $\text{FNR} = 1 - \text{TPR}$

so, by maximizing precision, we ensure that all of our predicted positives are
true positives and, by maximizing recall, we ensure that we have no false 
negatives (that is, we truly identify all of the real positives). 

In the medical context, you may encounter alternative metrics of 

- Sensitivity, equal to TPR
- Specificity, equal to TNR

This zoo of terms is admittedly quite confusing, but the Wikipedia article
on [Precision and Recall](https://en.wikipedia.org/wiki/Precision_and_recall#Definition)
has an excellent table. 

### Combining Metrics

If you look at the 2x2 table above, you can convince yourself that it has 
two "degrees of freedom:" having a high TPR implies nothing about the TNR 
(or equivalent metrics). That is, it is possible to have a classifier with
arbitrarily good TPR (equivalently, small FNR) and terrible TNR (equivalently,
small FPR). 

In fact, we can construct one trivially: 

$$\hat{f}(\cdot) = 1$$ 

That is, the function that predicts $1$ _always_. This classifier has zero false
negatives, because it has no negatives whatsoever, but it necessarily has many
false positives.[^rate] 

Conversely, 

$$\hat{f}(\cdot) = 0$$ 

has no false positives, but many false negatives. 

[^rate]: In this case, the FPR is exactly equal to the negative rate in the 
population. 

Clearly, we need a 'combination' metric that combines both degrees of freedom
into a single score. Popular choices include: 

- Accuracy: the fraction of correct predictions ($(n_{00}+n_{11})/n$)
- Balanced Accuracy: $\frac{\text{TPR} + \text{TNR}}{2}$
- $F_1$ Score: $2 *(1 - \text{FDR}) / (\text{TPR} - \text{FDR} +1)$

While these choices are popular, none of them are actually all that suitable
for a real application. To properly evaluate a classifier, we need to know
the real cost of false positives and false negatives, as well as the population
base rates, and balance things accordingly. 

For instance, if we are designing a 'baseline' medical diagnostic to be used
as part of annual physicals, we need to think about what happens with false
negatives and false positives. 

- If we give a false positive, patients will undergo a better test that costs
  \$1,000 but gives an accurate answer in all circumstances (a "gold standard").
- If we give a false negative, patients will leave a condition untreated until
  it becomes far worse, requiring expensive surgery costing \$50,000. 
  
In this case, we might desire a classifier that minimizes $$n_{10} + 50 n_{01}$$,
not any of the of the 'standard' metrics. Note also that the particular FPR, FNR
of this classifier will depend on the rate of the condition in the population,
so we can't actually find the best procedure by looking at FPR, FNR in isolation.

While this *decision-theoretic* approach is optimal, it is often quite difficult
to characterize in practice. In the example above, for instance, we made the
choice to only minimize cost, without seeking to minimize suffering. (What if
the untreated condition is _incredibly_ painful? Should we intervene more often?)
This sort of decision-making is ultimately very problem-specific and too often
ignored in the academic literature. In practice, designing (and optimizing)
the correct loss function is far more relevant to business outcomes than
finding the best classifier to optimize an irrelevant cost function. 
procedure requires 

### Trade-Off of Recall and Precision

Our discussion of 'trivial' classifiers that focus on one metric while completely
ignoring the other may remind you of the way this is typically solved in
statistical testing procedures. We specify a maximum acceptable false positive
rate (level or 1 - confidence) and then minimize the false negative rate
subject to that constraint; in statistical speak, we seek a test with 
_maximal power_. 

As we have seen with ridge and lasso regression, there is typically a 'knob'
(hyperparameter) we can pick to tune a particular classifier. It is common to
apply the classifier at all values of this knob, measure the TPR and TNR at all
levels, and plot them in a trade-off curve. This curve is, for historical 
reasons, known as the receiver operating characteristic (ROC)[^roc] curve and
it looks something like: 

```{r}
#| message: false
library(yardstick)
library(ggplot2)
library(dplyr)
library(ggplot2)
library(dplyr)
# Example from ?roc_curve
roc_curve(two_class_example, truth, Class1) %>%
  ggplot(aes(x = 1 - specificity, y = sensitivity)) +
  geom_path() +
  geom_abline(lty = 3) +
  coord_equal() +
  theme_bw()
```

Here we see that, for this particular classifier, increased specificity is
associated with decreased sensitivity or, in the alternative terms, increased TNR
is associated with decreased TPR. As illustrated by our example of constant
predictors above, this is a quite typical trade-off. 

Curves like this can be used in a few ways, none of which are fully satisfying: 

- If we have two curves, we can plot their ROCs on the same set of axes. If one
  classifier is _uniformly_ above the other, it is said to _dominate_ the other
  and is better for all applications. Sadly, this is rarely the case. 
- If we have two curves and plot their ROCs, but the ROCs cross, we don't have
  strong guidance on which model to prefer, but if one is _almost always_ above
  the other, we will likely still use it. 

A commonly reported metric is the _area under the ROC curve_ (AUC), which ranges
from 0.0 to 1.0, with higher values being better. This is a bit unsatisfying
compared to the decision theoretic analysis, but it is an improvement over
some other metrics as it at least addresses the fact there is a tradeoff. 

::: {.callout-info title="Pause for Reflection"}

What does an AUC less than 0.5 imply? If you get an AUC less than 0.5, what 
should you do? Why is AUC near 0 actually a good thing?

:::

[^roc]: Don't confuse this with _Rate of Change_. 

While the ROC is typically motivated by presence of a 'tuning knob', even
methods like logistic regression (which have no obvious knobs) have an implicit
knob we can choose to tune: the _classification threshold_. There is no law of
nature that says that a predicted probability of 55\% has to correspond to
a prediction of the positive class. If false positives are very expensive, we
may want to reduce the fraction of positives predicted and only predict 
the positive class when the estimated probability is over 80\%. Similarly, 
if false negatives are very bad, we may want to predict positive even when
the probability of a positive is in the 40\% range: this is particularly 
common in medical contexts where the impact of a missed diagnosis is quite
substantial, so doctors order extra diagnostic work 'just to be sure' even
if the test does not strongly suggest a disease. 

### Scoring Functions

_Scoring_ functions are loss functions particularly suited for evaluation of 
probabilistic classifiers. Unlike our confusion-matrix set of metrics, it's less
clear how we should evaluate probabilistic classifiers. 

It is clear that, if we predict something with 80% probability and it doesn't
happen, this is a 'worse' mistake than if we only had a 51% probability, but
how can we measure this formally? 

- Calibration and Tightness
- Log Scores

## Types of Classification Methods

In this course, we will consider two main types of classifiers: 

i) Generative classifiers, which attempt to jointly model $(\bX, \by)$ and use
   probabilistic mathematics (mainly Bayes' Rule) to infer $\by | \bX$ from
   the joint model. 
   
ii) Discriminative classifiers, which directly attempt to estimate the 'boundary'
    between the classes. 

As you will see, the terminology around these methods is _particularly_ terrible. 

## Building Mutliclass Classifiers from Binary Classifiers
  
As we introduce methods, we will typically derive them for the binary
classification task. If we want to apply them to the multiclass task, how can 
we extend them? It would be quite cumbersome if we needed _brand new_ methods
each time. 

In general, there are two main strategies we might use: 

- One vs Rest
- Each vs Each

For simplicity, assume we are predicting $K$ classes. In the "one-vs-rest"
strategy, we will build $K$ binary classifiers, of the form: 

- Class 1 vs Not Class 1 (*i.e.*, Class 2, 3, 4, $\dots$, $K$)
- Class 2 vs Not Class 2 (*i.e.*, Class 1, 3, 4, $\dots$, $K$)
- *etc.*
- Class $K$ vs Not Class $K$ (*i.e.*, Class 1, 2, 3, $\dots$, $K-1$)

To make an actual prediction, we take the *highest* score of each of these
classifiers. *E.g.*, if $K=3$ for categories "Red", "Green" and "Blue", we
might predict: 

- $\mathbb{P}(\text{Red}) = 0.8$ and $\mathbb{P}(\text{Not Red}) = 0.2$
- $\mathbb{P}(\text{Green}) = 0.4$ and $\mathbb{P}(\text{Not Red}) = 0.6$
- $\mathbb{P}(\text{Blue}) = 0.2$ and $\mathbb{P}(\text{Not Red}) = 0.8$

Our final prediction would then be the maximizer, "Red."

This strategy is relatively simple and intuitive, but for very large $K$, 
the maximum can be far less than 50%, so it may feel a bit funny. (Why is 
this not an issue for $K=2$?)

Conversely, in the "each vs each" (sometimes called "one vs one") setting, 
we train $\binom{K}{2} = K(K-1)/2$ classifiers for every possible _pair_ of
classes. To get the prediction, we then take a 'majority vote' of the classifiers. 

In our $K=3$ example above, we would have three classifiers: 

- Red vs Blue, which we fit after throwing out the "Green" points
- Red vs Green, which we fit after throwing out the "Blue" points
- Green vs Blue, which we fit after throwing out the "Red" points

If we make predictions for a single point, our binary classifications might be

- Red vs Blue: Red
- Red vs Green: Red
- Green vs Blue: Blue

In this case, we would take a majority vote to get "Red" as our final prediction.

I find this strategy a bit counterintuitive, but it is sometimes faster than
the one-vs-rest strategy since each individual model is fit to a subset
of the entire data set.

Certain modern ML strategies, in particular deep neural networks, predict
the probability of all $K$ classes from a single model. In this case, we 
typically select the single most probable class as our prediction. 

In applications where there are multiple contenders and the difference
between them is small, it is common to randomly select a class in
proportion to its predicted probability. This strategy is one of the things
that gives ChatGPT its randomness; it usually predicts the *most likely* next
word, but sometimes it does something a little bit weird.)

Note that it is actually usually pretty hard to make a model that predicts
a large set of probabilities simultaneously: in particular, it's hard to get
a vector that sums to one (like probabilities should). To address this, a
[softmax](https://en.wikipedia.org/wiki/Softmax_function) normalization is
applied. This is essentially nothing more than i) making sure each predicted
value is positive; and ii) dividing them all by the sum so they add to 1. 

## Generative Classifiers

We begin our study of classifiers with an analysis
of *generative* classifiers. While many generative
classifiers have appeared in the literature, they can 
essentially all be derived from a standard formula. We will discuss
this standard formula first and then turn our attention to specific
named instances. 

### Mixture Models

Recall from [STA 9715](https://michael-weylandt.com/STA9715) that
a (discrete) mixture model is a distribution of the form: 

$$X \sim \begin{cases} X_1 & Z = 1 \\ X_2 & Z = 2 \\ \dots \\ X_K & Z=K \end{cases} \text{ where } \textsf{supp}(Z) = \{1,\dots, K\} \text{ and } X_i \sim \mathcal{P}_i \text{ for all $i$ and } Z \perp X_i, X_i \perp X_j\, (i\neq j)$$

That's quite a bit to unpack, so let's take it piece by piece: the
random variable $X$ has a $K$-component *mixture* distribution if
we can sample $X$ by generating $K$ independent random variables
$\{X_1, \dots, X_K\}$ (with *different* distributions) and then
selecting one of them at random. 

Let's look at a concrete version of this: suppose we have
a population of male and female basset hounds that is 60\% female.
The weight of male basset hounds follows a $\mathcal{N}(65, 10)$
distribution while females have weight distributed as 
$\mathcal{N}(55, 10)$. The weight of a randomly selected basset
can be sampled as

```{r}
rbasset <- function(n){
    Z <- rbinom(n, size=1, prob=0.6) # Z=1 <- Female
    ifelse(Z, rnorm(n, mean=55, sd=sqrt(10)), 
              rnorm(n, mean=65, sd=sqrt(10)))
}
rbasset(1)
```

If we collect the weights of many basset hounds, we see that
the resulting distribution is not normally distributed: 

```{r}
hist(rbasset(10000), breaks=50, main="Distribution of Basset Hound Weights")
```

In fact, this is a *bimodal* distribution with each 'lump' representing
males or females.[^bimodal] Distributions of this form can often be
written  more compactly using a 'convex combination' notation: 

[^bimodal]: Note that a mixture of normals is not *always* multimodal
(*e.g.*, if the two means are very close together), but a mixture
of normals is a common model for multimodal data.

$$\text{Basset} \sim \frac{2}{5}\mathcal{N}(65, 10) + \frac{3}{5}\mathcal{N}(55, 10)$$

The above notation can be interpreted as specifying a draw from 
the first distribution with probability $2/5 = 40\%$ and from
the second with probability $3/5 = 60\%$; that is, it's a particular
2-component mixture of normals. 

In specifying this mixture, the probabilities of each component must
be non-negative and must sum to 1; these are exactly the same
restrictions we put on a convex combination of points earlier in
the course. In fact, the set of mixture distributions can be 
consider the 'convex set' containing all of its individual 
components. 

### Inference from a Mixture Distribution

Now suppose you are on a walk and you meet a very friendly 62 
pound basset hound. You of course want to pet this marvelous dog, but
you need to ask the owner's permission first: in doing so, should
you ask "may I pet him?" or "may I pet her?"

While the basset hound won't care if you misgender, you recall
that you can use [*Bayes'
Rule*](https://en.wikipedia.org/wiki/Bayes%27_theorem) to 'invert'
the mixture distribution. That is, while the mixture distribution
normally goes from sex to weight, Bayes' rule will let us go from
weight back to sex. 

Specifically, in this case, Bayes' rule tells us: 

$$\P(\text{Female}|\text{Weight=62}) = \frac{\P(\text{Weight=62}|\text{Female})\P(\text{Female})}{\P(\text{Weight=62}|\text{Female})\P(\text{Female}+\P(\text{Weight=62}|\text{Male})\P(\text{Male}))}$$

The math here is a bit nasty numerically, but we can do it in `R` quite
simply: 

```{r}
p_female <- 0.6
p_male <- 1-p_female

p_62_female <- dnorm(62, mean=55, sd=sqrt(10))
p_62_male   <- dnorm(62, mean=65, sd=sqrt(10))

p_female_62 <- (p_62_female * p_female) / (p_62_female * p_female +
                                           p_62_male * p_male)

paste0("There is a ", round(p_female_62 * 100, 2), "% chance that basset is a female.")
```
Given that calculation, you take a chance and ask whether you 
can pet *him*. 

What have we done here? We have *built* a basset-sex classifier from
a combination of the population distribution and Bayes' rule. This is
the essence of generative classifiers - *if* we know the distribution,
accurate probabilistic classification is a straightforward application
of Bayes' rule. In practice, we don't know the distribution *a priori*
and it must be estimated from data. Different generative classifiers
essentially are just different ways of estimating the population
(mixture) distribution. 

## Building Generative Classifiers

Let's formalize our approach for a two-class classifier though
the extension to multi-class is simple enough. Given a training set

$$\mathcal{D}_{\text{Train}} = \{(\bx_1, y_1), (\bx_2, y_2), \dots, (\bx_n, y_n)\}$$ 

we assume $\bX$ follows a two-component mixture where $Y$ is the 
mixing variable:

$$\bX | Y=0 \sim \mathcal{P}_0 \text{ and } \bX | Y=1 \sim \mathcal{P}_1$$

Clearly to fully specify this distribution, we need three things: 

- To know the distribution of $Y$
- To know the distribution $\mathcal{P}_0$
- To know the distribution $\mathcal{P}_1$

The distribution of $Y$ is relatively straightforward: since it only
takes values $\{0, 1\}$ it _must_ have a Bernoulli distribution and hence
the only thing we need to estimate is the probability, $\pi_1$, that $Y=1$. 
The simplest approach to estimating the Bernoulli probability is to simply
take the empirical probability from the training data set, *i.e.*, the fraction
of observations where $Y=1$, but this can also be estimated using knowledge of
a broader data set or a different population than the training data. 
For instance, if a data set is known to overindex on $Y=1$, *e.g.*, because it
contains patients who self-reported some disease, it may be more appropriate to
estimate $p$ based on the population-wide prevalance of the disease. These concerns
are typically rather problem specific. 

Estimation of $\mathcal{P}_0, \mathcal{P}_1$ may be more difficult. If the 
training data is _extremely_ large, we may use a *non-parametric* estimator, 
*e.g.*, a multivariate histogram, but in general we will need to use some sort
of restrictive (parametric) model. The most commonly used generative classifiers
assume some sort of multivariate normal structure, but you can create 'custom'
classifiers appropriate to specific domains. 

For now, let us assume $\mathcal{P}_{0}, \mathcal{P}_1$ have been estimated
and that they have PDFs $p_0(\cdot), p_1(\cdot)$. Applying Bayes' rule as before, 
with prior (baseline) probabilities $\pi_0, \pi_1 = 1-\pi_0$, we get the following
probabilities

$$\begin{align*}
\P(Y = 1 | \bx) &= \frac{\P(\bx | Y = 1)\P(Y=1)}{\P(\bx | Y = 0)\P(Y=0)+\P(\bx | Y = 1)\P(Y=1)} \\
&= \frac{p_1(\bx)\pi_1}{p_0(\bx)\pi_0 + p_1(\bx)\pi_1}
\end{align*}$$

If we are happy with a *probabilistic* classifier, we are essentially done. If
we want a true binary classifier, *i.e.*, a $\{0, 1\}$ valued 'point predictor', 
we also need to set a classification threshold, $\overline{p}$, to obtain: 

$$\delta(\bx) = \begin{cases} 0 & \frac{p_1(\bx)\pi_1}{p_0(\bx)\pi_0 + p_1(\bx)\pi_1} < \overline{p} \\ 1 & \text{ otherwise} \end{cases}$$

$\overline{p}$ can be chosen to optimize a relevant performance metric (if you 
practice good sample splitting hygiene) or via a full decision theoretic analysis,
as discussed last week. 

That's basically it! 


## Named Generative Classifiers

As mentioned above, the 'popular' generative classifiers are essentially 
defined by specific multivariate normality assumptions on 
$\mathcal{P}_0, \mathcal{P}_1$. Let us review these in ascending order of
complexity. 

Recall that a multivariate normal distribution is uniquely identified by its
*mean* and *(co)variance (matrix)*, just like a standard univariate normal 
distribution. When estimating means, we will simply use the class means for 
each classifier, though James-Stein time estimators could be used for a 
(likely imperceptible) improvement. Estimation of variances is a harder task: 
to estimate the full covariance matrix of $p$ random variables (or a random
$p$-vector), we must estimate: 

i) $p$ (marginal) variances and $\frac{p^2-3p}{2}$ correlations; or 
ii) $\binom{p}{2} = \frac{p(p-1)}{2}$ covariances

Making this even harder, we have to deal with the fact that the "variance of the
variance", *i.e.*, the kurtosis, is typically quite large. While we do not focus
on high-dimensional covariance estimation in these notes, it is quite a rich
subject: if you want dig further into it, the
[Ledoit-Wolf](http://www.ledoit.net/honey.pdf) estimator is a good place to start.

### Naive Bayes

In Naive Bayes, we assume $\mathcal{P}_0, \mathcal{P}_1$ are multivariate normal
distributions with no correlation among the features and a common covariance
across classes.[^isotropic] This *significantly* simplifies the covariance 
estimation problem, since we now have all $n$ samples available to estimate only
$p$ variances. Our estimates are given as follows: 

$$\hat{\mu}_0 = \frac{1}{|\mathcal{C}_0|} \sum_{i \in \mathcal{C}_0} \bx_i$$

where $\mathcal{C}_0$ is the set of observations from class 0 and 

$$\hat{\mu}_1 = \frac{1}{|\mathcal{C}_1|} \sum_{i \in \mathcal{C}_1} \bx_i.$$
The variance estimates are a bit trickier: 

$$\hat{\Sigma}_{ii} = \frac{1}{n}\left(\sum_{j\in\mathcal{C}_0} (x_{ji} - \hat{\mu}_{0i})^2 + \sum_{j\in\mathcal{C}_1} (x_{ji} - \hat{\mu}_{1i})^2\right)$$
This looks a bit nasty, but it's not too hard: we take the standard 
'average squared difference' estimator of the variance, but instead of using
the overall mean, we subtract the estimated mean from the relevant class. Note
that here we are only estimating $\Sigma_{ii}$, the diagonal elements, and not
a full matrix $\Sigma$. 

Naive Bayes classifiers are particularly important to the history of ML as
they were used for one of the first and most influential spam filters. 



[^isotropic]: Sometimes, we go further and assume the standard deviations of 
each feature are the same: this assumption is typically not made explicitly as 
such but is done 'off-camera' by pre-standardizing features.)

### Linear Discriminant Analysis, 

Generalizing a bit, we can relax the assumption that the covariance matrix 
$\Sigma$ is diagonal and only assume that it is constant across the two classes.
Here, 

## Quadratic Discriminant Analysis

## A Non-Standard Example
