---
session: "`r session <- 5; session`"
topic: "`r library(tidyverse); topic <- read_csv('../key_dates.csv', name_repair='universal') |> filter(Course.Element == 'Class Session', Item.Number==session) |> pull(Details); topic`"
date: "Last Updated `r Sys.Date()`"
---

{{< include _latex_defines.qmd >}}

We previously showed the following: 

- OLS can be derived purely as a 'loss minimization' problem without reference
  to specific probabilistic models; this is an instance of the general strategy
  of *empirical risk management*
- OLS is a *convex* problem, making it particularly easy to solve even if we
  don't use the closed-form solution
- The BLUE property of OLS is perhaps less interesting than it sounds as it
  requires very strong assumptions on the data generating process which are
  unlikely to hold in practice
- MSE, the loss we used to pose OLS, can be decomposed as 
  $\text{Bias}^2 + \text{Variance}$
- Because BLUE restricts to _unbiased_ estimators, the MSE of OLS is all variance

We begin this week by asking if we can do better than OLS. To keep things simple,
we begin by assuming we are under a linear DGP (so no 'model error') but that's
only a mathematical niceity. It's not something you should always assume - in fact,
it is really more important to think about how models do on non-linear DGPs. 
As we will see, it may still be useful to use linear models... 

Because OLS is BLUE under our assumptions, we know that we need to relax one
or more of our assumptions to beat it. For now, we will focus on relaxing the
U - unbiasedness; non-linear methods come later in this course. 

Recalling our decomposition: 

$$\MSE = \text{Bias}^2 + \text{Variance}$$

Our gambit is that we can find an alternative estimator with a bit more bias, 
but far less variance. Before we attempt to do so for linear regression, let's 
convince ourselves this is possible for a much simpler problem - estimating
means. 

## Estimating Normal Means

Suppose we have data from a distribution $$X_i \buildrel \text{iid} \over \sim 
\Ncal(\mu, 1)$$ for some unknown $\mu$ that we seek to estimate. Quite 
reasonably, we might use the sample mean 
$$\overline{X}_n = \frac{1}{n}\sum_{i=1}^n X_i$$
to estimate $\mu$. Clearly, this is an _unbiased_ estimator and it has variance
given by $1/n$, which isn't bad. In general, it's pretty hard to top this. 

We can verify all of this empirically: 

```{r}
#| message: false
library(ggplot2)
library(tidyr)
library(purrr)
library(dplyr)
```

```{r}
compute_mse_sample_mean <- function(mu, n){
    # Compute the MSE estimating mu
    # with the sample mean from n samples
    # We repeat this process a large number of times
    # to get the expected MSE
    R <- replicate(1000, {
        X <- rnorm(n, mean=mu, sd=1)
        mean(X)
    })
    
    data.frame(n=n, 
               mu=mu, 
               bias=mean(R - mu), 
               variance=var(R), 
               mse=mean((R - mu)^2))
}

MU_GRID <- seq(-5, 5, length.out=501)
N <- 10

SIMRES <- map(MU_GRID, compute_mse_sample_mean, n=N) |> list_rbind()
```

Our bias is essentially always zero: 

```{r}
library(patchwork)
p1 <- ggplot(SIMRES, aes(x=mu, y=bias)) + 
    geom_point() + 
    geom_abline(slope=0, 
                intercept=0, 
                color="black", 
                lwd=2, 
                lty=2) + 
    xlab(expression("True Parameter:" ~ mu)) + 
    ylab(expression("Bias")) + 
    ggtitle("Constant Zero Bias of Sample Mean Estimator") + 
    theme_bw()
p2 <- ggplot(SIMRES, aes(x=mu, y=bias^2)) + 
    geom_point() + 
    geom_abline(slope=0, 
                intercept=0, 
                color="black", 
                lwd=2, 
                lty=2) + 
    xlab(expression("True Parameter:" ~ mu)) + 
    ylab(expression("Bias"^2)) + 
    ggtitle("Constant Zero Bias of Sample Mean Estimator") + 
    theme_bw()

p1 + p2
```

Similarly, our bias is small, and constant. Specifically, it is around 
$1/n$ as predicted by standard theory: 

```{r}
ggplot(SIMRES, aes(x=mu, y=variance)) + 
    geom_point() + 
    geom_abline(intercept=1/N, 
                slope=0,
                color="black", 
                lwd=2, 
                lty=2) + 
    xlab(expression("True Parameter:" ~ mu)) + 
    ylab("Variance of Sample Mean") + 
    ggtitle("Constant Variance of Sample Mean Estimator") + 
    theme_bw()
```
As expected, the MSE is the sum of bias and variance, so it's basically just
variance here: 


```{r}
ggplot(SIMRES, aes(x=mu, y=mse)) + 
    geom_point() + 
    geom_abline(intercept=0, 
                slope=0, 
                color="black", 
                lwd=2, 
                lty=2) + 
    geom_abline(intercept=1/N, 
                slope=0,
                color="black", 
                lwd=2, 
                lty=2) + 
    xlab(expression("True Parameter:" ~ mu)) + 
    ylab("Sample Mean MSE ") + 
    ggtitle("Constant MSE of Sample Mean Estimator") + 
    theme_bw()
```

So far, it looks like the sample mean is hard to beat. In particular,
this curve is But... what if we know, *e.g.*, that $\mu$ is positive. We 
might still use the sample mean, but with the additional step that we set
it to zero if the sample mean looks negative. That is, our new estimator is
$$\hat{\mu} = (\overline{X}_n)_+ \text{ where } z_+ = \begin{cases} z & z > 0 \\ 0 & z \leq 0 \end{cases}$$

The $(\cdot)_+$ operator is known as the *positive-part*. How does this 
$\hat{\mu}$ do? 

```{r}
pospart <- function(x) ifelse(x > 0, x, 0)
compute_mse_positive_mean <- function(mu, n){
    # Compute the MSE estimating mu
    # with the positive part of the sample mean from n samples
    # We repeat this process a large number of times
    # to get the expected MSE
    R <- replicate(1000, {
        X <- rnorm(n, mean=mu, sd=1)
        pospart(mean(X))
    })
    
    data.frame(n=n, 
               mu=mu, 
               bias=mean(R - mu), 
               variance=var(R), 
               mse=mean((R - mu)^2))
}

SIMRES_POSPART <- map(MU_GRID, compute_mse_positive_mean, n=N) |> 
    list_rbind()
```


```{r}
ggplot(SIMRES_POSPART, aes(x=mu, y=mse)) + 
    geom_point() + 
    geom_abline(intercept=0, 
                slope=0, 
                color="black", 
                lwd=2, 
                lty=2) + 
    geom_abline(intercept=1/N, 
                slope=0,
                color="black", 
                lwd=2, 
                lty=2) + 
    xlab(expression("True Parameter:" ~ mu)) + 
    ylab("Positive Part Mean MSE") + 
    ggtitle("Constant MSE of Positive Part Mean Estimator") + 
    theme_bw()
```

Not surprisingly, we do _very poorly_ if we are estimating a negative $\mu$
but we assume it is positive. Let's zoom in on the area near 0 however. 


```{r}
SIMRES_POSPART |>
    filter(mu >= -0.5, 
           mu <= 1) |>
ggplot(aes(x=mu, y=mse)) + 
    geom_point() + 
    geom_abline(intercept=0, 
                slope=0, 
                color="black", 
                lwd=2, 
                lty=2) + 
    geom_abline(intercept=1/N, 
                slope=0,
                color="black", 
                lwd=2, 
                lty=2) + 
    xlab(expression("True Parameter:" ~ mu)) + 
    ylab("Positive Part Mean MSE") + 
    ggtitle("Constant MSE of Positive Part Mean Estimator") + 
    theme_bw()
```
Interesting! For some of these values, we do better than the sample mean. 

In particular, we do better in the following scenario: 

- True mean is positive
- Sample mean is negative
- Positive part of sample mean is zero, so closer than pure sample mean

The probability of step 2 (sample mean is negative) is near zero for large
$\mu$, but for $\mu$ in the neighborhood of zero, it can happen.


::: {.callout-note title="Pause to Reflect" collapse="true"}

Review Question: As a function of $\mu$, what is the probability that
$\overline{X}_n$  is negative? You can leave your answer in terms of the
standard normal CDF $\Phi(\cdot)$.

:::

This is pretty cool. We have made an additional assumption and, when that
assumption holds, it helps us or, worst case, doesn't really hurt us much. 
Of course, when the assumption is wrong ($\mu < 0$), we do much worse, but 
we can't really hold that against $(\overline{X}_n)_+$. 

Looking more closely, we can look at the bias of $(\overline{X}_n)_+$: 

```{r}
SIMRES_POSPART |>
    filter(mu >= -0.5, 
           mu <= 1) |>
ggplot(aes(x=mu, y=bias)) + 
    geom_point() + 
    geom_abline(intercept=0, 
                slope=0, 
                color="black", 
                lwd=2, 
                lty=2) + 
    geom_abline(intercept=1/N, 
                slope=0,
                color="black", 
                lwd=2, 
                lty=2) + 
    xlab(expression("True Parameter:" ~ mu)) + 
    ylab(expression("Bias")) + 
    ggtitle("Bias of Positive Part Mean Estimator") + 
    theme_bw()
```

We see here that our improvement came at the cost of some bias, particularly 
in the $\mu \in [0, 1]$ range. But for that bias, we see a good reduction in
variance: 

```{r}
SIMRES_POSPART |>
    filter(mu >= -0.5, 
           mu <= 1) |>
ggplot(aes(x=mu, y=variance)) + 
    geom_point() + 
    geom_abline(intercept=1/N, 
                slope=0,
                color="black", 
                lwd=2, 
                lty=2) + 
    xlab(expression("True Parameter:" ~ mu)) + 
    ylab(expression(Variance^2)) + 
    ggtitle("Non-Constant Variance of Positive Part Mean Estimator") + 
    theme_bw()
```

Here, we see that the variance is less than $1/n$ from $\mu \approx 0.5$ and 
down.  Let's plot variance and bias against each other: 

```{r}
#| warning: false
library(geomtextpath)
SIMRES_POSPART |>
    filter(mu >= 0, 
           mu <= 1) |>
    ggplot(aes(x=bias, y=variance)) + 
       geom_point() + 
       geom_textline(aes(x=bias^2, y=1/n - bias^2),
                     lty=2, color="red4", 
                     label="Breakeven") + 
       ylim(c(0, 0.1)) + 
       theme_bw() + 
       xlab(expression("Bias"^2)) + 
       ylab("Variance") + 
       ggtitle("Bias-Variance Tradeoff for Positive Part Sample Mean")
```
Here, all of the values of $\mu$ corresponding to points _below_ this line are
points where the positive part estimator does better than the standard sample
mean. 


::: {.callout-note title="Pause to Reflect" collapse="true"}

See if you can compute the bias and variance of $(\overline{X}_n)_+$ in closed
form. The [moments of](https://math.stackexchange.com/a/3281287) the 
[Rectified Normal Distribution](https://en.wikipedia.org/wiki/Rectified_Gaussian_distribution)
may be of use. 

:::


Ok - now let's start to generalize. Clearly, the first step is to change the
'positive' assumption. The easiest generalization is to restrict $\mu$ to an
interval $[a, b]$. In this case, it makes sense to replace the positive part
operator with a 'clamp' operator: 

$$(x)_{[a, b]} = \begin{cases} a & x \leq a \\ x & x\in(a, b) \\ b & x \geq b \end{cases}$$

The positive part operator we applied before is $(x)_+ = (x)_{[0, \infty)}$. 


::: {.callout-note title="Pause to Reflect" collapse="true"}

Extend the simulation above to characterize the estimation performance (bias and
variance) of $(\overline{X}_n)_[a, b]$. 

:::

A particularly useful version of this bound is taking $(\overline{X}_n)[-\beta, +\beta]$; 
that is, we don't know the sign of $\mu$, but we know it is less than $\beta$ in
magnitude. This is not an improbable assumption - we often have a good sense of
the plausible magnitude of a parameter (Bayesian priors anyone?) - but it feels 
a bit 'firm'. Can we relax this sort of assumption? We want $\mu$ to be 'not too
big', but we're willing to go big if the data takes us there. 

We can implement this as follows: 

$$\hat{\mu}_{\alpha} = \frac{\overline{X}_n}{1+\alpha}$$

Clearly, setting $\alpha = 0$ gets us back to the standard sample mean. Can this
be better than the sample mean? Let's do the calculations by hand. First we note
that $\E[\hat{\mu}_{\alpha}] = \frac{\mu}{1+\alpha}$ giving a bias of 
$$\text{Bias} = \E[\hat{\mu}_{\alpha}] - \mu = \mu\left(1 - \frac{1}{1+\alpha}\right) \implies \text{Bias}^2 = \mu^2\left(1 - \frac{1}{1+\alpha}\right)^2$$
and 

$$\text{Variance} = \V[\hat{\mu}_{\alpha}] = \frac{1}{(1+\alpha)^2}\V[\overline{X}_n] = \frac{1}{n(1+\alpha)^2}$$

so the total MSE is given by 


$$\begin{align*}
\MSE &= \E[(\hat{\mu}_{\alpha} - \mu)^2] \\
           &= \text{Bias}^2 + \text{Variance} \\
           &= \mu^2\left(1 - \frac{1}{1+\alpha}\right)^2 + \frac{1}{n(1+\alpha)^2}
\end{align*}$$

For suitable $\alpha, n$ this can be less than the standard MSE of $1/n$. For
instance, at $\mu = 5$ and $n = 10$, 

```{r}
shrunk_mean_mse <- function(mu, n, alpha){
    mu^2 * (1 - 1/(1+alpha))^2 + 1/(n * (1+alpha)^2)
}

shrunk_mean_mse(5, 10, 1e-4)
```

Not great - but an improvement! It's actually pretty hard to beat the sample mean
with an estimator of this form in the univariate case, but it can be incredibly
useful in more general settings. 

Let's actually take some time and work this out formally in one dimension:
suppose we have $n$ IID samples from a distribution with mean $\mu$ and variance
$\sigma^2$. The natural estimator of $\mu$ is the sample mean

$$\hat{\mu} = \frac{1}{n}\sum_{i=1}^n X_i$$

It is easy to see that this estimator is unbiased: 

$$\E[\hat{\mu}] = \E\left[\frac{1}{n}\sum_{i=1}^n X_i\right] = \frac{1}{n}\sum_{i=1}^n \E[X_i] = \frac{1}{n}\sum_{i=1}^n \mu = \mu \text{ for all } \mu$$
and the variance follows the usual $1/n$ pattern: 

$$\V[\hat{\mu}] = \V\left[\frac{1}{n}\sum_{i=1}^n X_i\right] = \frac{1}{n^2}\sum_{i=1}^n \V[X_i] = \frac{1}{n^2}\sum_{i=1}^n \sigma^2 = \frac{\sigma^2}{n} \text{ for all } \sigma^2, n$$

so the MSE associated with this estimator is: 

$$\MSE(\hat{\mu}) = \text{Bias}^2 + \text{Variance} = \frac{\sigma^2}{n}$$

Let's compare this to a 'scaled' estimator 

$$\hat{\mu}_{\alpha} = \alpha \hat{\mu} = \frac{\alpha}{n}\sum_{i=1}^n $$
This estimator is no longer unbiased: 

$$\text{Bias}(\hat{\mu}_{\alpha}) = \E[\hat{\mu}_{\alpha}] - \mu = \E[\alpha\hat{\mu}] - \mu = \alpha\E[\hat{\mu}] - \mu = \alpha \mu - \mu = \mu(1-\alpha) \neq 0$$

The variance is 

$$\V[\hat{\mu}_{\alpha}] = \V[\alpha \hat{\mu}] = \alpha^2 \V[\hat{\mu}] = \frac{\alpha^2\sigma^2}{n}$$

These two expressions are interesting: the (squared) bias has clearly increased 
above our initial value of 0, but the variance has gone down by a factor of 
$1 - \alpha^2$. Is there a range of values where this represents an MSE-improvement?

$$\begin{align*}
\MSE(\hat{\mu}_{\alpha}) &= \text{Bias}(\hat{\mu}_{\alpha})^2 + \V[\hat{\mu}_{\alpha}] \\
&= \mu^2(1-\alpha)^2 + \frac{\alpha^2\sigma^2}{n}
\end{align*}$$

This is a quadratic in $\alpha$ (and upward facing / convex), so we know it will
have a unique minimizer. It is unclear whether that minimizer is actually at
$\alpha = 1$ (the 'normal' sample mean). 

Let's take the derivative and set it equal to 0: 
$$\begin{align*}
\frac{\partial}{\partial \alpha}\MSE(\hat{\mu}_{\alpha})
&= \frac{\partial}{\partial \alpha}\left[\mu^2(1-\alpha)^2 + \frac{\alpha^2\sigma^2}{n}\right] \\
0 &= -2\mu^2(1-\alpha) + \frac{2\alpha\sigma^2}{n} \\
  &= -\mu^2 +\alpha\left(\frac{\sigma^2}{n} + \mu^2\right) \\
\implies \alpha_* &= \frac{\mu^2}{\mu^2 + \sigma^2/n}
\end{align*}$$

You can substitute this back into the MSE formula from above to get the
MSE at $\alpha^*$ or recall a basic set of facts about univariate quadratics: 

$$\begin{align*}
f(x) &= ax^2 + bx + c \\
\implies x_* &= -\frac{b}{2a} \tag{Minimizer} \\
f(x_*) &= c - \frac{b^2}{4a} \tag{Minimum}
\end{align*}$$

If we expand the MSE expression into this form: 

$$\begin{align*}
\MSE(\hat{\mu}_{\alpha}) &= \mu^2(1-\alpha)^2 + \frac{\alpha^2\sigma^2}{n} \\
&= \mu^2 - 2\mu^2\alpha + \mu^2\alpha^2 + \frac{\alpha^2\sigma^2}{n} \\
&= \underbrace{\left(\mu^2 + \frac{\sigma^2}{n}\right)}_a\alpha^2 -\underbrace{2\mu^2}_b \alpha + \underbrace{\mu^2}_c
\end{align*}$$

so we have a minimum possible MSE of: 

$$\begin{align*}
\MSE^* &= c - \frac{b^2}{4a}  \\
&= \mu^2 - \frac{(-2\mu^2)^2}{4\left(\mu^2 + \frac{\sigma^2}{n}\right)} \\
&= \mu^2 - \frac{4\mu^4}{4(\mu^2 + \sigma^2/n)} \\
&= \mu^2 - \frac{\mu^4}{\mu^2 + \sigma^2/n} \\
&= \mu^2 - \frac{\mu^2}{\mu^{-2}(\mu^2 + \sigma^2/n)} \\
&= \mu^2 - \frac{\mu^2}{1 + (\sigma/\mu)^2/n} \\
&= \mu^2\left(1-\frac{\mu^2}{1 + (\sigma/\mu)^2/n}\right)
\end{align*}$$

Whew! That was a lot of algebra. Let's do a quick check to make sure this
actually holds. Without loss of generality, let's take $\sigma^2 = 1$ and 
$n = 4$ and vary $\mu$: 

```{r}
MU_GRID <- seq(-3, 3, length.out=201)

SIM_RESULTS <- map(MU_GRID, \(mu){
  replicate(1000, {
    n <- 4
    sd <- 1
    X <- rnorm(n=n, mean=mu, sd=sd)
    mu_hat <- mean(X)
    
    alpha_star <- mu^2/(mu^2 + sd^2/length(X))
    
    mu_hat_alpha <- alpha_star * mu_hat
    
    data.frame(mu=mu, 
               n=n, 
               sigma=sqrt(sd), 
               alpha_shrink = alpha_star, 
               alpha_no_shrink = 1, 
               mu_hat_shrink = mu_hat_alpha, 
               mu_hat_no_shrink = mu_hat, 
               theory_mse_shrink = mu^2  - mu^4 / (mu^2 + sd^2/n),
               theory_mse_no_shrink = sd^2/n)
  }, simplify=FALSE) |> bind_rows()
}) |> bind_rows()

SIM_RESULTS |> 
    mutate(err_shrink = mu_hat_shrink - mu, 
           err_no_shrink = mu_hat_no_shrink - mu,
           err2_shrink = err_shrink^2, 
           err2_no_shrink = err_no_shrink^2) |>
    group_by(mu) |> 
    mutate(var_shrink = var(err_shrink), 
           var_no_shrink = var(err_no_shrink)) |>
    select(-mu_hat_shrink, -mu_hat_no_shrink, -n, -sigma) |>
    summarize(across(everything(), mean)) |>
    pivot_longer(cols=-mu, 
                 names_to="metric", 
                 values_to="value") |>
    mutate(scaling = !str_detect(metric, "no_shrink"), 
           facet_name = case_when(
               str_detect(metric, "var") ~ "Variance", 
               str_detect(metric, "alpha") ~ "Scale Factor", 
               str_detect(metric, "theory") ~ "Theoretical MSE", 
               str_detect(metric, "err2") ~ "Observed MSE", 
               str_detect(metric, "err") ~ "Bias"
           )) |>
    group_by(scaling, facet_name) |>
    mutate(value_smooth = runmed(value, k=9, endrule="constant")) |>
    ggplot(aes(x=mu, color=scaling)) + 
    geom_point(aes(y=value), alpha=0.2) + 
    geom_line(aes(y=value_smooth)) + 
    theme_bw() + 
    theme(legend.position="bottom") + 
    xlab(expression(mu)) + 
    facet_wrap(~facet_name, scales="free_y") + 
    scale_color_brewer(type="qual", 
                       palette=2,
                       name="Scale Factor Applied")
```

Lots of interesting things here: 

- Both the theoretical and the actual MSE of the unscaled (plain sample mean)
  estimator are basically constant and does not depend on $\mu$. (Fluctuations
  in observed MSE are due to sample randomness.) This makes sense as the sample 
  mean is location invariant. 

TODO

## James-Stein Estimation of Multivariate Normal Means

TODO
