---
session: "`r session <- 9; session`"
topic: "`r library(tidyverse); topic <- read_csv('../key_dates.csv', name_repair='universal') |> filter(Course.Element == 'Class Session', Item.Number==session) |> pull(Details); topic`"
date: "Last Updated `r Sys.Date()`"
---

{{< include _latex_defines.qmd >}}

To this point, we have created individual models for regression and
classification tasks. We have covered regression-type models for
low-dimensional settings (GLMs: mainly linear and logistic), generative
models for classification (NB, LDA, QDA), purely discriminative models
(SVM), and flexible non-parametric models (KNN). Building on these tools,
we have developed non-linear extensions (splines and kernels) suitable
for the $n \gg p$ setting and regularization techniques (ridge and lasso)
necessary for the $p \gg n$ setting. 

In each of these, we have argued that certain models (more precisely,
certain estimators) have different amounts of bias and variance and
given a rule of thumb that in low-data settings high-bias/low-variance
models are preferred while in high-data settings, we should feel free
to use low-bias models since the variance will be controlled by the
large training set. 

Furthermore, we have discussed the importance of 'hold-outs' (test/train
splits, cross-validation, *etc.*) as a tool to avoid overfitting. If we
evaluate our models on new data (or at least new to that model), we 
minimize the chance of 'regurgitation' and ensure that our models are 
learning true(-ish) generalizabile relationships. Importantly, we know
that _any time_ we make a selection, it is important to have a new data
set for downstream tasks. So: 

- If we just want to to train a single pre-specified model with no 
  hyperparameters and use it blindly, we can use all of our data
- If we want to to train a single pre-specified model with no 
  hyperparameters and assess its predictive performance, we need to
  split our data into two sets (training + test)
- If we want to train a model and select hyperparameters but do not
  need to assess its predictive performance, we need to split our
  data into two sets (training + 'validation')
- If we want to train a model, select hyperparameters, and provide
  an unbiased assessment of predictive performance, we need to split 
  our data into three sets (training + validation + test)
  
Personally, I find the jargon of these different sets a bit confusing--what
is testing and how is it different than validation?--but the 'new data
for each task' heuristic is easy to apply and generalize. 

Finally, we have noted that - while hold out techniques are an excellent tool
for maximizing predictive performance - they aren't the best to ensure 
*reliability of interpretation*. If *insights* are more important than pure
*performance*, stability techniques are typically a better strategy for 
tuning hyperparameters. 

For the next two weeks, we are going to go beyond this 'single model' 
paradigm and consider how we might handle scenarios where we have two
or more models available for our use. This is the domain of *ensemble 
learning* - building models consisting of multiple 'sub-models', termed
'base learners'. Ensemble learning is particularly powerful when 
applied in conjunction with fast and flexible base learners, so we will
also use this as a jumping off point for one more class of models. 

## Ensemble Learning

We begin with a discussion of *ensemble learning*, the task of combining
multiple (simpler) models into a 'super model' to achieve a task. 
Ensembles can be used for essentially any ML task, but we will focus on 
ensemble learning for the two tasks we have focused on to date: regression
and classification. (The differences in these two tasks will not really
have any bearing on our 'ensembling' discussion.)

Suppose that we want to predict the outcome of some variable $Y$ using
two *statistically independent* predictors $\hat{Y}_1, \hat{Y}_2$: if
*arguendo* we assume these are unbiased, the MSE of each prediction
is simply

$$\begin{align*}
\MSE_{\hat{Y}_i} &= \E_{Y, \hat{Y}_i}[(Y - \hat{Y}_i)^2] \\
                       &= \E_{Y, \hat{Y}_i}[(Y - \E[Y] + \E[Y] - \hat{Y}_i)^2] \\
                       &= \E_{Y, \hat{Y}_i}[(Y - \E[Y])^2 +2(Y - \E[Y])(\E[Y] - \hat{Y}_i)+(\E[Y] - \hat{Y}_i)^2] \\
                       &= \E_{Y}[(Y - \E[Y])^2 +2\E_{Y, \hat{Y}_i}[(Y - \E[Y])(\E[Y] - \hat{Y}_i)]+\E_{\hat{Y}_i}[(\E[Y] - \hat{Y}_i)^2] \\
                       &= \sigma_Y^2 + 2\E_{Y}[Y - \E[Y]]\E_{\hat{Y}_i}[\E[Y] - \hat{Y}_i)] + \sigma^2_{\hat{Y}_i} \\
                       &= \sigma_Y^2 + \sigma^2_{\hat{Y}_i}
\end{align*}$$

(Why do the cross terms vanish?) That is, the MSE is simply sum of the 
'irreducible error' associated with the best possible prediction ($\sigma_Y^2$
against the optimal prediction $\E[Y]$) and the variance (error) in $\hat{Y}_i$
as an estimator of $\E[Y]$.

If $\hat{Y}_1, \hat{Y}_2$ are both predictors of $Y$, it is natural to ask
if we can do better using a *combination* of the two of them than we can with
either separately. Indeed, if we let $\widehat{Y} = \frac{1}{2}(\hat{Y}_1 + \hat{Y}_2)$, the above analysis tells us: 

$$\begin{align*}
\MSE_{\widehat{Y}} &= \sigma_Y^2 + \sigma_{\widehat{Y}}^2 \\
&= \sigma_Y^2 + \V\left[\frac{\hat{Y}_1 + \hat{Y}_2}{2}\right] \\
&= \sigma_Y^2 + \sigma_{\hat{Y}_1}^2/4 + \sigma_{\hat{Y}_2}^2/4
\end{align*}$$

How does this compare to the separate predictions? Well, if 
$\sigma_{\hat{Y}_1} = \sigma_{\hat{Y}_2}$, then we have clearly reduced our 
expected MSE by half of the variance error. (By definition, we can't do anything
about the irreducible error.) If, on the other hand, one predictor is much
better than the other, say $\sigma_{\hat{Y}_1}^2 = 400 \sigma_{\hat{Y}_2}^2$,
then we have

$$\begin{align*}
\MSE_{\hat{Y}_1} &= \sigma_Y^2 + 400 \sigma_{\hat{Y}_2}^2 \\
\MSE_{\hat{Y}_2} &= \sigma_Y^2 + \sigma_{\hat{Y}_2}^2 \\
\MSE_{\widehat{Y}} &= \sigma_Y^2 + 100.25 \sigma_{\hat{Y}_2}^2 \\
\end{align*}$$

So we don't beat the superior predictor ($Y_2$), but we comfortably beat the
inferior predictor ($Y_1$). Notably, if we didn't know which predictor
was better and had to select randomly (50/50), we would have an expected
MSE of $\sigma_Y^2 + 200\sigma_{\hat{Y}_2}^2$, so the 'averaging' predictor
is still better. More generally, if $\hat{Y_1}$ is sometimes better and
$\hat{Y_2}$ is better at other times, a suitable averaging strategy will 
do better  *in the long run* than using a single predictor. (We should pick
the averaging weights as a function of the variance of the two approaches and
the relatively frequencies of the $\hat{Y}_1$-better and $\hat{Y}_2$-better
scenarios.)

This example - simple as it is - gets at the core insight of stacking: if we
have several 'good enough' models, we can do better - sometimes much better -
by using a combination of them. This is not guaranteed - if $\hat{Y}_2$ is 
always the 'lower variance' model, adding in some $\hat{Y}_1$ pretty much
always hurts - but we do not expect to have a single 'dominant' model. 
(See earlier discussion of the *No Free Lunch* Theorem.) In 
practice, we typically find ourselves armed with models of similar-enough
performance (logistic regression, linear SVMs, RBF Kernel SVMs, RBF Kernel 
logistic regression) and find that different models perform better on different
inputs. *E.g.*, the 'linear' models might do well far from the decision boundary,
while the 'kernelized' models might improve performance near the decision
boundary while suffering from extra 'wiggliness' (variance) far from the 
decision boundary. 

So what can we take away from this discussion? 

> Low correlation among predictors is helpful. If our base learners give
  the exact same predictions at every point, there's nothing to be gained by 
  combining and comparing them. 

Put another way, ensemble learning benefits from a *diverse* set of base 
learners. We hope to 'combine strength' from different approaches to build
an ensemble predictor that is better than any of its individual components.

The three major ensembling techniques we will cover - stacking, bagging, and 
boosting - essentially come down to different ways of getting this diversity.

## Stacking

So this is all well and good, but where do we *actually get* different 
predictors? Here, we're going to change notation to $\hat{f}_1, \hat{f}_2, \dots
\hat{f}_K$ to emphasize that we want to learn a combination of _predictors_ 
(functions estimated from data), not _predictions_. 

Well... perhaps the easiest thing to do is to train different models on the
same data. Logistic Regression, LDA, and SVMs will likely find _similar_ 
decision boundaries, but they won't be exactly the same. Suppose we have

- $\hat{f}_{\text{LR}}$: Logistic Regression (for simplicity, plain and
  unregularized)
- $\hat{f}_{\text{LDA}}$: Linear Discriminant Analysis
- $\hat{f}_{\text{SVM}}$: A Linear SVM with the default hyperparameter 
  from `sklearn`
  
Further, let's assume that we have the binarized predictions ($\{0, 1\}$, no
soft labels for now) from each model. How can we combine these? 

Perhaps the simplest rule is the 'majority vote' rule: if $\hat{f}_{\text{LR}}(\bx) = \hat{f}_{\text{LDA}}(\bx) = 1$ and $\hat{f}_{\text{SVM}}(\bx) = 0$, then $\hat{f}_{\text{Majority}}(\bx) = 1$. 

This approach is quite easy to implement, but it treats all the inputs as
equally reliable. While this isn't the _worst_ assumption, we can do better.
In particular, what if we _learned_ an optimal set of weights. Specifically, 
we want to learn a vector of weights $\bw \in \R^3$ to maximize
predictive accuracy of the linear combination $w_1 \hat{f}_{\text{LR}} + w_2 \hat{f}_{\text{LDA}} + w_3 \hat{f}_{\text{SVM}}$. This gives us *another*
classification problem, so let's use logistic regression to determine the 
combination weights:

$$\hat{\bw} = \argmin_{\bw} \sum_{i=1}^n -y_i \bw^{\top}\hat{\bf} + \log(1 + e^{\bw^{\top}\hat{\bf}})$$
where $\hat{\bf}$ is a vector of the three base learner predictions.

The solution to this problem, $\hat{\bw}$, gives the _stacking weights_
for our three base learners. Once we have $\hat{\bw}$, we can make
predictions at our our new test point, $\tilde{\bx}$, by: 

1. Apply three base learners separately: 
   - $\hat{y}_{\text{LR}} = \hat{f}_{\text{LR}}(\tilde{\bx})$
   - $\hat{y}_{\text{LDA}} = \hat{f}_{\text{LDA}}(\tilde{\bx})$   
   - $\hat{y}_{\text{SVM}} = \hat{f}_{\text{SVM}}(\tilde{\bx})$
2. Determine the weighted combination: 
   - $\text{Stacking Score} = w_1\hat{y}_{\text{LR}} + w_2\hat{y}_{\text{LDA}}+w_3\hat{y}_{\text{SVM}}$
3. Make a prediction using the stacking score: 
   $$\hat{y}_{\text{Ensemble}} = \begin{cases} 1 & \text{ if } \text{Stacking Score} > \theta \\ 0 & \text{ if } \text{ Stacking Scoore} \leq \theta \end{cases}$$
   where $\theta$ is a threshold chosen to minimize the problem-specific loss.
   (Alternatively, set $\theta = 0$ and include an intercept term.)
   
We note a few practical points here: 

1. What data should we use to estimate $\bw$? This is a new step in our pipeline,
   so we need a new 'chunk' of our data. This split of the data is typically
   called either a query or validation set, but I tend to describe it as an
   'ensembling' set to make its purpose clear. Regardless, it serves to 
   give us a new data set to fit our new ensemble model. (If we use the
   training set used to train each $\hat{f}_i$, the base learners may look
   'too good' to the ensemble learning process.)
2. Nothing actually requires us to use the binarized predictions. For models
   that provide soft labels (class probabilities), we can and should use those
   in the ensemble process (both in training the ensemble weights and in
   constructing the predictor vectors $\hat{\bf}$). Dropping the
   probabilities is really just throwing away data. 
3. We motivated stacking by weighted averages. As such, it is common to add
   additional constraints to the stacking problem to make it look more like
   averaging. For example, you will often see: 
   - $\bw \geq 0$: The stacking weights are non-negative. This makes sense
     if you assume each predictor is generally reasonable. (If you have a 
     prediction that is *anti-correlated* with your response, you should probably
     pause and look into that further first.) 
   - $\sum w_i = 1$: The stacking weights sum to 1. Again, a natural
     generalization of averaging. 
4. When fitting *large* ensembles (like any large model), it is not uncommon
   to see ridge or lasso penalties. These are *particularly important* in ensemble
   building since the base learners are generally *highly correlated*. 
5. Suitably tuned (and that's a big assumption!) stacking should never do worse
   than any individual predictor because we can always pick $\bw = (1, 0, 0)$ 
   if the first base learner strictly dominates the others. 
   
Stacking creates a 'model of models' - the turducken of machine learning - and 
all of the ML tricks we have learned to date can be used in creating this
meta-model. Like any modeling step, we have to practice good data hygiene
(splitting and hold outs) to make sure we don't overfit and to generate a good
estimate of predictive accuracy, but otherwise it is particularly straightforward.

An interesting fact about stacking is that it treats each of the base learners
$\hat{f}_i$ as black boxes and doesn't really need to know what is inside them
or where they came from. This lets us use stacking as a 'work-around' for certain
types of practical problems: 

- the models are all pre-trained and we don't have the ability to alter
  them (*e.g.*, building an ensemble that 'averages' out the advice from ChatGPT,
  Google Gemini, and Anthropic)
- we are unable to move the data from a restricted setting for privacy
  reasons (data trained on patient records held by different medical providers)
  
Stacking can also be used as a form of 'post-processing' to introduce additional
desiderata that the original model training process was not aware of. I have
[explored this in the fairness context](https://arxiv.org/abs/2206.00074) with 
some collaborators. Here, we consider a situation in which the base learners
are learned to only maximize accuracy, but we include fairness constraints in
the stacking process, effectively eliminating models that are highly predictive,
but whose predictive power comes from systemtic biases. 

Despite these strengths, stacking can only take us so far: the value-added of 
stacking comes from differences between the base learners and, when trained on 
the same data set, we don't expect too much diversity among the base learners. 
In practical settings, stacking is particularly useful when we are given 
'standard' or 'baseline' models for a task that we want to combine without 
changing the individual models, but for real improvements we may want to modify
our pipeline to explicitly generate more diversity. 

This brings us to our next family of ensembling techniques - resampling methods,
the most famous of which is *bootstrap aggregation* or bagging. 

## Resampling

In our discussion above, we argued that diversity of (variance between) base 
learners was key to ensemble performance. When these base learners are trained
on the same training data, we can only get so much variance. Since most of our
models are not terrible, they generally pick out the same major patterns in
the data. 

So how can we get more variance? We could further split our training data, using
a small chunk for each model, but this seems likely to bring about variance
problems. If we want to fit large ensembles of 10, 20, or even 100 models, we
might wind up using less than 1% of the overall data to train a single model,
which is clearly subpar. Stacking might give us some power at the margin, but
it's hard to make up for the weakness associated with using such a small portion
of our data. 

We can get around this problem using _sampling_ or _resampling_ techniques.
We can train our esnemble members on randomly selected subsets of our data 
(or subsets of the features). Because these models have different training sets,
they will be more varied. Typically, this sort of 'data randomization' induces
more variance than just changing hyperparamters or model families.)

These randomization schemes have different names: 

- Minibatching: training on random subsets of samples (rows)[^mb]
- Random Subspace Method: training on random subsets of features (columns)
- Minipatching: training on random subsets of samples and features 
  (row and columns)
  
[^mb]: Most of these methods were actually designed to speed up 
training/fitting and that remains the main use case for minibatching today
(*cf.* *stochastic* gradient descent or certain dropout schemes). We won't 
focus too much on implications for speed today, but it certainly is nice when a 
technique invented to solve one problem turns out to be useful in other scenarios. 
  
If the model is not too sensitive to the sample size (or if we use a high
sampling rate), these strategies can work well. But we have a bit
of a bind here: if we want to use large training sets for each learner, we go
back to the overlapping training scenario we were trying to avoid. 

Can we be a bit more creative on our sampling? We want to minimize overlap
but also get 'full sized' training sets. 

### Bootstrapping

To get around this problem, we rely on a key idea of late 20th century 
statistics: **the bootstrap.**[^bootstrap] Recall the core idea of bootstrapping: 

- We have samples $\{(\bx_1, y_1), (\bx_2, y_2), \dots, (\bx_n, y_n)\}$ from
  an unknown distribution. 
- We would like to have more samples from this distribution, but we don't have
  the ability to gain more data. 
- If our data set is large, we expect that the _empirical distribution_
  $\hat{\P}_n$ will be close to the 'real' distribution $\P$. Formally, the law
  of large numbers (and its fancy variants like Glivenko-Cantelli) guarantees
  us that $\hat{P}_n \to \P \text{ as } n \to \infty$.[^lln]
- Sampling from $\hat{P}_n$ is 'close' to sampling from $\P$

[^lln]: The Law of Large Numbers gives us a pointwise convergence here, while
Glivenko-Cantelli is needed to give us *uniform* convergence guarantees. The 
distinction matters, but we're not going to use it here. 

So how can we sample from the empirical distribution $\hat{P}_n$? We sample
_with replacement_. That is, we pick one of the training points $n$ times 
**independently** (with no regard for prior selections). Intuitively, this
captures the idea of IID sampling - it's also necessary so we don't just 
reproduce the (shuffled) training data.[^replacement]

[^replacement]: If this seems weird, step back from thinking of resampling data
and just think about sampling from a distribution. If you are sampling from a
$\text{Binomial}(n=5, p=0.5)$ distribution, you will get *several* `3`s in a 
data set of reasonable size. That's just a fact of life with discrete 
distributions (it's a "probablility zero" event in the continuous case) and we
take the empirical distribution $\hat{P}_n$ as a discrete uniform distribution
supported on the training set. This is why it's so important to think of 
bootstrapping as sampling from *a distribution* not as sampling from *the data*
(even though that distribution is estimated from the data). 

This scheme - sampling with replacement - is the essence of 
**bootstrap sampling**. In STA 9719, you will see how it can be used to do
things like estimate sampling variability of complex statistics. Here, our goal
is a bit more straightforward: we just want a way to 'shake things up.'

### Bootstrap Aggregation (Bagging)

We can use a similar strategy to generate 'new' training data for our base
learners: 

- Repeat $b=1, \dots, B$ times: 
  - Sample from the training data $\Dcal_{\text{train}}$ $n$ times 
    *with replacement* to get a bootstrap set $\tilde{\Dcal_b}$
  - Train a base learner $\hat{f}_b$ on $\tilde{\Dcal_b}$
- Create the ensemble predictor: 
  $$\hat{f}(\bx) = \frac{1}{B}\sum_{b=1}^B \hat{f}_b(\bx)$$
  
This ensembling strategy is called **bootstrap aggregation** or more simply
**bagging**. 

So how much diversity / variance can we actually expect from this strategy? It
depends on how much overlap we get in our sampling. This is a fun little
probability exercise: 

::: {.callout-tip title="Probability of a Given Sample Being in a Bootstrap Data Set"}

In our scenario above, what is the chance that $\bx_1$ is in
$\tilde{\Dcal}_b$? 

For each sample, there is a $1/n$ chance that we select $\bx_1$, and hence
a $1-1/n$ chance that we don't. If we repeat this process $n$ times, there is
a $(1-1/n)^n$ chance that we never select $\bx_1$. For a large data set, 
this converges to

$$\lim_{n\to\infty}\left(1-\frac{1}{n}\right)^n = e^{-1} \approx \frac{2}{3}$$

so about $2/3$ of our data is used for each base learner. _Unlike_ straight
subsampling, however, this process includes repeats, so we have $n$ samples, 
guaranteeing us repeats. 

A more detailed analysis can show that the _number_ 
of times a sample appears is asymptotically Poisson: rare events ($1/n$) and
large samples ($n \to \infty$) strike again. 

:::

So how does bagging actually work? We know that we're always trying to control
bias and variance so let's look at those two terms separately: 

- Bias: Because we are fitting the same model $B$ times, there's really no
  impact on bias. (*E.g.*, if we are fitting a linear model, the average of
  $B$ lines is just another line so whatever approximation error we have is
  unchanged.)
- Variance: Suppose that the base learners $\hat{f}_i$ have variance $\sigma^2$
  and correlation $\rho$. (These are constant for all $\hat{f}_i, \hat{f}_j$ by
  the IID sampling structure used). Then the variance of $\hat{f}$ (the bagged 
  average) is given by 
  
$$\begin{align*}
\V[\hat{f}] &= \V\left[\frac{1}{B}\sum_b \hat{f}_b\right] \\
            &= B^{-2}\V\left[\sum_b \hat{f}_b\right] \\
            &= B^{-2}\left(\sum_{b=1}^B \V[\hat{f}_b] + \sum_{\substack{b, b'=1 \\ b \neq b'}}^B \C[\hat{f}_b, \hat{f}_{b'}]\right)\\
            &= B^{-2}\left(\sum_{b=1}^B \sigma^2 + \sum_{\substack{b, b'=1 \\ b \neq b'}}^B \rho\sigma^2\right) \\
            &= B^{-2}\left(B\sigma^2 + (B^2 - B) \rho\sigma^2\right) \\ 
            &= B^{-2}\left(B(\rho\sigma^2 + (1-\rho)\sigma^2) + (B^2 - B) \rho\sigma^2\right) \\ 
            &= B^{-2}\left(B\rho\sigma^2 + B(1-\rho)\sigma^2) + B^2\rho\sigma^2 - B\rho\sigma^2\right) \\ 
            &= B^{-2}\left(B(1-\rho)\sigma^2) + B^2\rho\sigma^2\right) \\ 
            &= \rho\sigma^2 + \frac{1}{B}(1-\rho)\sigma^2
\end{align*}$$

This is an interesting formula. As always, we can understand it best by taking
the extremes: 

- $B = 1$: In this case, the variance reduces to $\sigma^2$, the variance of a 
  single predictor (as we would expect for 'ensemble of one'). 
- $B \to \infty$: The variance decreases to $\rho\sigma^2$, but no further. 
  Practically, we see diminishing returns in the $B\approx 200-500$ range, which
  you will see as the default for most software that has a bagging step. 
- $\rho \to 1$: As the base learners become more correlated, the variance
  converges to $\sigma^2$, the variance of a single predictor. As we would
  expect since we have no diversity among base learners when $\rho \to 1$. 
- $\rho \to 0$: As the base learners become uncorrelated, the variance converges
  to $\sigma^2/B$, the variance of a standard average. (You might ask why not
  take $B \to \infty$ in this case to have no bias: we really can't get an 
  infinite number of uncorrelated[^uncor] models, no matter how hard we try, 
  if they are modeling the same underlying DGP.)
  
[^uncor]: Well, *technically* we can get a bunch of uncorrelated models if we
are willing to accept models that have no predictive power, *e.g.*, a bunch of
distinct random number generators. The reasons this is a bad strategy should be
obvious.
  
So we see we want a *flexible* (low-bias) base learner that is also going to give
us low correlation (and low variance) among the bagged predictors. How can we
get such a base learner? We'll cover this more below. 

[^bootstrap]: If you have seen bootstrapping before, it was likely in the
context of estimating the sampling variance of an estimator. (If you haven't
seen this, go look it up - it's awesome!) Here, we're using bootstrap sampling
to induce some variance, not to estimate variance, but the sampling strategy
is the same. 

## Summary of Bagging

To review: e discussed the ensemble learning technique known as 
*bagging*. Bagging, a contraction of 'boostrap aggregation' can be used
to create an ensemble by training $B$ copies of a given base learner on $B$
distinct bootstrap samples, *i.e.*, 'faux' data sets of $n$ points obtained
by sampling the original training data $n$ times _with replacement_. 
Because these $B$ trained base learners are essentially interchangeable, 
it is conventional to create the bagged prediction by a straight average
(or majority vote if classification) of the individual base learners, though
you will occasionally see the bagging base learners combined in a 'stacked'
fashion. 

We showed that the MSE associated with a bagged ensemble of size $B$ is: 

$$\MSE_{\text{Bagging}} = \text{Bias}^2_{\text{Base Learner}} +\underbrace{\rho\sigma^2 + \frac{1}{B}(1-\rho)\sigma^2}_{\text{Variance}} + \text{Irreducible Error}$$

and noted that, by sending $B \to 0$, we reduce the 'bootstrapping variance'
to 0 but that some 'inherent variance' remains from the training
data.[^training_variance] As we noted last week, this implies that our
search for the optimal base learner should do three things: 

- Have low bias
- Have low correlation among the bagged learners
- Best fast enough that we can take $B \to \infty$ and cancel that
  term
  
If we can find something that does all of these, we will have a (near)
optimal predictor ($\MSE \approx \text{Irreducible Error}$). This
is our primary goal for the day. 

[^training_variance]: There is randomness in our selection / acquisition of
training data and nothing we do can completely remove it. Three cheers for
large and well-constructed training sets, the unsung heros of machine
learning!

### 'Out of Bag' Error

Before we move to our long-promised final base learner, let's consider
another advantage of bagging. We know that it is important to have 'new'
data when assessing model performance. To date, we have obtained this 'new'
data through variants of the hold-out principle, such as the test-train
split or cross-validation: if it is important to have new data, it is
important enough to set aside some of our putative training data for model
assessment. 

But bagging gives us another source of 'unseen' data: if we sample $n$
points from a population of size $n$ with replacement, approximately
$n/e \approx 2n/3$ points will be sampled, leaving $n(1-e^{-1}) \approx n/3$
points unseen by each base learner. If these points were the same across
each base learner, we would have a readily available test set. But then
our whole bagging strategy would be essentially useless... 

We can estimate the so-called 'out-of-bag' error (OOB) associated with
each training point by computing the predictive performance 
_on the ensemble of learners that haven't seen that point_. That is, 
instead of testing the performance of the whole ensemble, we take a 
'subensemble' of the base learners that never saw that point. Repeating
this process over all $n$ training points (and creating $n$ different
sub-ensembles) we obtain an estimate of the OOB error. Clearly, OOB is not
test error of the entire ensemble, but it is a useful quantity nevertheless.

Formally, we obtain the bagged ensemble and associated OOB error as follows: 

- **Inputs**:
  - Base learner family $\Fcal$
  - Training set of size $n$: $\Dcal=\{(\bx_1, y_1), (\bx_2, y_2), \dots, (\bx_n, y_n)\}$
  - Number of bootstrap samples $B$
- For $b=1, \dots, B$:
  - Create bootstrap training set $\Dcal_b = \{(\bx_1^{(b)}, y_1^{(b)}), (\bx_2^{(b)}, y_2^{(b)}), \dots, (\bx_n^{(b)}, y_n^{(b)})\}$
  - Fit base learner $\hat{f}_b$ to $\Dcal_b$
- Build ensemble $\hat{f}(\bx) = B^{-1}\sum_{b=1}^B \hat{f}_b(\bx)$
- For $i=1, \dots, n$: 
  - Let $b_i^c = 0$ be the number of bootstrap samples not containing $i$
  - Let $y_i^c = 0$ be the average prediction of bootstrapped learners not trained on $i$
  - For $b = 1, \dots, B$: 
    - If $i \notin \Dcal_b$: 
      - $b_i^c := b_i^c+1$
      - $y_i^c := y_i^c + \hat{f}_i(\bx_i)$
  - Let $\hat{y}_i^c = y_i^c / b_i^c$ (or majority vote if classifying)
  - Let $\text{OOB}_i = \text{Loss}(\hat{y}_i^c, y_i)$
- Estimate OOB error as $\text{OOB} = n^{-1} \sum_{i=1}^n \text{OOB}_i$
- **Return** $\hat{f}, \text{OOB}$

For each training point, we expect it to occur in approximately $B/e$ 
bootstrap samples and hence we have approximately $1 - B/e \approx B/3$
bootstrapped base learners that we can use to computing the OOB error. This isn't 
quite as good as the full $B$ samples we would use for a true test point, but 
since variance decreases quadratically in $B$, this isn't really a major 
loss.[^oob_var]

[^oob_var]: Taking a typical choice of $B = 500$, the OOB ensemble of 
(expected) $316$ base learners will have a variance that is approximately
$500/316 \approx 1.6$ times higher than the 'true' ensemble. If $B$ is
large enough $(1-\rho)\sigma^2/B \approx 0$,
$(1-\rho)\sigma^2/(B(1-e^{-1})) \approx 1.6(1-\rho)\sigma^2/B$ will 
also be negligible. Put another way, the bulk of the error in bagged
ensembles comes from i) bias; ii) _training data_ variance; and iii) 
irreducible error. The 'not enough bootstraps' contribution is typically
quite small.

## Decision Trees

Ok - now it's time to consider the final base learner of this course: 
_decision trees_. You are almost certainly already familiar with 
decision trees through diagrams like this: 

```{r}
#| echo: true
library(rpart)
library(rpart.plot)
library(MASS)
dtree <- rpart(medv ~ . , data=Boston)

rpart.plot(dtree)
```

Here, the tree is evaluated by applying a series of _univariate[^ob] binary_
decision rules 'downwards': that is, at each step, we ask one step about
one variable and take the left or right branch as appropriate. The final
terminus (or 'leaf') of the tree is then our prediction.

[^ob]: This model can be generalized to decisions like $3x_1 + 2x_2 < 5$ instead
of simple univariate rules like $x_1 < 2.5$. These sorts of trees, often called
*oblique* trees because they have decision boundaries that are not axis-aligned,
are rather more difficult computationally, so you won't see them used as much,
but they are popular among theory-minded researchers. 

Decision trees are incredibly popular because they are easy to apply
'mentally' and suggest natural interpretations. Variables which are
used to split early or split often are more important. Interestingly, 
decision trees also can be used to give an 'intermediate prediction'
which is what we would get by stopping the prediction at that point. 
Splits which materially change the intermediate prediction are 
also more important. 

But how should we *actually fit* a decision tree to data? This is, like
bagging and computing OOB error, the sort of thing which is easy to 
understand intuitively, but comes with a significant amount of bookkeeping
in practice. 

Essentially, trees are fit by repeatedly asking the question: "What split
would most reduce the error if my predictions stop here?" So, for the 
first split, the tree looks at each variable to determine its optimal
split point, and then finds the optimal variable to split on. To find
the optimal split point, we need to sort the data on that variable
and then compute the 'before' and 'after' variance for each value. 
After finding the first split, the two 'second level' splits repeat
this process _on only the data in their path_. That is, the two 'second
level' splits are found on _different_ data. The four 'third level' splits
are found on four different data sets, *etc.* 

Because the later splits are computed on different data sets, they 
almost always select different variables at each level (as in our
example above, where the two 'level two' splits use the `lstat` and `rm`
variables). Predictions are made by taking the average (or majority)
of data left in that split. 

Note that the splits need not be _and typically are not_ 'even' splits
along a variable. *E.g.*, if we were doing a split on a variable that
looks like

| Variable | Value | 
|----------|-------|
| 1.0      | 1.0   | 
| 1.1      | 0.9   | 
| 1.2      | 1.1   | 
| 1.3      | 1.4   | 
| 100      | 500   | 
| 105      | 550   | 

The optimal split is clearly something like $\texttt{Variable} < 1.3$ with
conditional predictions of $1.1$ and $525$ in each split and split sizes
of $4$ and $2$. 

Clearly, if the tree is allowed to run 'to completion', there will be
one data point left in each leaf and the tree can obtain 0 training error. 
That is, maximal decision trees are a very high complexity (low bias, 
high variance) model. On the other extreme, a 'single split' tree (often
called a 'stump') will be low complexity (high bias, low variance). So
we can take tree depth (number of levels) as a good proxy for model
complexity. Like $K$-Nearest Neighbors, we can pick depth (complexity)
by a suitable cross-validation or train-test scheme, but will soon see
a better way. 

Some other notes on trees: 

- You will sometimes see decision trees refered to as `CART`
  (Classification and Regression Trees) or `rpart` (Recursive Partitioning)
  but most folks will know the term "decision trees" so I just use that. 
- Trees extend _very naturally_ to categorical variables and do not require
  numeric encoding: we simply have to adapt the decision rule from 
  $x \leq y$ to $x \in \Ycal$ where $x$ is now a categorical feature
  and $\Ycal$ is a set of lables. 
- Trees also handle missing values nicely: we can simply treat `NA` as
  _very high_ or _very low_ or use an explicit `is.na` decision rule
- Trees are rather robust to outliers: if a data point has an extreme
  value, it will be 'split away' rather early and the 'main branch' of
  the tree will be fit on the rest of the data. Non-outlier test points
  will funnel into this 'main branch' and it will be like the outlier
  essentially never existed
  
There are tons of tiny little decisions in building trees, so I don't
recommend writing your own implementation. See, *e.g.*, `?rpart.control`
for the fitting parameters used by the `rpart` package: other software
will have similar options. 

## Random Forests

We are now ready to introduce the concept of _random forests_, which are
_almost_ bagged decision trees. Recall our big goal for bagging: constructing
an ensemble of *minimally correlated* *very flexible* base learners. Clearly, 
'deep' trees, *i.e.*, trees with many splits, can be very flexible. (In the
limit, sufficiently deep trees can fit the training data perfectly *à la* 
$1$-nearest neighbor methods.) And row-subsampling (bagging) gives us a good
source of variance. So "bagged decision trees" are pretty darn good. But can we
go further? 

Indeed! We have one other knob we can turn: random use of features. When building
a single tree, we always built our next split by looking for the *best* feature
to split on out of *all* of our features. But we don't actuallly need to use 
the best feature at each step. Instead, we can randomly select a subset of
features at each split, pick the best of those, and go on our merry way. 

Clearly, this is going to give us more variability, but are we concerned it 
might hurt us? If we (randomly) leave out the most important features, are we 
just producing a worse tree overall? This is where the true genius of random
forests comes in: the feature is not *removed from the whole tree*, it is only
removed from that one split. We can use it (if randomly selected) in later 
(deeper) splits. That is, if the feature is important early and we happen to 
miss it, the tree will wind up partially reconstructing it later. 

So that's the key construction of *random forests*: 

- Build a bunch of decision trees; 
- by randomly subsampling rows (observations) for each tree; and
- randomly sampling features (columns) *within* each split. 

This basically hits everything we want for ensemble learning: high variance
flexible learners in the form of deep decision trees that are constructed in 
a way that gives us lots of tree-to-tree variance (resulting from both our row
and column sampling) without loss of power in any individual tree (because
the missing column in split $i$ can reappear in all later splits). 

Additionally, because trees are relatively quick to code and quick to apply 
(computers are very good at less than/greater than checks and binary logic), 
random forests are *fast*. Additionally, we note that for any bagging scheme-
including random forests-the base learners are built and applied in parallel, so
we can get even more speed by just throwing more compute power at the problem. 

Here, you might ask whether we can combine *stacking* and *random forests*. 
Instead of using a simple sample average of trees, can we get an optimally 
weighted average of our trees? These sorts of *optimized* forests definitely
have been explored, but they don't tend to add too much above random forests. 

As a general rule, random forests are pretty near optimal for applied ML on
tabular data. They are: 

- easy to apply
- fit quickly
- don't have too many *important* tuning parameters[^tuning] 
- have a built in error metric that can be used to make sure training doesn't go
  off the rails (OOB error)
- have great software support in almost every programming environment
- can fit linear and non-linear relationships quite well

For this reason, random forests are an indispensible part of the ML toolbox. 
You can throw them at almost any (tabular) problem and they will perform within
some reasonable percentage of optimality. Being clear here: random forests
are not *optimal* for every problem -- no method is -- but they do very well
on almost all problem with minimal analyst effort. It often takes real work
to beat a good random forest implementation. 

When I face a challenge, *e.g.* in work with scientific collaborators, I often
start by throwing a simple linear model (possibly with $k$ lasso selected features)
and a random forest and seeing how the performance compares. If the linear model
is 80-90% as good as the random forest, I usually just stop there and report back
"this is as good a linear model as you are likely to get without massive 
overfitting". If the linear model is 10% as good as the RF, I know there is more
work to be done and more signal that could be captured. Note that this strategy
makes sense in my typical context - developing an interpretable model for 
scientific discovery - but it may not always work for you. 

Due to this usefulness, almost every variation of random forests you can imagine
has already been explored. We won't explore these as most of them add little
general purpose value (many are useful in specific scenarios however). 

[^tuning]: Random Forests actually have a *lot* of tuning parameters: *e.g.*, 

    ```{r}
    #| message: false
    library(ranger)
    args(ranger)
    ```
    
    but most of these have pretty marginal impact. 

## Boosting

So "stack-'em-and-bag-'em" is a great pair of techniques for ensemble 
construction. Can we go further? 

In both strategies, we have a set of base learners built more-or-less independently
and are looking to combine them. This assumes that other learners can 'make up'
for incorrect answers learned by one base learner. (Either through averaging or
through a simple majority-vote scheme.) But this requires a bit of luck. 
Can we instead build an ensemble models that are designed to improve on each
other? 

This is the fundamental idea of *boosting*: given one learner, what if we fit
another learner whose goal is not to predict $y$ *per se*, but to fix where
the prior learner went wrong? While boosting can be applied for both classification
and regression (and was originally developed for classification), I find the 
math a bit more intuitive in the regression case. 

Suppose we have a set of training data $(\bX, \by)$ and we fit a predictor
$\hat{f}_1$. We can construct a set of 'first stage' predictions

$$\hat{\by}_1 = \hat{f}_1(\bX)$$ 

and associated residuals

$$\hat{\br}_1 = \by - \hat{\by}_1.$$  *If* we could predict $\hat{\br}_1$ 
reasonably well, say with another function $\hat{f}_2$, we could get better
predictions by combining the two models together: 

$$\hat{\by}_{1:2} = \hat{\by}_1 + \hat{\br}_1 = \hat{f}_1(\bX) + \hat{f}_2(\bX)$$

::: {.callout-caution title="Why not Use $\by$?"}

You might ask why note fit $\hat{f}_2$ to take $\hat{\by}_1$ or $\hat{\br}_1$
as inputs. We can't use anything that depends on $\by$ as inputs to any
$\hat{f}(\cdot)$  in the test (unknown $\by$) context. 

:::

At this point, your BS alarm should be going off. If it were possible to predict
$\hat{r}_1$ meaningfully in the first place, why didn't our original $\hat{f}_1$
do this? After all, we know from the gradient condition from OLS that: 

$$\bX^{\top}(\by - \bX \hat{\bbeta}) = \bzero \implies \bx_j^{\top}\br = 0 \text{ for all } j = 1, \dots, p$$

so OLS leaves no linear relationship "on the table." Many other models have
similar guarantees, but not all! 

As we have discussed throughout this course, we often pick models that don't 
"go to the mattresses" and take on their most flexible forms (think $K$-NN with
$K > 1$: it's not the 'all out' version of $K$-NN). If we pick a base learner
with minimal fitting ability, it might leave something on the table. 

So what is our laziest base learner? Let's go back to our beloved trees and
consider a tree with only a single split: 

```{r}
#| echo: false
library(rpart)
library(rpart.plot)
library(MASS)
rpart.plot(rpart(medv ~ . , data=Boston, maxdepth=1))
```

This type of minimal tree is often referred to as a *stump* in the ML
community.[^stump] Clearly, a stump is a very high-bias (inflexible) 
low variance learner. Even if we do not go all the way to a stump, shallow
decision trees still form a good basis for *boosting* or iterative model
improvement. 

This gives us our basic Boosting-type algorithm: 

- Initialize: $\hat{f}_0(\cdot) = 0$ (constant predictor)
- For $b = 1, \dots, B$: 
  - Compute residuals: $\hat{\br}_b = \by - \hat{f}_{b-1}(\bX)$
  - Build new learner $\tilde{f}_b$ on training data $(\bX, \hat{\br}_b)$
  - Update ensemble predictor: 
    $$ \hat{f}_b(\cdot) = \hat{f}_{b-1}(\cdot) + \lambda \tilde{f}_b(\cdot) \Leftrightarrow \hat{f}_b(\cdot) = \sum_{i=1}^b \lambda\tilde{f}_b(\cdot)$$
    
Wait? What is this new $\lambda$ term? It's a form of regularization that
we use to avoid overfitting. Instead of using *all* of the residual 
predictor we construct, we use a small fraction thereof. $\lambda$ is
typically chosen to be small, on the order of $\lambda \approx 0.01$. The
hope is that, by taking such small steps, and not running boosting *ad infinitum*
we can identify a good boosted predictor. 

::: {.callout-caution title="Boosting and Overfitting"}

In *bagging* scenarios, we can run $B \to \infty$ without risk of overfitting
because we are resampling data but still only fitting a model of limited 
capacity. The resampling only exists as a variance taming mechanism. 

In *boosting* scenarios, running $B \to \infty$ is a recipe for overfitting as
we will continue to fit increasingly more complex models to capture more and more
residual variance. It is important to have good data hygiene and use some sort
of hold-out/cross-validation scheme to avoid overfitting. 

:::

At this point, you might recognize a bit of a gradient descent 'flavor' to 
boosting: we take small steps in a direction that further reduces our loss. 
This is not just a coincidence: as noted above, regression residuals are intimately
tied to the gradient at a particular $\hat{\beta}$. If we want to extend boosting
to classification contexts, where residuals are not well-defined, one commonly
used strategy (not surprisingly called *gradient boosting*) fits each successive
base learner to the *gradient* of the loss function at each training point: that
is, we fit the new base learner so that it pushes us "just a bit" in a direction
that further minimizes the loss. 

While other forms of boosting exist (`AdaBoost`, `LogitBoost`), gradient boosting
is the most commonly used today. In particular, a software framework known as
[`xgboost`](https://xgboost.ai) dominates the boosting landscape. 
`xgboost` - with the `x` standing for e**X**treme - is not just the 
blue-raspberry flavor of machine learning. It uses small decision trees as the
base learner, additionally using a random-forest feature randomization step to
get some additional variability. The authors of `xgboost` have put significant
effort into `xgboost` to make it as efficient as possible on a variety of
computer architectures and have added many additionally useful features like
automatic feature selection, a penalization scheme to downweight certain trees,
and a Newton-like step instead of a simple gradient step for constructing the
ensemble. See the [official documentation](https://xgboost.readthedocs.io/en/stable/tutorials/model.html)
for a more detailed discussion of these and the other powerful features of the 
`xgboost` library. 

Like random forests, `xgboost` is also a widely used "nearly optimal" tool of
choice for much applied predictive modeling and you are highly encouraged to
familiarize yourself with it and add it to your toolbox. 

Below, I demonstrate both random forests and xgboost to predict the `price` feature
in the `diamonds` data you might have seen used to demonstrate `ggplto2`: 

```{r}
library(rsample)
library(tidyverse)
library(ranger) # Provides random forest
library(xgboost) # Provides xgboost


set.seed(9890)
split <- initial_split(diamonds, prop = 0.8)
train_diamonds <- training(split)
test_diamonds  <- testing(split)

diamonds_lm <- lm(price ~ ., data=train_diamonds)
diamonds_rf <- ranger(price ~ ., data=train_diamonds)
# No formula interface
diamonds_xg <- xgboost(x = model.matrix(diamonds_lm), y=train_diamonds$price)

y_hat_lm <- predict(diamonds_lm, newdata=test_diamonds)
y_hat_rf <- predict(diamonds_rf, data=test_diamonds)$predictions

## Some awful magic to process test_diamonds into the expected format
## Ignore this!
diamonds_lm_terms <- diamonds_lm |> terms() |> delete.response()
m <- model.frame(diamonds_lm_terms, test_diamonds, xlev = diamonds_lm$xlevels)
diamonds_lm_X_test <- model.matrix(diamonds_lm_terms, m, contrasts.arg = diamonds_lm$contrasts)

y_hat_xg <- predict(diamonds_xg, newdata=diamonds_lm_X_test)

list(y=test_diamonds$price, 
     y_hat_lm=y_hat_lm, 
     y_hat_rf=y_hat_rf, 
     y_hat_xg=y_hat_xg) |> 
  bind_cols() |> 
  pivot_longer(cols=starts_with("y_hat"), 
               names_to="method", 
               values_to="prediction") |>
  mutate(method = recode_values(method, 
    "y_hat_lm" ~ "Linear Regression", 
    "y_hat_rf" ~ "Random Forest", 
    "y_hat_xg" ~ "xgBoost")) |> 
  ggplot(aes(x=y, y=prediction)) + 
  geom_point(alpha=0.3) + 
  facet_wrap(~method, ) + 
  theme_bw() + 
  xlab("True y") + 
  ylab("Predicted y") + 
  geom_abline(slope=1, intercept=0, color="red4", linewidth=2) +
  scale_x_continuous(labels=scales::dollar) + 
  scale_y_continuous(labels=scales::dollar)
```

This is quite clunky - and we didn't even do any tuning. Still, we see that
a naive application of random forest 
(RMSE \$`r yardstick::rmse_vec(test_diamonds$price, y_hat_rf) |> round(digits=2)`)
or xgboost
(RMSE \$`r yardstick::rmse_vec(test_diamonds$price, y_hat_xg) |> round(digits=2)`)
gives a real improvement over a linear model
(RMSE \$`r yardstick::rmse_vec(test_diamonds$price, y_hat_lm) |> round(digits=2)`). 
For a more fluid UX, I recommend the use of the `tidymodels` package instead. 

TODO: Tidymodels Example

[^stump]: To my eye, this is more like the very sad Charlie Brown Christmas
tree with one branch than a true *stump*, but if the worlds of statistics
and ML can barely agree on terminology, what chance is there for botany
and ML to agree on terms? 
