---
session: "`r session <- 11; session`"
topic: "`r library(tidyverse); topic <- read_csv('../key_dates.csv', name_repair='universal') |> filter(Course.Element == 'Class Session', Item.Number==session) |> pull(Details); topic`"
date: "Last Updated `r Sys.Date()`"
---

{{< include _latex_defines.qmd >}}

```{r}
#| echo: false
set.seed(9890)
```

For our first topic in unsupervised learning, we turn to 
_clustering_, the task of dividing _observations_ into 
meaningful groups. Before we get to methods, what are
some reasons why we might want to perform clustering on
a given data set? 

- Simplification of structure: by dividing a complex
  population into several smaller and more homogeneous
  subpopulations, we hope to make it easier to understand
  some underlying phenomenon. 
  
  This is common in things like medical studies: rather than
  dealing with _cancer_, as a large and undifferentiated
  phenomenon, we have divided cancers into types and subtypes
  of related disease. Note that these are certainly not all identical
  within a type - the random genetic abnormalities caused by excessive
  UV exposure in my skin aren't necessarily the same random genetic
  abnormalities caused by excessive UV exposure in your skin - but they
  are still much more homogeneous than two totally different cancers, *e.g.*,
  pancreatic and brain. By recognizing the subgroups, we have a simpler
  task to understand four or five groups separately than building one 
  "super-model" that covers them all. 
  
  Statistically, you can think of this as being a special case of one of
  our favorite properties of variance: 
  
  $$\V[Y | X ] \leq \V[Y] \text{ for any random variables } X, Y$$ 
  
  but we do not necessarily know the relevant $X$ *a priori*. (Compare
  this to regression, where we typically have a model - often implicit -
  about $X$ and the relationship between $X$ and $Y$).
  
- Data compression: when dealing with _large_ data sets, it may be too
  expensive to treat all points separately. If we can identify repeated
  (or nearly repeated) points, we combine them and add a weight of "2" 
  to our algorithm, earning us some savings in computational time and in
  memory usage. 
  
- Denoising: when faced with noisy observations, we might get a better of
  the 'true' underlying signal by averaging several similar observations. 
  Obviously, we don't want to accidentally mix in signals that are
  fundamentally different, but slight differences might be ok in a
  "bias-variance" sort of way. 
  
In each of these tasks, we're faced with a bit of a trade-off in our
task: 

- If we divide our data into _more_ clusters, we have more homogeneity 
  within the groups, but we get less of the benefits of clustering because
  we still have more groups to deal with. 
- If we divide our data into _fewer_ clusters, we get more simplicity (fewer
  clusters) but our simplified (clustered) representation is lossier. 
  
In our discussion of clustering, this is the principal "tuning parameter" we 
will have to wrestle with. (Sometimes this will be an explicit "how many
clusters?" hyperparameter, while other times we will have a "how much
combining?" parameter which will only implicitly determine the number of 
clusters when we apply it to a data set.) As discussed last week, there's not
a simple "just holdout"-type heuristic we can use to assess the performance of
clustering and our validation strategies typically need to be a bit more 
problem specific. 

::: {.callout-caution title="This isn't Economists' Clustering"}

If your regression training had some an economics-flavored background, 
you may have heard the term _clustering_ used in a regression context,
referring to groups of related observations that should be not modeled as
independent observations (*e.g.*, standardized test performance of different
classes within the same school). This is not that type of clustering! 

In that supervised context, we know the underlying group structure and use it
to inform our regression modeling. In the unsupervised context we consider,
we do not know the underlying group structure (though we assume one exists!)
and we instead seek to discover it.

:::

## Distance-Based Clustering

We will build up our basic clustering framework from a _distance-based_ 
perspective. If two points $\bx_1, \bx_2$ are relatively close (in a distance
sense), they are more suitable candidates for clustering than two points
$\bx_3, \bx_4$ which are very far apart. This will not always be the case - 
if $\bx_4$ is very far away from anything else, $\bx_3$ may still be its 
"nearest-neighbor" - but it gives us a good place to start. 

We will begin with clustering based on Euclidean ($\ell_2$) distances, but 
most of the methods we consider can be "kernelized" - allowing for other
distances in a feature-expanded space - or "sparsified" - allowing the method
to identify and cluster based only on distances in a few relevant features. 
Because we start with $\ell_2$ distance, these methods are going to perform
their best when applied to data with nice isotropic (circular) structure. 

This type of data will be relatively easy to cluster: 

```{r}
#| echo: false
#| message: false
#| fig-asp: 1
library(tidyverse)
D <- data.frame(c = rep(c("A", "B"), each=50)) |>
    mutate(x_mean = if_else(c == "A", -2, 2), 
           y_mean = if_else(c == "A", -0, 0), 
           x = rnorm(n(), mean=x_mean), 
           y = rnorm(n(), mean=y_mean))
ggplot(D, aes(x=x, y=y, color=c)) + 
    geom_point(alpha=0.5) + 
    theme_bw() + 
    scale_color_brewer(palette=2, 
                       type="qual", 
                       name="True Cluster") + 
    theme(legend.position="bottom") + 
    xlab(NULL) + 
    ylab(NULL) + 
    geom_point(aes(x=x_mean, y=y_mean), 
               size=10, 
               pch=18)
```

Of course, we do not generally get our data so nicely labeled and instead
are faced with something like: 

```{r}
#| echo: false
#| message: false
#| fig-asp: 1
ggplot(D, aes(x=x, y=y)) + 
    geom_point() + 
    theme_bw() + 
    xlab(NULL) + 
    ylab(NULL)
```

While the cluster structure is less clear, a first pass would lead to
placing a circle around the 

Clearly, if you were to cluster this data, drawing a circle around the group
on the left and the group on the right: 

```{r}
#| echo: false
#| message: false
#| fig-asp: 1
library(ggforce)
ggplot(D, aes(x=x, y=y)) + 
    geom_point() + 
    theme_bw() + 
    xlab(NULL) + 
    ylab(NULL) + 
    geom_circle(aes(x0=x_mean, y0=y_mean, r=2))
```

Clearly not a perfect clustering - we have a few uncircled points to deal with
and the choice of the radius is debatable - but a darn good first start. Notably,
the points in each circle are generally closer to the center of the circle (and
hence to each other) than they are to points in the other circle.

Compare this to a situation in which there is strong correlation among the 
two coordinates, even if the two clusters are still centered in the same place. 

```{r}
#| echo: false
#| message: false
#| fig-asp: 1
library(tidyverse)
rho <- 0.8
sqrt_1_rho2 <- sqrt(1-rho^2)
D <- data.frame(c = rep(c("A", "B"), each=50)) |>
    mutate(x_mean = if_else(c == "A", -2, 2), 
           y_mean = if_else(c == "A", -0, 0), 
           x = rnorm(n(), mean=x_mean), 
           y = rnorm(n(), mean=y_mean - rho * (x - x_mean), sd=sqrt_1_rho2))
ggplot(D, aes(x=x, y=y, color=c)) + 
    geom_point(alpha=0.5) + 
    theme_bw() + 
    scale_color_brewer(palette=2, 
                       type="qual", 
                       name="True Cluster") + 
    theme(legend.position="bottom") + 
    xlab(NULL) + 
    ylab(NULL) + 
    geom_point(aes(x=x_mean, y=y_mean), 
               size=10, 
               pch=18)
```

The same "draw a circle" strategy gives us 

```{r}
#| echo: false
#| message: false
#| fig-asp: 1
ggplot(D, aes(x=x, y=y)) + 
    geom_point() + 
    theme_bw() + 
    xlab(NULL) + 
    ylab(NULL) + 
    geom_circle(aes(x0=x_mean, y0=y_mean, r=2))
```

This still isn't a terrible outcome - it's just not that hard a problem! - but 
it's clearly _harder_ than the previous problem, even though the clusters 
themselves aren't any further apart and the variance in each direction isn't 
changed. If we were able to represent this data in a more 'diagonal' way that
captured the structures of each cluster, we might do better. (*Hint, hint*: PCA)

With that aside complete, our basic clustering problem is thus: 

- Divide all points into one of $K$ groups
- Minimize the variance within the groups
- Maximize the distance _between_ the groups

Here, we can parameterize each group by its _center_ point (mean), which is 
known as the _centroid_.[^centroid] We can modify the above to: 

[^centroid]: You might ask whether it is fair to swap "distance from point A to
point B" for "distance from A to the midpoint of A, B". It turns out that, for
Euclidean distance, this is basically fine. You might remember an alternate
definition of variance we sometimes use for random variables: 
$\E[(X - X')^2] = 2\sigma^2$ where $X, X'$ are independent samples from a
distribution with variance $\sigma^2$. We're applying essentially the same 
argument here. 

- Divide all points into one of $K$ groups
- Minimize the distance from each point to its nearest centroid
- Maximize the distance _between_ the centroids

So let's write this as an optimization problem: 

$$\argmin_{\substack{c_1, c_2, \dots, c_n \in \{1, 2, \dots, K}\} \\ C_1, C_2, \dots, C_K \in \R^p} \sum_i \|X_i - C_{c_i}\|_2^2$$

That's a mess! But what does it mean?

i) We want to find labels $c_1, \dots, c_n$. We have one label for each of our
   $n$ points and, rather than being arbitrary numbers, we only get to pick
   the integers $1$ to $K$.
ii) Additionally, we want to estimate $K$ centroids $C_1, \dots, C_K$. 
iii) Our objective function is the sum of squared distances from each point
     $X_i$ to its corresponding centroid $C_{c_i}$. For instance, if our third
     point was mapped to centroid $2$, the corresponding term in the objective
     $\|X_3 - C_2\|_2^2$
     
So not too complex once we take it apart, but we will have trouble solving it. 
Due to the discrete labels $\{c_i\}$, this is a _combinatorial_ (discrete) 
problem and we don't get the convexity magic. 

Our first two clustering methods are based on on _approximately_ solving this
problem. 

### The $K$-Means Model

Returning to our problem above, while it is hard to solve the problem 
*as written*, we can break it into two relatively simple problems: 

- If we know the centroids, it's easy to determine the _closest_ centroid for
  each point
- If we know the cluster labels, we can estimate the corresponding centroids by
  averaging all the points in the cluster. 
  
Putting these together, we get a pretty great clustering algorithm. We start with
a guess of either the labels and then we iteratively update the centroids and 
labels, using our answer from the previous label update to get better centroids
and then using our latest centroids to get better labels. We run this process
until the labels stop changing (which implies the centroids stop changing) and
we have a pretty good guess as to our clustering. 

Let's see this in action. I'm going to generate data with three clusters. These
will be visually apparent so you can see the algorithm in action. 

```{r}
D <- data.frame(c = rep(c("A", "B", "C"), each=40)) |>
    mutate(r = 1, 
           theta = recode_values(c, "A" ~ 0, "B" ~ 2*pi/3, "C" ~ 4*pi/3), 
           z_mean = r * exp((0 + 1i) * theta), 
           x_mean = Re(z_mean), 
           y_mean = Im(z_mean),
           x = rnorm(n(), mean=x_mean, sd=0.75), 
           y = rnorm(n(), mean=y_mean, sd=0.75), 
           id = row_number())

ggplot(D, aes(x=x, y=y)) + 
  geom_point() + 
  theme_bw()
```

We see three pretty clear clusters, with a few possibly ambiguous points.
To start our algorithm, let's randomly assign each point to one of three
clusters, indicated by colors: 

```{r}
D <- D |> mutate(c_hat = sample(c("A", "B", "C"), n(), replace=TRUE))
ggplot(D, aes(x=x, y=y, color=c_hat)) + 
  geom_point() + 
  theme_bw() + 
  scale_color_brewer(type="qual", 
                     palette=2, 
                     name="Initial Guessed Clusters")
```

Obviously, by doing this randomly, we're not going to do well. But with
these clusters, we can now estimate cluster centroids by taking the 
average of all points of that color: 

```{r}
D <- D |> 
  group_by(c_hat) |>
  mutate(x_centroid = mean(x), 
         y_centroid = mean(y)) |>
  ungroup()

ggplot(D) + 
  geom_point(aes(x=x, y=y, color=c_hat), alpha=0.5) + 
  geom_point(aes(x=x_centroid, y=y_centroid, color=c_hat), pch=12, size=3) + 
  theme_bw() + 
  scale_color_brewer(type="qual", 
                     palette=2, 
                     name="Initial Guessed Clusters")
```

With these, we get to update our cluster estimates: 

```{r}
D1 <- D |> select(-c_hat, -x_centroid, -y_centroid)
D2 <- D |> select(c_hat, x_centroid, y_centroid) |> distinct()

D <- cross_join(D1, D2) |> 
  mutate(dist_to_centroid = (x - x_centroid)^2 + (y - y_centroid)^2) |>
  group_by(id) |> 
  slice_min(dist_to_centroid) |> 
  ungroup() |> 
  select(-ends_with("centroid"))

ggplot(D) + 
  geom_point(aes(x=x, y=y, color=c_hat)) + 
  theme_bw() + 
  scale_color_brewer(type="qual", 
                     palette=2, 
                     name="Guessed Clusters - 1 Round")
```

Not perfect, but much improved! From here, we again get new estimated 
centroids and plot them: 

```{r}
D <- D |> 
  group_by(c_hat) |>
  mutate(x_centroid = mean(x), 
         y_centroid = mean(y)) |>
  ungroup()

ggplot(D) + 
  geom_point(aes(x=x, y=y, color=c_hat), alpha=0.5) + 
  geom_point(aes(x=x_centroid, y=y_centroid, color=c_hat), pch=12, size=3) + 
  theme_bw() + 
  scale_color_brewer(type="qual", 
                     palette=2, 
                     name="Guessed Clusters - 1 Round")
```

We can repeat this process a few times and see that the 
algorithm converges quickly: 

```{r}
D1 <- D |> select(-c_hat, -x_centroid, -y_centroid)
D2 <- D |> select(c_hat, x_centroid, y_centroid) |> distinct()

D <- cross_join(D1, D2) |> 
  mutate(dist_to_centroid = (x - x_centroid)^2 + (y - y_centroid)^2) |>
  group_by(id) |> 
  slice_min(dist_to_centroid) |> 
  ungroup() |> 
  select(-ends_with("centroid")) |>
  group_by(c_hat) |>
  mutate(x_centroid = mean(x), 
         y_centroid = mean(y)) |>
  ungroup()

ggplot(D) + 
  geom_point(aes(x=x, y=y, color=c_hat), alpha=0.5) + 
  geom_point(aes(x=x_centroid, y=y_centroid, color=c_hat), pch=12, size=3) + 
  theme_bw() + 
  scale_color_brewer(type="qual", 
                     palette=2, 
                     name="Guessed Clusters - 2 Rounds")
```

```{r}
D1 <- D |> select(-c_hat, -x_centroid, -y_centroid)
D2 <- D |> select(c_hat, x_centroid, y_centroid) |> distinct()

D <- cross_join(D1, D2) |> 
  mutate(dist_to_centroid = (x - x_centroid)^2 + (y - y_centroid)^2) |>
  group_by(id) |> 
  slice_min(dist_to_centroid) |> 
  ungroup() |> 
  select(-ends_with("centroid")) |>
  group_by(c_hat) |>
  mutate(x_centroid = mean(x), 
         y_centroid = mean(y)) |>
  ungroup()

ggplot(D) + 
  geom_point(aes(x=x, y=y, color=c_hat), alpha=0.5) + 
  geom_point(aes(x=x_centroid, y=y_centroid, color=c_hat), pch=12, size=3) + 
  theme_bw() + 
  scale_color_brewer(type="qual", 
                     palette=2, 
                     name="Guessed Clusters - 3 Rounds")
```

```{r}
D1 <- D |> select(-c_hat, -x_centroid, -y_centroid)
D2 <- D |> select(c_hat, x_centroid, y_centroid) |> distinct()

D <- cross_join(D1, D2) |> 
  mutate(dist_to_centroid = (x - x_centroid)^2 + (y - y_centroid)^2) |>
  group_by(id) |> 
  slice_min(dist_to_centroid) |> 
  ungroup() |> 
  select(-ends_with("centroid")) |>
  group_by(c_hat) |>
  mutate(x_centroid = mean(x), 
         y_centroid = mean(y)) |>
  ungroup()

ggplot(D) + 
  geom_point(aes(x=x, y=y, color=c_hat), alpha=0.5) + 
  geom_point(aes(x=x_centroid, y=y_centroid, color=c_hat), pch=12, size=3) + 
  theme_bw() + 
  scale_color_brewer(type="qual",
                     palette=2, 
                     name="Guessed Clusters - 4 Rounds")
```

At this point, the algorithm is terminated (since repeating the updates won't
change anything) and we actually have a pretty good result. You could quibble 
with some points on the boundary, but this is actually pretty impressive if you 
think back to our starting point just a few iterations earlier.

The algorithm is known as *Lloyd's Algorithm* and it solves the so-called
"$K$-Means" problem (of estimating $K$ cluster means). Despite its simplicity,
it usually converges quite quickly to a reasonable solution. In `R`, we implement
this via the `kmeans()` function: 

```{r}

```

Which we visualize as before: 

```{r}
D1 <- D |> select(-c_hat, -x_centroid, -y_centroid)
D |> 
  select(x, y) |> 
  as.matrix() |> 
  kmeans(centers=3) |> 
  useful::fortify.kmeans() |> 
  cross_join(D1) |> 
  mutate(dist_to_centroid = (x - .x)^2 + (y - .y)^2) |>
  group_by(id) |> 
  slice_min(dist_to_centroid) |> 
  ungroup() |> 
  select(-ends_with("centroid")) |> 
  ggplot(aes(color=.Cluster)) + 
  geom_point(aes(x=x, y=y), alpha=0.5) + 
  geom_point(aes(x=.x, y=.y), size=3, pch=10) + 
  theme_bw() + 
  scale_color_brewer(type="qual",
                     palette=2, 
                     name="Guessed Clusters - kmeans()")
```

Which is quite nice! 

But this actually doesn't quite match our results from above.
What happened? 

Lloyd's Algorithm is sensitive to initialization - with
different starting values, we get different final results.
(Recall that this can't happen with convex problems!) To 
get around this, we typically run $K$-means several times with different
initial values and pick the best result (defined as the
minimum within cluster variance). 

```{r}
D1 <- D |> select(-c_hat, -x_centroid, -y_centroid)
D |> 
  select(x, y) |> 
  as.matrix() |> 
  kmeans(centers=3, nstart = 100) |> 
  useful::fortify.kmeans() |> 
  cross_join(D1) |> 
  mutate(dist_to_centroid = (x - .x)^2 + (y - .y)^2) |>
  group_by(id) |> 
  slice_min(dist_to_centroid) |> 
  ungroup() |> 
  select(-ends_with("centroid")) |> 
  ggplot(aes(color=.Cluster)) + 
  geom_point(aes(x=x, y=y), alpha=0.5) + 
  geom_point(aes(x=.x, y=.y), size=3, pch=10) + 
  theme_bw() + 
  scale_color_brewer(type="qual",
                     palette=2, 
                     name="Guessed Clusters - kmeans(nstart=100)")
```

Here, the data is small enough and the algorithm is fast enough we could run it
many times and get an improved result. 

Note that there are two ways in which the results can differ: 

- A "label switching" problem in which cluster "A" is called cluster "B", but
  the actual groupings don't change
- Different groupings, where two points are in the same group in one solution but
  different groups in a second solution
  
The first makes for some annoying book-keeping, but is ultimately harmless. The
latter is what we worry about when using $K$-means. 



i) If we initialize our centers 

### An EM Algorithm

### Convex Clustering

## Spectral Clustering

## Hierarchical Clustering

## Density Based Clustering

## Validation of Cluster Results

```{r}
#| include: false
1+1
```
