---
session: "`r session <- 1; session`"
topic: "`r library(tidyverse); topic <- read_csv('../key_dates.csv', name_repair='universal') |> filter(Course.Element == 'Class Session', Item.Number==session) |> pull(Details); topic`"
date: "Last Updated `r Sys.Date()`"
---

```{r}
#| echo: false
set.seed(100)
```

## A Taxonomy of Machine Learning

Where classical statistics focuses on learning information about a
population from a (representative) sample,  Machine Learning 
focuses on out-of-sample prediction accuracy. 

For example, given a large set of medical records, the natural
instinct of a statistician is to find the indicators of cancer in
order assess them via some sort of follow-on genetic study, while a
ML practitioner will typically start by building a predictive
algorithm to predict who will be diagnosed with cancer. The
statistician will perform calculations with $p$-values, test
statistics, and the like to make sure that any discovered
relationship is accurate, while the ML practitioner will verify
the performance by finding a new (hopefully similar) set of medical
records to test algorithm performance. 

Clearly, this difference is more one of style than substance: the
statistician might see what features are important in the ML model to
decide what to investigate, while the ML modeler will use statistical
tools to make sure the model is finding something real and not just
fitting to noise. In this course, the distinction may be even
blurrier as our focus is *statistical machine learning* - that little
niche right on the boundary between the two fields. 

In brief, "statistics vs ML" is a bit of a meaningless distinction as
both fields draw heavily from each other. I tend to say one is *doing
statistics* whenever the end-goal is to *better understand* something
about the real world (*ie.*, the end product is *knowledge*), while
one is *doing ML* whenever one is building a system to be used in an
automated fashion (*ie.*, the end product is *software*), but
definitions vary.[^culture]

[^culture]: There are some cultural differences that come from ML's
CS history vs Statistics' "Science-Support" history: notably, ML
folks think of (binary) classification while statisticians think of
regression as the first task to teach.

### Types of Learning

In this course we will use the following taxonomy borrowed from the
ML literature: 

- Supervised Learning: Tasks with a well-defined target variable (output)
that we aim to predict

  Examples: 
  - Given an image of a car running a red-light, read its license plate.
  - Given attendance records, predict which students will not pass a course.
  - Given a cancer patient's medical records, predict whether a certain drug
    will have a beneficial impact on their health outcomes. 
  
- Unsupervised Learning: Given a whole set of variables, none of which is
  considered an output, learn useful underling structure. 

  Examples: 
  - Given a set of climate simulations, identify the general trends of a 
    climate intervention.
  - Given a set of students' class schedules, group them into sets. (You might
    imagine these sets correspond to majors, but without that sort of "label" 
    (output variable) this is only speculative, so we're unsupervised)
  - Given a social network, identify the most influential users.

There are other types of learning tasks: *e.g.* semi-supervised, online,
reinforcement, but the Supervised/Unsupervised distinction is the main one
we will use in this course. 

Within Supervised Learning, we can further subdivide into two major categories:

- Regression Problems: problems where the response (label) is a
  real-valued number
- Classification Problems: problems where the response (label) is a
  category label.
  
  - Binary classification: there are only two categories
  - Multi-way or Multinomial classification: multiple categories
  
Linear Regression, which we will study more below, is the canonical
example of a *regression* tool for *supervised* learning.

At this point, you can already foresee one of the (many) terminology
inconsistencies will will encounter in this course: _logistic regression_
is a tool for _classification_, not regression. As modern ML is the intersection
of many distinct intellectual traditions, the terminology is rarely 
consistent.[^uk]

[^uk]: You may recall the famous quip about the US and the UK: "Two countries,
separated by a common language."

### Types of Data

It is also worth thinking of the many types of data we might analyze: much of
remarkable success of modern ML comes from an openness (and willingness) to 
analyze types of data historically 'out of bounds' (or, at least, not in the 
mainstream) for classical statistics.

Let's divide this into two axes: i) _organizational_, what is the natural 
way to store and represent this data on a computer; and ii) _modal_, what 
"type" of thing is each individual datum. While these are useful questions to
ask, these are not _fixed_ within a single problem. For instance, if given
a large text sample (such as doctor's notes on a medical patient), it is possible
to change this data from "sequence" data (things in a row) to "tabular" data,
by simply counting the occurence of each word. This transformation is
sometimes called a "bag-of-words" representation and it is lossy, but it can 
often still be good enough for many applications: *e.g.*, if the notes have
the word "cancer" appearing many times, it is a good guess that the patient
should receive advanced cancer screening; while it is _theoretically_ possible
that the doctor wrote "no sign of cancer" after each visit, this is unlikely.

This example highlights two ways of arranging data: 

- Tabular Data: this is your classic "spreadsheet"-style data, with rows
  representing separate observations and columns corresponding to well-defined
  features. (In [STA 9750](https://michael-weylandt.com/STA9750/), you may have
  heard this referred to as a "data frame".)
- Sequence Data: this is data presented as an organized sequence of 'tokens',
  where the ordering has meaning. Examples of sequence data include natural
  language (words as tokens), DNA sequence (the four amino acids as tokens), and
  certain representations of computer signals (0/1 binary). Sequence data is
  typically characterized by inconsistent length (sentences can be short or long)
  as opposed to tabular data where we represent observations with fixed-length
  vectors. 
  
Other data structures may include image data (data represented on a 2D grid), 
3D imaging data (as often occuring in medical imaging), video data (image data
with an additional time dimension), functional data (where we observe a 
'function' on a continuous domain, *e.g.*, an audio recording), or even shape
data (*e.g.*, inferring the malignancy of a strange growth on your skin
from its _shape_). 

Within these organizational patterns, we can also categorize data by the 
_modality_ of each datum. We spoke a bit about continuous and binary
data above (in the context of regression and classification), but you may also
see ordinal data (rankings, not corresponding to specific values, like 
survey responses to  "do you strongly agree, agree, slightly agree, *etc.*-type
questions) and categorical data (where there is a finite set of possible
unranked answers). Within categorical data, it is often useful to further
distinguish low-cardinality and high-cardinality problems: low-cardinality refer
to situations where there are relatively few possible categories
("summer/fall/winter/spring"); high-cardinality problems are those where the 
set of values is so impractically large that you can't handle each value 
separately. It is typical to model text data, such as LLMs, as a 
high-cardinality problem because, while there is technically a finite set of 
possible output words from an LLM, the problem of guessing the next LLM output
doesn't bear too much practical resemblence with a standard yes/no classification.

In this course, we focus on tabular real-valued data, but we will discuss
extensions to other data where appropriate. 

## Test and Training Error

As we think about measuring a model's predictive performance, it becomes
increasingly important to distinguish between _in-sample_ and _out-of-sample_
performance, also called training (in-sample) and testing (out-of-sample)
performance. 

In previous courses, you likely have assessed model fit by seeing how well
your model fits the data it was trained on: statistics like $R^2$, SSE, SSR in
regression or $F$-tests for model comparison do just this. As you have used them,
they are primarily useful for comparison of _similar_ models, *e.g.*, OLS with
different numbers of predictor variables. But it's worth reflecting on this
process a bit more: don't the model with more predictors always fit the data
better? If so, why don't we always just include all of our predictors? 

Of course, as you know, the answer is that we want to avoid "overfitting." Just
because a model fit the data a bit better doesn't mean it is actually better. 
If you need 1,000 variables to get a 0.1% reduction in MSE, do you really
believe those features are doing much? No! 

You likely have a sense that a feature needs to "earn its keep" to be worth
including in a model. Statisticians have formalized this idea very well
in some contexts: quantities like _degrees of freedom_ or _adjusted_ $R^2$
attemps to measure whether a variable provides a _statistically significant_
improvement in performance. These calculations typically rely on subtle
calculations involving nice properties of the multivariate normal distribution
and ordinary least squares, or things that can be (asymptotically) considered
essentially equivalent. 

In this class, we don't want to make those sorts of strong distributional and
modeling assumptions. So what can we do instead? Well, if we want to see if 
Model A predicts more accurately than Model B on new data, why don't we just
do that? Let's get some new data and compare the MSEs of Model A and Model B:
whichever one does better is the one that does better.[^sig]

This is a pretty obvious idea, so it's worth asking why it's not the baseline
and why statisticians bothered with all the degrees of freedom business to start
with. As always, you have to know your history: statistics comes from a lineage
of scientific experimentation where data is limited and often quite expensive to
get. If you are running a medical trial, you can't just toss a few hundred extra
participants in - this costs money! If you are doing an agricultural experiment,
it may take several years to see whether a new seed type actually has higher
yield than the previous version. It's also not clear how one should separate
data into training and test sets: if you are studying, *e.g.*, friendship
dynamics on Facebook, you don't have an (obvious) "second Facebook" that you can
use to assess model accuracy. 

By contrast, CS-tradition Machine Learning comes from a world of "internet-scale"
where data is plentiful, cheap, and is continuously being collected.[^anec] Not
all problems fall in this regime but, as we will see, enough do that it's worth 
thinking about what we should do in this scenario. Excitingly, if we don't
demand a full and exhaustive mathematical characterization of a method before we
actually apply it, we can begin to explore much more complex and interesting
models.

::: {.callout-tip title="Advice"}

A good rule of thumb for applied statistical and data science work: begin by
asking yourself what you would do if you had access to the whole population
(or an infinitely large sample) and then adapt that answer to the limited data
you actually have. You always want to make sure you are asking the right 
question, even if you are only able to give an approximate finite-data answer,
rather than giving an 'optimal' answer to a question you don't actually care
about.

:::

So, for the first two units of this course, we will put this idea front and
center: we will fit our models to a _training set_ and then see how well they
perform on a _test set_. Our goal is to not to find find models which perform
well on the test set *per se*: we really want to find models that perform well
on the all future data, not just one test set. But this training/test split will
certainly get us going in the right direction. 

Looking ahead, let's note some of the key questions we will come back to again
and again: 

- If we don't have an explicit test set, where can we get one? Can we 'fake it
  to make it'? 
- What types of models have small "test-training" gap, *i.e.* do about as well
  on the test and training sets, and what models have a large gap? 

[^anec]: Anecdotally, a Google Data Scientist once mentioned to me that they
rarely bother doing $p$-value or significance calculations. At the scale of
Google's A/B testing on hundreds of billions of ad impressions, _everything_
is statistically significant. 

[^sig]: You might still worry about the randomness of this comparison process:
if we had a slightly different sample of our new data, would the better model
still look better? You aren't wrong to worry about this, but we have at a minimum
made the problem much easier: now we are just comparing the means of two error
distributions - classic $t$-test stuff - as opposed to comparing two (potentially
complex) models.


::: {.callout-tip title="Pause for Reflection"}

Take the time to make sure you deeply understand what "overfitting" means. When
we talk about overfitting, we typically assume that there is a "correct" 
estimate and that the correct estimate is "best". But if we fit _better_ than
best, what is really going on?

:::

## Generalization and Model Complexity 

So far, we have two useful concepts:

- Training Set (the data we used to select our model)
- Test Set (the new data we used to assess our model)

The performance of a model on these two sets is, unsurprisingly, known as
_training error_ and _test error_, respectively. With these in hand, we 
introduce a third quantity that gives us a trivial, but surprisingly useful, 
inequality: 

$$\text{Test Error} = \text{Training Error} + \text{Generalization Gap}$$

Here, the "Generalization Gap" is defined as the difference between the training
error and the test error.[^testerror] Essentially, the Generalization Gap measures
the "optimism" bias obtained by measuring accuracy on the original training data.
If the Generalization Gap is large, the model will look much better on the 
training data then when we actually go to put it into practice. Conversely, if
the Generalization Gap is small, the performance we estimate from the training
data will continue when we deploy our model. 

This is a all a bit circular, but it lets us break our overarching goal 
(small test error) into two parts: 

- We want a model with small training error; and
- We want a model with small generalization gap. 

We can only be guaranteed that we'll have a small test error when these *both* 
of these are true. Again - and just to be clear - having a small training error
*is important*, but it is only *necessary* and not *sufficient* to have a small
test error. 

We can simply observe training error, so much of our theoretical analysis
focuses on understanding the generalization gap. For now, let's think about
the generalization gap of plain linear regression (OLS). It is not hard to show
(and we might show in class next week) that, if the OLS model is true, the
expected training MSE is: 

$$\mathbb{E}[\text{Training MSE}] = \mathbb{E}\left[\frac{1}{n}\sum_{i=1}^n(y_i - \sum_{j=1}^p x_{ij}\hat{\beta}_j)^2\right] = \frac{\sigma^2(n-p)}{n} = \sigma^2\left(1-\frac{p}{n}\right)$$

Here $\sigma^2$ is the 'noise variance' of the OLS model. (We will review the OLS
model in much more detail next week.) 

This is somewhat remarkable: if we knew the exact true model $\beta_*$, our MSE
would be $\sigma^2$, but our training MSE is less than that. How can we do better
than the optimal and exactly correct model? **Overfitting** - our OLS fits our
training data a bit 'too well' and it manages to capture the true signal *and* 
the noise. Whatever noise is in the data doesn't extend into (generalize to) 
the test set, so we get a bit of overfitting. 

Even a simple model like OLS is vulnerable to overfitting. It isn't too
big in this context and our formula above actually lets us see how it behaves: 

- As $n\to\infty$, the amount of overfitting goes down. This is of course the
  behavior we want: as we get more data, we should stop fitting noise and only
  fit signal.
- As $p \to n$, the amount of overfitting goes up. That is, as we add more 
  features (covariates) to our regression, we expect more over fitting. When
  we supply OLS with more 'degrees of freedom', some of those wind up fitting
  noise. 
  
The $n\to\infty$ behavior shouldn't surprise you: the theme of "more data yields
better estimation and smaller error" is ubiquitous in statistics. The behavior
as $p$ increases may not be something you have seen before. Later in this course,
we will actually ask what happens if $p > n$. Clearly, our formula from above 
can't hold as it predicts _negative_ MSE! But we'll get to that later...

As $p$ gets larger, OLS is more prone to overfitting. It turns out that this is
not a special property of OLS - basically all methods will have this property
to one degree or another. While there are many ways to justify this, perhaps
the simplest is a story about "complexity": with more features, and hence more
coefficients, OLS becomes a more complex model and more able to fit both the 
signal and the noise in the training data. Clearly, complexity is not necessarily
bad - we want to be able to capture all of the signal in our data - but it is
dangerous. 

This _complexity_ story is one we will follow through the rest of this course. 
A more _complex_ model is one which is able to fit its training data easily. 
Mathematically, we actually measure complexity by seeing how well a model fits
pure noise: if it doesn't fit it well at all (because there is no signal!), 
we can usually assume it won't overfit on signal + noise. But if it fits
pure noise perfectly, it has _by definition_ overfit the training data. 

I like to think of complexity as "superstitious" or "gullibility": the model
will believe (fit to) anything we tell it (training data), whether it is true
(signal) or not (noise). 

We don't want a model that is too complex, but we also don't want a model that
is too simple. If it can't fit signal, it is essentially useless. In our metaphor,
an overly simple (low complexity) model is like a person who simply doesn't believe
anything at all: they are never tricked, but they also can't really understand
the world. 


[^testerror]: I'm being a bit sloppy here. This is more precisely the 
*population test error*, not the *test set test error*. 

## Model Families and Estimation

Of course, the story we have told so far skims over a fundamental idea: 
model _fitting_. If we are given a set of parameters,-or, more generally, a 
prediction function--we might be able to discuss how well they fit (or overfit) 
the data, but that's not really the situation we often find ourselves in. 
In practice, we have some data and are asked to come up with the predictor
function ourselves. How can we do this? 

Well, this is a _machine_ learning course, so it's a reasonable starting 
assumption that we want our predictor to be implemented on a (finite-sized)
computer. Taking a reasonable upper bound for the size of the computer's memory, 
say 32 GB, this restricts us to an enormous, but technically finite, set of 
possible models.[^gb] Ignoring resource constraints, we _could_ in theory try
out all of these models and pick the one that fits our data "best". 

Obviously, this has some practical problems about runtime (and our power bill), 
but there's also the very real possibility that there's not a single "best" fit.
We can address both of these possibilities by restricting our attention to a
much smaller universe of prediction functions than "all possible programs". For
example, we might restrict our attention to univariate linear regression models,
parameterized by a single slope (one "double" value, or 64 bits) and a single
intercept (64 bits again). This gives us 128 bits total, so "only" $2^{128}$ 
possible lines to check.

This is obviously still deeply impractical, but this restriction allows us to
use much smarter algorithms than simply "try everything". The mathematical 
concept that lets us find the "best" predictor efficiently is called _convexity_
and we will study it later in this course. Furthermore, this set of "all possible
lines" is small enough that there will almost always be a single "best" line, so
our problem of finding the best predictor _within the class of simple linear 
regression models_ is well-posed. 

So restriction to a set of possible models does (at least) two useful things
for us: 

1) It gives us a fighting-chance of finding the best possible model 
   _in that category_
2) It makes it possible to have one (or at least a small-ish number) of 
   best possible models
   
This restriction does even more for us: it can be used to control the size of
the generalization gap we talked about above. When we start with our "big" class
of all possible programs, it's clear that there are trillions of programs that
will fit our training data perfectly but can go totally wild when exposed to new
data.[^rng] When we instead focus on linear regression, there is an "upper limit"
of terribleness. If our model was doing reasonably well in the past and 
our training data resembles our test data, we can feel reasonably confident 
things won't go off the rails at test time. 

This final idea is _statistical complexity_: if we pick a "reasonable" set of
models, we will be able to argue that the generalization gap is not too big. 
So building an ML system is really a two stage process: 

1) We need to select a class of models that is interesting enough to reasonably
   capture our training data, but not so complex that it could have a huge 
   generalization gap. 
2) We use smart algorithms to pick the _best_ model within the model class. 

While we won't rigorously characterize the complexity of model classes
in this course, it's still helpful to be familiar with the concept in a 
"know it when you see it" way.[^wain]

[^gb]: A mere $32 = 1024^3 * 8 = 2^{38} = 274,877,906,944$ bits, giving 
$2^{274877906944}=2^{2^{38}} = 2^{2^{2^{19}}}$ possible programs, but who's 
counting? 

[^wain]: The book *High-Dimensional Statistics* by Wainwright introduces these
ideas beautifully, but they are unavoidably a bit technical. 

## Overture

Before we turn to the first model class we'll consider in this class, let's take
a quite look ahead and what we'll be covering for the next few weeks: 

```{r}
#| message: false
#| echo: false
library(tidyverse)
get_topic <- function(n){
    read_csv("../key_dates.csv") |> 
        filter(`Course Element` == "Class Session") |> 
        filter(`Item Number` == n) |> 
        pull(Details) |> 
        str_extract(".*: [IV]+. (.*)", group=1)
}
```

- Week 02: `r get_topic(2)`. We will first review some of the basic linear
  algebra used throughout this course. Linear algebra is essential to machine
  learning in several respects: 
  
  i) Tabular data is organized as a matrix, the fundamental object of study in 
     linear algebra.
  ii) Linear methods (*e.g.*, OLS) are a useful baseline for many different 
      problems.
  iii) Most computational math is built from linear algebra tools. Even 
       non-linear problems are typically solved by repeatedly applying linear
       algebra techniques to linear approximations (think Taylor series).
       
- Week 03: `r get_topic(3)`. Next, we'll take a deep dive into what it means
  for a ML system to perform _well_. While familiar measures like MSE might
  make sense in some contexts, their applicability is much less clear when
  applied to things like LLMs. A key theme here will be _metrical
  determinism,_[^argmin_blog] the idea that once we settle on what our actual
  goal is and the way we will be measuring it, the best procedure is often
  automatically determined. We will also explore the popular MSE loss in more
  detail, showing how it can be broken into several easily-interpretable terms
  and relating these terms to model complexity. 
- Week 04: `r get_topic(4)`. Our last week of foundations will consider the role
  of computers in ML. We will discuss ways of finding that "best fit" predictor
  in a given class (mathematical optimization) and the uses and limits of 
  simulation studies as a way of understanding the usefulness of ML models. 
- Week 05: `r get_topic(5)`. As we begin our study of supervised learning, we
  will briefly introduce the idea of _statistical shrinkage_, the idea of 
  willingly accepting a small systematic error to eliminate the possibility of
  large random errors. We will explore this in the classic context of mean 
  estimation before applying shrinkage ideas to more interesting prediction 
  problems.
- Week 06: `r get_topic(6)`. We will apply shrinkage to our beloved OLS using
  two fundamental forms of shrinkage: $\ell_1$ and $\ell_2^2$-penalization. 
  While mathematically similar, these induce very different behaviors in OLS
  estimates. We will particularly focus on the role of _sparsity_ and model 
  selection in high-dimensional problems. 
- Week 07: `r get_topic(7)`. We will next turn to classification problems, with
  a focus on generative (probability-based) classifiers. This (hopefully) coincides
  with Baruch's "Ethics Week" so we will also discuss the perils of (societal)
  bias in classification contexts. 
- Week 08: `r get_topic(8)`. We will explore another tradition of classification
  problems based on geometry (not probability). These models play particularly
  nicely with non-linear decision boundaries, so we will also introduce methods
  for turning linear methods into non-linear predictors. (Note that these can
  be applied to both regression and classification problems.)
- Week 09: `r get_topic(9)`. Finally, we will consider _ensemble learning_, the
  practice of building predictors out of several simpler predictors. Ensembling
  is particularly powerful when applied to flexible but simple models, like
  decision trees, and this will let us introduce two of the most widely-used
  "fire and forget" ML models: random forests and `xgboost`. 
- Week 10: `r get_topic(10)` to Unsupervised Learning. After finishing our 
  discussion of supervised learning, we will turn to the _unsupervised_ context.
  Unsupervised learning poses extra challenges around validation and performance
  measurement, so we will begin with a discussion of the aims (and limits) of
  unsupervised learning. 
- Week 11: `r get_topic(11)`. We will cover the important unsupervised task of
  finding "typical" or "representative" simple structure in data distributions.
  In certain cases, we can "simplify" data into these patterns, providing methods
  of dimension reduction. Special emphasis will be placed on the most important
  statistical dimension reduction technique, PCA. 
- Week 12: `r get_topic(12)`. TODO
- Week 13: `r get_topic(13)`. TODO


[^argmin_blog]: I picked up this term from Ben Recht's blog, but it is explored
in more detail in [this paper](https://arxiv.org/abs/2509.04546).

## Nearest Neighbor Methods

Let us now take a quick detour into a flexible _family_ of models and see how
performance relates to the complexity story. Specifically, let's look at
how complexity plays out for a very simple classifier, 
$K$-Nearest Neighbors (KNN). KNN formalizes the intuition of 
"similar inputs -> similar outputs." KNN looks at the $K$ most similar points
in its training data  ("nearest neighbors" if you were to plot the data) and
takes the average label to make its prediction.[^details]

[^details]: There are details here about how we measure similarity, but for
now we will restrict our attention to simple Euclidean distance. 

```{r}
library(class) # Provides a KNN function for classification
args(knn)
```

You can see here that KNN requires access to the full training set at
prediction time: this is different than something like OLS where we reduce our
data to a set of regression coefficients (parameters).[^np]

[^np]: KNN is a *non-parameteric* method, but not all *non-parametric* methods
require access to the training data at test time. We'll cover some of those
later in the course.

We'll also need some data to play with. For now, we'll use synthetic data:

```{r}
#' Make two interleaving half-circles
#'
#' @param n_samples Number of points (will be divided equally among the circles)
#' @param shuffle Whether to randomize the sequence
#' @param noise Standard deviation of Gaussian noise applied to point positions
#'
#' @description Imitation of the Python \code{sklearn.datasets.make_moons} function.
#' @return a \code{list} containining \code{samples}, a matrix of points, and \code{labels}, which identifies the circle from which each point came.
#' @export
make_moons <- function(n_samples=100, shuffle=TRUE, noise=0.25) {
  n_samples_out = trunc(n_samples / 2)
  n_samples_in = n_samples - n_samples_out
  
  points <- matrix( c(
    cos(seq(from=0, to=pi, length.out=n_samples_out)),  # Outer circle x
    1 - cos(seq(from=0, to=pi, length.out=n_samples_in)), # Inner circle x
    sin(seq(from=0, to=pi, length.out=n_samples_out)), # Outer circle y
    1 - sin(seq(from=0, to=pi, length.out=n_samples_in)) - 0.5 # Inner circle y 
  ), ncol=2) 
  
  if (!is.na(noise)) points <- points + rnorm(length(points), sd=noise)
  
  labels <- c(rep(1, n_samples_out), rep(2, n_samples_in))
  
  if (!shuffle) {
    list(
      samples=points, 
      labels=labels
    )
  } else {
    order <- sample(x = n_samples, size = n_samples, replace = F)
    list(
      samples=points[order,],
      labels=as.factor(ifelse(labels[order] == 1, "A", "B"))
    )
  }
}
```

This function comes from the `clusteringdatasets` R package, but 
the underlying idea comes from a function in `sklearn`, a popular Python ML 
library.

Let's take a look at this sort of data:

```{r}
TRAINING_DATA <- make_moons()
```

```{r}
#| message: false
library(ggplot2)
library(tidyverse)
data.frame(TRAINING_DATA$samples, 
           labels=TRAINING_DATA$labels) |>
  ggplot(aes(x=X1, y=X2, color=labels)) + 
  geom_point(size=3)
```

We can make this a bit more attractive:

```{r}
MY_THEME <- theme_bw(base_size=20) + theme(legend.position="bottom")
theme_set(MY_THEME)
```

```{r}
data.frame(TRAINING_DATA$samples, labels=TRAINING_DATA$labels) %>%
  ggplot(aes(x=X1, y=X2, color=labels)) + 
  geom_point(size=2) + 
  ggtitle("Sample Realization of the Moons Dataset")
```

Much better!

Let's try making a simple prediction at the point (0, 0):

```{r}
knn(TRAINING_DATA$samples, 
    cl=TRAINING_DATA$labels, 
    test=data.frame(X1=0, X2=0), k=3)
```

::: {.callout-note title="Pause to Reflect" collapse="true"}

Does this match what you expect from the plot above? Why or why not?
The following image might help: 

```{r}
library(ggforce)
data.frame(TRAINING_DATA$samples, labels=TRAINING_DATA$labels) %>%
  ggplot(aes(x=X1, y=X2, color=labels)) + 
  geom_point() + 
  ggtitle("Sample Realization of the Moons Dataset") + 
  geom_circle(aes(x0=0, y0=0, r=0.2), linetype=2, color="red4")
```

Why is `R` returning a `factor` response here? What does that tell us about
the type of ML we are doing? 

:::

We can also visualize the output of KNN at every point in space:

```{r}
visualize_knn_boundaries <- function(training_data, k=NULL){
    xrng <- c(min(training_data$samples[,1]), max(training_data$samples[,1]))
    yrng <- c(min(training_data$samples[,2]), max(training_data$samples[,2]))
    
    xtest <- seq(xrng[1], xrng[2], length.out=101)
    ytest <- seq(yrng[1], yrng[2], length.out=101)
    
    test_grid <- expand.grid(xtest, ytest)
    colnames(test_grid) <- c("X1", "X2")
    
    pred_labels = knn(training_data$samples, 
                      cl=training_data$labels, 
                      test_grid, 
                      k=k)
    
  ggplot() + 
  geom_point(data=data.frame(TRAINING_DATA$samples, 
                             labels=TRAINING_DATA$labels), 
             aes(x=X1, y=X2, color=labels), 
             size=3) + 
  geom_point(data=data.frame(test_grid, pred_labels=pred_labels), 
             aes(x=X1, y=X2, color=pred_labels), 
             size=0.5) + 
  ggtitle(paste0("KNN Prediction Boundaries with K=", k))
}

visualize_knn_boundaries(TRAINING_DATA, k=1)
```

If we raise $K$, we get smoother boundaries: 

```{r}
visualize_knn_boundaries(TRAINING_DATA, k=5)
```

And if we go all the way to $K$ near to the size of the training data, 
we get very boring boundaries indeed:

```{r}
visualize_knn_boundaries(TRAINING_DATA, k=NROW(TRAINING_DATA$samples)-1)
```

::: {.callout-note title="Pause to Reflect" collapse="true"}

What does this tell us about the _complexity_ of KNN as a function of $K$?

:::

In the terminology we introduced above, we see that increasing $K$ 
decreases model complexity (wiggliness). 

Let's now see how training error differs as we change $K$: 

```{r}
TEST_DATA <- make_moons()

TRAINING_ERRORS <- data.frame()
for(k in seq(1, 20)){
    pred_labels_train <- knn(TRAINING_DATA$samples, cl=TRAINING_DATA$labels, TRAINING_DATA$samples, k=k)
    true_labels_train <- TRAINING_DATA$labels
    err <- mean(pred_labels_train != true_labels_train)
    cat(paste0("At k = ", k, ", the training (in-sample) error of KNN is ", round(100 * err, 2), "%\n"))
    TRAINING_ERRORS <- rbind(TRAINING_ERRORS, data.frame(k=k, training_error=err))
}

ggplot(TRAINING_ERRORS, aes(x=k, y=training_error)) + 
    geom_point() + 
    ggtitle("Training (In-Sample) Error of KNN")
```

Compare this to the test error: 

```{r}
TESTING_ERRORS <- data.frame()
for(k in seq(1, 20)){
    pred_labels_train <- knn(TRAINING_DATA$samples, cl=TRAINING_DATA$labels, TEST_DATA$samples, k=k)
    true_labels_train <- TEST_DATA$labels
    err <- mean(pred_labels_train != true_labels_train)
    cat(paste0("At k = ", k, ", the test (out-of-sample) error of KNN is ", round(100 * err, 2), "%\n"))
    TESTING_ERRORS <- rbind(TESTING_ERRORS, data.frame(k=k, test_error=err))
}

ggplot(TESTING_ERRORS, aes(x=k, y=test_error)) + 
    geom_point() + 
    ggtitle("Test (Out-of-Sample) Error of KNN")
```

The difference between the two is clearer if we put them on the same figure:

```{r}
ERRS <- inner_join(TRAINING_ERRORS, TESTING_ERRORS, by="k") |>
        pivot_longer(-k) |>
        rename(Error=value, Type=name)
ggplot(ERRS, aes(x=k, y=Error, color=Type)) + 
  geom_point() + geom_line()
```

We notice a few things here: 

1) Training Error increases in $K$[^rev], with 0 training error 
   at $K=1$. (Why?)
2) Test Error is basically always higher than test error
3) The best training error does not have the best test error

[^rev]: It is easier to understand this 'in reverse': as 
$K \downarrow 1$, training error decreases to 0, so as 
$K \uparrow n$, training error increases. 

We can also look at the gap between training and test error: this is called
`generalization error` or `optimism`: 

```{r}
inner_join(TRAINING_ERRORS, TESTING_ERRORS, by="k") |>
    mutate(optimism=test_error - training_error) |>
    ggplot(aes(x=k, y=optimism)) + 
    geom_point() + 
    geom_line()
```

::: {.callout-note title="Pause to Reflect" collapse="true"}

Consider the following questions:

- What is the relationship between optimism and model complexity
- What is the best value of $K$ for this data set?
- How should we pick the best value of $K$?
- How might that change if we increase the number of training samples?
- How might that change if we increase the number of test samples? 

:::

## Model Complexity (Redux)

Let us understand this using the tools of complexity we discussed above. 

- For large $K$, the model became quite simple, barely fit the training data, but
  did just as well on the training data as the test data (poorly on both). This
  is indicative of a *small generalization gap* and a *low complexity* model. 
- For python $K$, the model became rather complex, fit the training data well, 
  but didn't perform as well as we would like on the test data. This is indicative
  of a *large generalization gap* and a *high complexity* model. 
  
In this case, we could see the complexity visually by looking at the decision
boundaries (the lines separating predictions of class "A" from class "B"). This
isn't universally true[^olscmplx], but the intuition of "high complexity = more
wiggles" is usually pretty good.

The classical story of ML and complexity is given by something like this:[^fig]

![](https://mlstory.org/assets/gen-ushaped.svg)

For now, you can interpret "risk" as "test error" and
"empirical risk" as "training error". 

::: {.callout-caution title="A more complex story of complexity" collapse="true"}

Recent research has suggested that the story is not
quite so simple. Some methods exhibit a "single descent"
curve

![](https://mlstory.org/assets/gen-single.svg)

and some even have a "double descent" curve:

![](https://mlstory.org/assets/gen-double.svg)

The story of these curves is still an active topic
of research, but it's pretty clear that _very large_ 
and _very deep_ neural networks exhibit something
like a double descent curve. My own guess is that we're
not quite measuring 'complexity' correctly for these
incredibly complex models and that the classical story 
holds if we go measure complexity in the right way, but
this is far from a universally held belief. 

:::

So how do we measure complexity? For OLS, it seems
to be proportional to $p$, while for KNN it seems to
be inverse to $K$. In this class, we won't really 
focus too much on the actual measurements. For most of
the methods we study, it is usually pretty clear what
drives complexity up or down, even if we can't quite
put a number on it. 

### Stability and Generalization

It turns out there is a deep connection between
generalization and a suitable notion of "stability". 
A model is said to be relatively stable if changes
to an input point do not change the output predictions
significantly.[^be]

At an intuitive level, this makes sense: if the model
is super sensitive to individual inputs, it must be
very flexible and hence quite complex. (A very simple
model cannot be sensitive to all of its inputs.) 

We can apply this idea to understand some rules-of-thumb
and informal practices you might have seen in previous
statistics courses: 

- Regression leverage: [Leverage](https://en.wikipedia.org/wiki/Leverage_(statistics))
  measures how much a single data point can change 
  regression coefficients. This is stability!
  
- Removing outliers: we often define outliers as
  observations which have a major (and assumed
  corrupting) influence on our inferences. By removing
  outliers, we guarantee that the resulting inference
  is not too sensitive to any of the remaining data
  points. Here, the 'pre-step' of outlier removal
  increases stability (by changing sensitivity of those
  observations to zero) and hopefully makes our
  inferences more accurate (better generalization)
  
- Use of robust statistics, *e.g.* medians instead of
  means. These explicitly control the stability of
  our process.

Perhaps most importantly: this justifies why the 
large $n$ (big sample) limit seems to avoid overfitting.
If our model is fit to many distinct data points, it
can't be too sensitive to any of them. At least, that's
the hope...

The sort of statistical models you have seen to date --
so-called *parametric* models -- have a 'stabilizing' 
effect. By reducing lots of data to only a few
parameters, those parameters (and hence the model
output) can't depend too much on any individual
input point.[^thirty_rule] This 'bottleneck' in the
parameter space seems to improve performance. 

Other methods, like $1$-Nearest-Neighbor, become
increasingly more complex as we get more data and do
not benefit from this sort of 'bottleneck' effect. 

At this point, you might think that stability and
variance are closely related concepts - you are not 
wrong and we will explore the connection in more 
detail next week. 

[^be]: For details, see [O. Bousquet and A. Elisseeff, 
*Stability and Generalization* in **Journal of Machine
Learning Research 2**,
pp.499-526.](https://jmlr.org/papers/volume2/bousquet02a/bousquet02a.pdf)

[^thirty_rule]: Some regression textbooks advocate a
rule that you have 30 data points for every variable
in your model. This is essentially guaranteeing that
the $p/n$ ratio that controls the generalization of OLS
(see above!) stays quite small.


## Key Terms and Concepts

- Supervised vs Unsupervised
- Regression vs Classification
- Training Error
- Test Error
- Generalization Gap
- Complexity
- Overfitting
- Stability
- $K$-Nearest Neighbors
- Decision Boundaries

[^fig]: Figures from HR Chapter 6, *Generalization*.

[^olscmplx]: In OLS, the complexity doesn't make the model 'wigglier' in a normal
sense - it's still linear after all - but you can think of it as the additional
complexity of a 3D 'map' (*i.e.*, a to-scale model) vs a standard 2D map. 


[^rng]: Consider programs of the form: 

    ```
if x in TRAINING DATA: 
    return corresponding y
else:
    return SOME RANDOM FIXED VALUE
    ```
    
    In training error, these models are all perfect, but depending on the
    choice of "else" value, they will have very different test errors. And it's
    not at all hard to imagine that we can do much better than a single fixed
    value in the "else" branch. 
