---
session: "`r session <- 2; session`"
topic: "`r library(tidyverse); topic <- read_csv('../key_dates.csv', name_repair='universal') |> filter(Course.Element == 'Class Session', Item.Number==session) |> pull(Details); topic`"
date: "Last Updated `r Sys.Date()`"
---

$$\newcommand{\E}{\mathbb{E}} \newcommand{\R}{\mathbb{R}} \newcommand{\bx}{\mathbf{x}}\newcommand{\bbeta}{\mathbf{\beta}} \newcommand{\bX}{\mathbf{X}} \newcommand{\by}{\mathbf{y}} \newcommand{\bz}{\mathbf{z}} \newcommand{\bA}{\mathbf{A}} \newcommand{\bb}{\mathbf{b}} \newcommand{\bc}{\mathbf{c}} \newcommand{\bH}{\mathbf{H}} \newcommand{\bI}{\mathbf{I}} \newcommand{\bB}{\mathbf{B}} \newcommand{\bU}{\mathbf{U}} \newcommand{\bV}{\mathbf{V}} \newcommand{\bD}{\mathbf{D}}$$
```{r}
#| include: false
set.seed(9890)
```

## Linear Algebra Review

In mathematics, a _vector_ - random or otherwise - is a fixed-length
_ordered_ collection of numbers. When we want to be precise about the
size of a vector, we often call it a "tuple", *e.g.*, a length-three vector
is a "triple", a length-four vector is a "4-tuple", a length-five vector
is a "5-tuple" *etc.*. 

So, these are all vectors: 

- $(3, 4)$
- $(1, 1, 1)$
- $(1, 5, 6, 10)$

When we want to talk about the set of vectors of a given size, we use
the _Cartesian product_ of sets. For two sets, $A, B$, the product set
$A \times B$ is the set of all pairs, with the first element from $A$ and
the second from $B$. In mathematical notation, 

$$A \times B = \left\{(a, b): a \in A, b \in B\right\} $$

This _set-builder_ notation is read as follows: "The Cartesian Product of $A$
and $B$ is the set of all pairs $(a, b)$ such that $a$ is in $A$ and $b$
is in $B$." 

If $A$ and $B$ are the same set, we define a Cartesian _power_ as follows:

$$A^2 = A \times A = \left\{(a_1, a_2): a_1 \in A, a_2 \in A\right\}$$

Note that even though the sets $A$ and $A$ in this product are the same, 
the elements in each pair may vary. For example, if $A = \{1, 2, 3\}$, we have

$$A^2 = \left\{(1, 1), (1, 2), (1, 3), (2, 1), (2, 2), (2, 3), (3, 1), (3,2), (3, 3)\right\}$$

Note that vectors are _ordered_ pairs so $(2, 1) \neq (1, 2)$. From here, it
should be pretty easy to convince yourself that set sizes play nicely with
Cartesian products: 

- $|A \times B| = |A| |B|$
- $|A^k| = |A|^k$

The most common set of vectors we use are those where each element is an
arbitrary real number. The set of vectors of length $n$ ($n$-tuples) is thus
$\R^n$. We rarely mix vectors of different lengths, so we don't really have
a name or notation for the "combo pack" $\R^2 \cup \R^3 \cup \R^4$. 

Conventionally, vectors are written in bold (if on a computer) or with a
little arrow on top (hand written): so a vector called "x" would be
denoted $\mathbf{x}$ or $\vec{x}$. The elements of $\bx$ are denoted by 
subscripts $\bx = (x_1, x_2, \dots, x_n)$. 

### Vector Arithmetic

We have three arithmetic operations we can perform on general vectors. 
The simplest is _scalar multiplication_. A _scalar_ is a non-vector number, 
*i.e.*, a 'regular' number. Scalar multiplication consists of applying the
scalar independently to each element of a vector. 

$$\alpha\, \bx = \alpha\, (x_1, x_2, \dots, x_n) = (\alpha x_1, \alpha x_2, \dots, \alpha x_n)$$

For example, if $\bx = (3, 4)$ and $\alpha = 2$, we have
$$\alpha\, \bx = (6, 8)$$

Note that the output of scalar multiplication is always a vector of the
same length as the input. 

We also have the ability to add vectors. This again is performed element-wise.

$$\bx + \by = (x_1, \dots, x_n) + (y_1, \dots, y_n) = (x_1 + y_1, \dots, x_n + y_n) $$

Note that we can't add vectors of different lengths (recall our "no mixing"
rule) and the output length is always the same as the input lengths. 

Finally, we have the vector _inner product_, defined as: 

$$\langle \bx, \by \rangle = x_1y_1 + x_2y_2 + \dots + x_ny_n $$

You might have seen this previously as the "dot" product. The inner product
takes two length-$n$ vectors and gives back a scalar. This structure might
seem a bit funny, but as we'll see below, it's actually quite useful. 

You might ask if there's a "vector-out" product: there is one, with the
fancy name "Hadamard product", but it doesn't play nicely with other tools, 
so we don't use it very much. 

These tools play nicely together: 

- $\alpha(\bx + \by) = \alpha \bx + \alpha \by$ (Distributive)
- $\langle \alpha \bx, \by \rangle = \alpha \langle \bx, \by \rangle$ (Associative)
- $\langle \bx, \by \rangle = \langle \by, \bx \rangle$ (Commutative)

### Vector Length and Angle

We sometimes want to think about the "size" of a vector, analogous to
the absolute value of a scalar. In scalar-world, we say "drop the sign"
but there's not an obvious analogue to a sign for a vector. For instance, 
if $\bx = (3, -4)$ is $\bx$ "positive", "negative" or somewhere in beetween? 

We note a trick from scalar-land: $|x| = \sqrt{x^2}$. We can use the same idea
for vectors: 

$$ \|\bx\| = \sqrt{\langle \bx, \bx\rangle} = \sqrt{\sum_{i=1}^n x_i^2}$$

This quantity, $\|\bx\|$, is called the _norm_ or _length_ of a vector.
We use the double bars to distinguish it from the absolute value of
a scalar, but it's fundamentally the same idea. 

In $\R^2$, we recognize this formula for length as the Pythagorean theorem: 

$$ \|(3, 4)\| = \sqrt{3^2 + 4^2} = \sqrt{25} = 5 $$

We also sometimes want to define the _angle_ between two vectors. We can
define this as: 

$$ \cos \angle(\bx, \by) = \frac{\langle \bx, \by\rangle}{\|\bx\|\|\by\|} \Leftrightarrow \angle(\bx, \by) = \cos^{-1}\left(\frac{\langle \bx, \by\rangle}{\|\bx\|\|\by\|}\right)$$ 

We won't use this formula too often for implementation, but it's good to have
it for intuition. In particular, we note that angle is a proxy for sample
correlation, justifying the common vernacular of "orthogonal", meaning "at right
angles", for "uncorrelated" or "unrelated."

### Matrices

An $n$-by-$p$ array of numbers is called a _matrix_; here the first dimension
is the number of rows while the second is the number of columns. So 
$$\bA = \begin{pmatrix} 1 & 2 \\ 3 & 4 \\ 5 & 6 \end{pmatrix}$$
is a 3-by-2 matrix. We denote the set of matrices of a given size as 
$\R^{n \times p}$, extending slightly the notation we use for vectors. 

In this course, we will use matrices for two closely-related reasons: 

1) To organize our data
2) To specify and manipulate linear models

Specifically, if we have $n$ training samples, each of $p$ covariates
(predictors), we will arrange them in a matrix traditionally called 
$\bX \in \R^{n \times p}$. Here, each _row_ of $\bX$ corresponds to an 
observation. Statisticians tend to call this matrix a _design_ matrix because
(historically) it was something designed as part of an experiment; the name 
got carried forward into the observational (un-designed) setting. You may also
hear it called the 'data matrix'. 

Suppose we have a design matrix $\bX \in \R^{n \times p}$ and a vector of 
regression coefficients $\mathbf{\beta} \in \R^p$. We can use _matrix_
multiplication to make predictions about all observations simultaneously. 

Specifically, recall that the standard (multivariate) linear model looks like: 

$$\hat{y} = \sum_{i=1}^p x_i\beta_i$$

For a single observation, this can be written in vector notation as

$$\hat{y} = \bx^{\top}\bbeta$$ 

If we have $n$ observations, we can stack them in a vector as: 

$$\begin{pmatrix} \hat{y}_1 \\ \hat{y}_2 \\ \vdots \\ \hat{y}_n \end{pmatrix} = 
\begin{pmatrix} \bx_1^{\top}\bbeta \\ \bx_2^{\top}\bbeta \\ \vdots \\ \\bx_n^{\top}\bbeta \end{pmatrix}$$

We can connect this with our design matrix $\bX$ by using the above as a 
definition of matrix-vector multiplication: 

$$\hat{\by} = \bX\bbeta =  
\begin{pmatrix} \bx_1^{\top}\bbeta \\ \bx_2^{\top}\bbeta \\ \vdots \\ \\bx_n^{\top}\bbeta \end{pmatrix}$$

Here, matrix multiplication proceeds by taking the inner product of each _row_
with the (column) vector $\bbeta$. This may feel a bit unnatural at first, but
with a bit of practice, it will become second nature. Note that we don't need
a transpose on $\bX$: the multiplication 'auto-transposes' in some sense. 

It is always helpful to keep track of dimensions when doing matrix
multiplication: checking that matrices and vectors have the right size is a
useful way to make sure you haven't done anything _too wrong_. In general, 
we can only multiply a $m$-by-$n$ matrix with a $n$-by-$p$ matrix and the result
is a $m$-by-$p$ matrix (the $n$-dimension gets reduced to a scalar by the inner
product). Formally, we have something like

$$ \R^{m \times n} \times \R^{n \times p} \to \R^{m \times p}$$

For purposes of this, you can always think of an $n$-vector as a $n$-by-1 
"column" matrix, giving us: 

$$\underbrace{\bX}_{\R^{n \times p}} \underbrace{\bbeta}_{\R^p} = \underbrace{\hat{\by}}_{\R^{n \times 1}}$$

### Matrix as Linear Function

In this set up, notice that $\bX$ plays an interesting role: it takes one vector
$\bbeta$ and outputs a different vector $\by$. Looking at it this way, we can 
think of $\bX$ as a _function_: 

$$\bX: \bbeta \in \R^p \to \by \in \R^n$$

This is a bit unnatural from a statistical point of view: we tend to think of
the parameters ($\bbeta$) acting on the data (each $\bx$) to produce outputs,
but mathematically it's entirely natural. 

To step away from the 'data' interpretation, let's just talk about a general
matrix $\bA \in \R^{m \times n}$. We can (pre-)multiply any $\bw \in \R^{m}$ by
$\bA$ to get back a new vector $\bz \in \R^n$. Symbolically, this is just

$$\bz = \bA \bw$$

which is much easier to say and to write. The _function_ defined by $\bA$ has
several nice properties: 

- it plays nicely with addition: 

  $$\begin{align*}
  \bz_1 &= \bA \bw_1 \\
  \bz_2 &= \bA \bw_2 \\
  \bz_1 + \bz_2 &= \bA\bw_1 + \bA\bw_2 \\
  &= \bA(\bw_1 + \bw_2)
  \end{align*}$$
  
  (Why is this true?)
- it plays nicely with scalar multiplication: 

  $$ \bA(\alpha \bw) = \alpha (\bA \bw)$$
  
  (Again, why is this true?)
  
Taken together, we say that these properties make $\bA$ a _linear_ function or
a _linear_ operator. Note that, in this context, _linear_ is maybe a bit more
restrictive than the 'middle school' definition. Our properties require "zero-in/
zero-out": if we take $\alpha = 0$ in the scalar multiplication rule, we have: 

$$\bA(0 \, \bw) = 0\, (\bA \bw) \implies \mathbf{0}_n = \bA \mathbf{0}_m$$

where $\mathbf{0}_{m}, \mathbf{0}_n$ are the all-zero vectors of length $m, n$
respectively. As such, a 'linear' function you might have seen before like

$y = 2x + 3$ 

is not linear in our sense because it maps $0$ to $3$, not back to $0$ 
Linear + constant is called "affine" in advanced mathematics, so middle school
'linear functions' are really affine.

This might feel like a limitation, but true (non-affine) linearity is incredibly
useful in mathematical modeling as it _guarantees_ us that rules like "double the
input, double the output" or "negate the input, negate the output" will always
hold. These types of relationships are so powerful and ubiquituous that this
slightly-restrictive is definitely worth using.[^var]

[^var]: In addition to the 'physics' of it (push twice as hard, go twice as far),
we are used to these types of linear type relationships in statistics: if we 
have some measurements in yards and re-express them in feet by multiplying by 
three, it's very useful that the standard deviation is just multipled by 3
without any added terms. 

While it is clear that any 'matrix multiply function' is a linear function, it
actually turns out that any linear function can be written as a matrix 
multiplication![^rev] This deep fact is the fundamental reason why linear algebra
is such a beautiful and beloved part of mathematics: any linear function can be
expressed using a small set of numbers--great for computational work--and any
set of numbers arranged in a rectangular array can be understood as a function.
By giving simultaneous to 'data tools' and 'function tools', we have a lot of
power at our fingertips. 

[^rev]: This is not too hard to show, but the book-keeping is annoying. In 
essence, given a linear function, we 'build' a matrix for it by seeing how it
applies to the standard basis vectors and then, by linearity, argue that that
matrix has to give the same value as the function for all possible inputs. 

### Combining Matrices

Given two matrices, how might we combine them? For addition, the answer is
relatively obvious: we simply add the corresponding elements together, which
we can write symbolically as: 

$$(\bA + \bB)_{ij} = \bA_{ij} + \bB_{ij}$$

This reads as "the $(i, j)$ element of the matrix $\bA + \bB$ is the $(i, j)$ element
of matrix $\bA$ plus the $(i, j)$ element of matrix $\bB$. From this definition 
it is clear that we can only add together matrices of corresponding dimension. 

This follows from the idea that a matrix is just a set of vectors "squished 
together" and our rule for vector addition. 

Scalar multiplication again is straightforward and follows from the corresponding
vector operation: 

$$(\alpha\, \bA)_{ij} = \alpha\, \bA_{ij}$$

With these, we can define the (additive) identity: that is, the matrix which, 
when added to another doesn't change the value, by $\mathbf{0}_{m \times n}$, 
the $m \times n$ matrix of all zeros. The additive inverse (the matrix that
you add to get back to the additive identity) is just 

$$(-\bA)_{ij} = -(\bA_{ij})$$

*i.e.*, elementwise negation. 

Multiplication is trickier: recall that a matrix $\bA$ also defines a function
$f_{\bA}. If we want matrix multiplication to follow normal associativity rules,
we need to ensure

$$\bA(\bB \bx) = (\bA\bB) \bx \quad \text{ for all } \bx, \bA, \bB$$

Achieving this is a little less obvious. First, let's look at the dimensions, 
suppose $\bx \in \R^{n}$ and $\bB \in \R^{m \times n}$. This gives us 
$\bB\bx \in \R^m$ so the second dimension of $\bA$ must be $m$. We can actually
allow arbitrary size for the first dimension of $\bA$ so we take 
$\bA \in \R^{l \times m}$, $\bB \in \R^{m \times n}$, and $\bx \in \R^n$. 
From this we see that $\bA, \bB$ don't have to be the same dimension (as we 
required) for addition, only that they _conform_, *i.e.*, the number of columns
of $\bA$ equals the number of rows of $\bB$. 

From here, it is possible, albeit cumbersome, to show that

$$(\bA\bB)_{ij} = \langle \bA_{i\cdot}, \bB_{\cdot j} \rangle = \sum_{k} \bA_{ik}\bB_{kj}$$

Here the $\cdot$ subscript means "take the whole row/column". This definition
is not so simple, but if you take a look at it, you'll see that it's really built
up from a bunch of vector multiplications, so maybe it's not too weird. 

Looking at our definition, we note that $\bA\bB$ does not necessarily equal 
$\bB\bA$. In fact, it's entirely possible for $\bB\bA$ to not even exist if
the sizes don't match up. 

A multiplicative identity is not too hard to identify: we call it $\bI$ and it
is given by 

$$\bI = \begin{pmatrix} 1 & 0 & 0 & \dots & 0 \\
                        0 & 1 & 0 & \dots & 0 \\
                        0 & 0 & 1 & \ddots & 0 \\
                        0 & 0 & 0 & \dots & 1 \end{pmatrix}$$
                        
That is, $\bI$ is a square matrix with ones on the diagonal and zeros everywhere
else. You should be able to convince yourself that $\bI\bx = \bx$ for all
$\bx$: check that you can show this directly (by the definition of matrix
multiplication) or conceptually (thinking in terms of functions, what do we need
$\bI$ to look like as a "do nothing" function). 

Matrix (multiplicative) inverses are harder to say much about. Given a matrix
$\bA$, we define its inverse as a matrix $\bA^{-1}$ that satisfies
$\bA\bA^{-1}=\bA^{-1}\bA = \bI$. If such a matrix exists, it is unique and so
we can justify referring to $\bA^{-1}$, but we don't know whether it actually 
always exists. It's not easy to just look at a matrix and tell whether it has
an inverse, but we will discuss some conditions below. 

The potential non-existence of inverses shouldn't bother you _too much_ - it's 
the matrix equivalent of the fact that $0^{-1}$ isn't defined for 'regular' 
numbers - but it is an annoying little technical thing we need to be aware of. 

### Rotation / Orthognal Matrices

While all linear functions are nice, certain operations are particularly nice
and get their own names. One is the concept of 'rotation': operations which 
take a $n$-vector  and return a different $n$-vector with the same length.[^rot]
(No stretching, only turning - hence rotation.) Clearly, these are _square_ 
matrices since they go from $\R^n$ to $\R^n$ and hence are $n\times n$ matrices.
Less obvious is 

[^rot]: I am oversimplifying here: there are also 'reflections' that satisfy
this requirement (*e.g.*, a matrix mapping $(0, 1)$ to $(0, -1)$ but 
$(1, 0)$ to $(1,0)$ - this flips across the $y$-axis but isn't a "rotation"). 
This distinction doesn't really come up in our class and the more general 
category of "orthogonal" is all we really need anyways.

A somewhat remarkable fact is that rotation matrices have a very simple inverse:
the matrix _transpose_. That is, the matrix you get by flipping across the diagonal

$$\bA^{\top}_{ij} = \bA_{ji}.$$

If a matrix satisfies $\bA^{\top}\bA = \bA\bA^{\top} = \bI$, we refer to it as
_orthogonal_. These orthogonal matrices are the matrix equivalent of the roots
of unity (*i.e.*, roots of $+1$ for numbers: $\pm 1$ for square roots, 
$1, -0.5 \pm \sqrt{0.75}i$ for cube roots, *etc.*). 

### Matrix Decompositions

A matrix is a rather complex thing, so we can often understand them better
through the use of matrix _decompositions_: that is, breaking an arbitrary
matrix into simpler parts. There are many decompositions that are useful, but 
the only one we will use regularly in this class is the 
_singular value decomposition_ or SVD. Given a matrix $\bA \in \R^{m \times n}$, 
its SVD is a three factor decomposition 

$$\bA = \bU\bD\bV^{\top}$$

where 

- $\bU \in \R^{m \times m}$ is an $m \times m$ orthogonal matrix
- $\bD$ is an $\R^{\m \times n}$ matrix which is all zeros except on the
  main diagonal, which is non-negative. 
- $\bV \in \R^{n \times n}$ is an $n \times n$ orthogonal matrix

(There are other set ups where we make $\bD$ square and one of $\bU, \bV$ 
non-square, but this is typically the easiest form to work with). 

*E.g.*, for a random matrix: 

```{r}
A <- matrix(rnorm(35), nrow=7, ncol=5)
A
```

We have an SVD given by: 

```{r}
#| echo: false
SVD <- svd(A, nu=dim(A)[1], nv=dim(A)[2])

U <- SVD$u
V <- SVD$v
D <- diag(SVD$d, nrow=NROW(A), ncol=NCOL(A))
```

```{r}
U
```

which satisfies[^zapsmall]:

[^zapsmall]: Computers are not perfectly accurate, so if you do this calculation
yourself, you'll get numbers around $\pm 3 \times 10^{-16}$. These are basically
zero and we can use the `zapsmall` function to set them to zero. 

```{r}
zapsmall(U %*% t(U))
```

and 

```{r}
zapsmall(t(U) %*% U)
```

and 

```{r}
V
```

satisfying

```{r}
zapsmall(V %*% t(V))
```

and

```{r}
D
```

which give us

```{r}
all.equal(U %*% D %*% t(V), A)
```

The non-zero elements of $\bD$ are known as the _singular values_ of the matrix,
hence the name SVD and they are roughly analogous to the absolute value
of a standard number (except there are more of them!). In particular, they are
non-negative and, when they are zero, we have an issue around division as we
discuss below. 

The SVD is a super useful way to 'break a matrix' into very simple pieces: 
multiplication by orthogonal matrices (roughly equivalent to multiplying by 1)
two times and multiplying by a non-negative diagonal matrix (essentially, just
stretching a vector). The fact that _any_ matrix and hence any linear function
can be broken up into "rotate -> stretch -> rotate" is actually quite remarkable.

### SVDs of Important Matrices

Let's talk through the SVD of different matrices: if we start with the all
zeros matrix ($\mathbf{0}_{m \times n}$), it's clear that the 'stretch factors'
should all be zero, since the zero matrix sends _everything_ to zero. Hence

$$\mathbf{0}_{m \times n} = \bU \mathbf{0}_{m \times n} \bV^{\top}$$

where any orthogonal $\bU, \bV$ will work (so the SVD is not unique). 

Similarly, we can look at the SVD of $\bI$. In this case, we have a square matrix
so $\bU, \bV$ must both be $n \times n$ matrices. Since the identity matrix 
doesn't change its inputs, its 'stretch factors' must all be 1 (no stretching
or shrinking) so we get: 

$$\bI = \bU \bI \bV^{\top}$$

To make this hold, it suffices to take $\bU = \bV$ and to recall that these
are orthogonal matrices so $\bU \bV^{\top} = \bU\bU^{\top} = \bI$, as desired.
Geometrically this makes sense, a general recipe for "do nothing" (identity) is
to rotate, do a "non-stretch" (multiply by 1) and to rotate back to where you
started. 

Finally, consider a general matrix $\bA = \bU \bD\bV^{\top}$. Suppose that
$\bA^{-1}$  exists (which it might not) - what would the SVD of $\bA^{-1}$ look 
like? Well, let's define $\bA^{-1} = \tilde{\bU}\tilde{\bD}\tilde{\bV}$ as
the SVD of $\bA^{-1}$. We then have: 

$$\bI = \bA\bA^{-1} = \bU\bD\bV^{\top}\tilde{\bU}\tilde{\bD}\tilde{\bV}^{\top}$$

To get terms to start cancelling out, let's take $\tilde{\bU} = \bV$ 
(since $\bV^{\top}\bV = \bI$). Similarly, if we start with $\bI = \bA^{-1}\bA$, 
we get $\tilde{\bV} = \bU^{\top}$. Putting these together, we have the 
pair of SVDs: 

$$\begin{align*}
\bA &= \bU \bD \bV^{\top} \\
\bA^{-1} &= \bV \tilde{\bD} \bU^{\top}
\end{align*}$$

So now we just need to have $\bD\tilde{\bD}$ cancel ($\bD\tilde{\bD} = \bI$). 
This is hard in general, but since $\bD, \tilde{\bD}$ are diagonal matrices,
finding the inverse is not hard: 

$$\tilde{\bD}_{ii} = 1/\bD_{ii}$$

Putting these together, we have: 

$$\bA^{-1} = \bV\bD^{-1}\bU^{\top}$$

where $\bD^{-1}$ is found by taking elementwise inverses of $\bD$ on the diagonal
and leaving zeros elsewhere. This recipe is quite powerful - and it actually 
shows us exactly the way in which a matrix can fail to have an inverse. If 
one or more of the singular values are 0, we have a "divide by zero" problem if
we try to compute $\bA^{-1}$ using the SVD of $\bA$. This is basically the 
_only_ way that a square matrix can fail to have an inverse: other 
characterizations you'll see (zero determinant, degenerate spectrum, *etc*) are 
just other ways of saying the same thing. 

Thinking about this SVD/inverse formula using our 

### Spectral Properties of Symmetric Matrices

An important class of matrices we will consider are _symmetric_ matrices, which
are just what the name sounds like. These come up in several key places in
statistics, none more important than the _covariance_ matrix, typically denoted
$\Sigma$. Recall that the covariance operator $\mathbb{C}$ is symmetric
($\mathbb{C}[X, Y] = \mathbb{C}[Y, X]$) so the covariance matrix of a random
vector turns out to be symmetric as well. 

Another common source of symmetric matrices is when a matrix is multiplied by
its transpose: you should convince yourself that 
$\bX^{\top}\bX \in \R^{p \times p}$ is a symmetric matrix. 



## Ordinary Least Squares

Suppose we want to fit a linear model to a given data set (for any
of the reasons we discuss in more detail below): how can we
choose which line to fit? (There are infinitely many!)

Since our goal is minimizing test error, and we hope training error
is at least a somewhat helpful proxy for training error, we can pick
the line that *minimizes training error*.  To do this, we need to 
commit to a specific measure of error. As the name *Ordinary Least
Squares* suggests, OLS uses (mean) squared error as its target. 

Why is MSE the right choice here? It turns out that MSE is very nice
computationally, but the reason is actually more fundamental: given a
random variable $Z$, suppose we want to minimize the $(Z - \mu)$ for 
some $\mu$: it can be shown that

$$ \E[Z] = \text{argmin}_{\mu \in \R} \E[(Z - \mu)^2]$$


::: {.callout-note title="Pause to Reflect" collapse="true"}

Prove this for yourself. 

*Hint*: Differentiate with respect to $\mu$. 

:::

That is, the quantity that minimizes the MSE is the mean. So when we
fit a line to some data by OLS, we are implicitly trying to fit to
$\E[y]$ - a very reasonable thing to do!

Specifically, given training data $\mathcal{D} = \{(\bx_i, y_i)\}_{i=1}^n$ where each $\bx_i \in \R^p$, OLS finds $\bbeta \in \R^p$ such that
$$ \hat{\bbeta} = \text{argmin}_{\bbeta \in \R^p} \frac{1}{n}\sum_{i=1}^n \left(y_i - \sum_{j=1}^p \beta_jx_j\right)^2 = \text{argmin}_{\bbeta \in \R^p} \frac{1}{n}\sum_{i=1}^n \left(y_i - \bx^{\top}\bbeta\right)^2$$

Some possibly new notation here: 

  - $\bx^{\top}\bbeta$ is the (inner) product of two vectors: defined as the sum of their elementwise products. 
  - Optimization problems: 
    - $$\hat{x} = \text{argmin}_{x \in \mathcal{C}} f(x)$$
    - $$f_* = \text{min}_{x \in \mathcal{C}} f(x)$$
     
    These problems say: find the value of $x$ in the set 
    $\mathcal{C}$ that minimizes the function $f$. $\text{argmin}$ 
    says 'give me the minimizer' while $\min$ says
    give me the minimum value'. These are related by 
    $f_* = f(\hat{x})$
    
    The function $f$ is called the *objective*; the set $\mathcal{C}$ 
    is called the *constraint set*.
    
*Ordinary Least Squares* refers to the use of an MSE objective
without any additional constraints.

Note the general structure of our approach here: 

- Define the loss we care about
- Set up an optimization problem to minimize loss on the training set
- Solve optimization problem

ML folk call this *empirical risk minimization* (ERM) since we're
minimizing the risk (average loss) on the data we can see (the
training data). Statisticians call this $M$-estimation, since it
defines an estimator by **M**inimization of a measure of 'fit'.
Whatever you call it, it's a very useful 'meta-method' for coming
up with ML methods. 


::: {.callout-note title="Pause to Reflect" collapse="true"}


1. How does this compare with *Maximum Likelihood Estimation*? 

2. We set up this ERM method using *mean squared error* - what happens 
   with other errors? Specifically, formulate this ERM for 

   - *mean absolute error*
   - *mean percent error*
   
   and compare to OLS. 
   
:::

So far we've set up OLS as 
$$ \hat{\bbeta} = \text{argmin}_{\bbeta \in \R^p} \frac{1}{n}\sum_{i=1}^n \left(y_i - \bx^{\top}\bbeta\right)^2$$

We can clean this up to make additional analysis easier: 

- Let $$\by = \begin{pmatrix} y_1 \\ y_2 \\ \vdots \\ y_n \end{pmatrix}$$ 
  be the (vertically stacked) vector of responses. 
- Next look at our predictions: 
  $$\hat{\by} = \begin{pmatrix} \bx_1^{\top}\bbeta \\ \bx_2^{\top}\bbeta \\ \vdots \\ \bx_n^{\top}\bbeta \end{pmatrix} = \begin{pmatrix} \bx_1^{\top} \\ \bx_2^{\top} \\ \vdots \\ \bx_n^{\top} \end{pmatrix}\bbeta = \bX\bbeta$$
  
Hence, OLS is just $$\hat{\bbeta} = \text{argmin}_{\bbeta \in \R^p} \frac{1}{n} \|\by - \bX\bbeta\|_2^2$$ Here $\|\cdot\|_2^2$ is the (squared Euclidean or $L_2$) norm of a vector: defined by $\|\bz\|_2^2 = \sum z_i^2$


::: {.callout-note title="Pause to Reflect" collapse="true"}


1. Do we actually need the $1/n$ part? 

2. Why is this called linear?

:::

We've formulated OLS as 
$$\hat{\beta} = \text{argmin}_{\bbeta \in \R^p} \frac{1}{n} \|\by - \bX\bbeta\|_2^2.$$ 
In order to solve this, it will be useful to modify it slightly to
$$\hat{\beta} = \text{argmin}_{\bbeta \in \R^p} \frac{1}{2} \|\by - \bX\bbeta\|_2^2$$ 

::: {.callout-note title="Pause to Reflect" collapse="true"}

Why is this ok to do?

:::

You will show in [Report #01](../reports/repot01.html) that the solution is
given by 

$$\hat{\beta} = (\bX^{\top}\bX)^{-1}\bX^{\top}\by$$

::: {.callout-note title="Pause to Reflect" collapse="true"}

Prove this for yourself. What conditions are required on $\bX$ for the
inverse in the above expression to exist?

:::

With this estimate, our in-sample predictions are given by: 

$$\hat{\by} = \bX\hat{\bbeta} = \bX(\bX^{\top}\bX)^{-1}\bX^{\top}\by $$

This says that our predictions are, in some sense, just a linear function of
the original data $\by$. The matrix

$$\bX(\bX^{\top}\bX)^{-1}\bX^{\top} = \bH$$

is sometimes called the 'hat' matrix because it puts a hat on $\by$
($\hat{\by}=\bH\by$). 

$\bH$ can be shown to be a special type of matrix called a _projector_, meaning
that it has eigenvalues all 0 or 1. For the hat matrix specifically, we can show 
that eigenvalues of $\bH$ are $p$ zeros and $n - p$ ones. 

This in turn implies that it is _idempotent_, meaning $\bH^2 = \bH\bH = \bH$. 
(To show this, simply express $\bH$ in terms of its eigendecomposition.)
We can use this property to finally justify the in-sample MSE of OLS we have
cited several times. 

The in-sample MSE is given by: 

$$\begin{align*}
\text{MSE} &= \frac{1}{n}\|\by - \hat{\by}\|_2^2 \\
           &= \frac{1}{n}\left\|\by - \bH\by\right\|_2^2 \\
           &= \frac{1}{n}(\by - \bH\by)^{\top}(\by - \bH\by) \\
           &= \frac{1}{n}(\by^{\top} - \by^{\top}\bH)(\by - \bH\by) \\
           &= \frac{\by^{\top}\by - \by^{\top}\bH\by - \by^{\top}\bH\by + \by^{\top}\bH\bH\by}{n} \\
           &= \frac{\by^{\top}\by - \by^{\top}\bH\by}{n} \\
           &= \frac{\by^{\top}(\bI - \bH)\by}{n} \\
\implies \E[\text{MSE}] &= \E\left[\frac{\by^{\top}(\bI - \bH)\by}{n}\right] \\
&= \frac{1}{n}\E\left[\by^{\top}(\bI - \bH)\by\right]
\end{align*}$$

To finish this, we need to know that the [expectation of a symmetric quadratic 
form](https://en.wikipedia.org/wiki/Quadratic_form_(statistics)) satisfies 

$$\bx \sim (\mu, \Sigma) \implies \E[\bx^{\top}\bA\bx] = \text{Tr}(\bA\Sigma) + \mu^{\top}\bA\mu$$

for any random vector $\bx$ with mean $\mu$ and covariance matrix $\Sigma$. 

To apply this above, we note $\by \sim (\bX\beta_*, \sigma^2 \bI)$, so we get

$$\begin{align*}
\E\left[\by^{\top}(\bI - \bH)\by\right] &= \text{Tr}((\bI - \bH) \sigma^2 \bI) + (\bX\bbeta_*)^{\top}(\bI - \bH)(\bX\bbeta_*) \\
&= \sigma^2 \text{Tr}(\bI - \bH) + \bbeta_*^{\top}\bX^{\top}(\bI - \bH)\bX\bbeta_* \\
&= \sigma^2 \text{Tr}(\bI - \bH) + \bbeta_*^{\top}(\bX^{\top}\bX - \bX^{\top}\bH\bX)\bbeta_* \\ 
&= \sigma^2 \text{Tr}(\bI - \bH) + \bbeta_*^{\top}(\bX^{\top}\bX - \bX^{\top}\bX(\bX^{\top}\bX)^{-1}\bX^{\top}\bX)\bbeta_* \\ 
&= \sigma^2 \text{Tr}(\bI - \bH) + \bbeta_*^{\top}(\bX^{\top}\bX - \bX^{\top}\bX)\bbeta_* \\ 
&= \sigma^2 \text{Tr}(\bI - \bH) + \bbeta_*^{\top} \mathbf{0}\bbeta_* \\
&= \sigma^2 \text{Tr}(\bI - \bH) \\
&= \sigma^2 (\text{Tr}(\bI) - \text{Tr}(\bH))
\end{align*}$$

Recall that the trace is simply the sum of the eigenvalues, so this last term
becomes $\sigma^2(n - p)$, finally giving us: 

$$\E[\text{MSE}] = \frac{\sigma^2(n-p)}{n}  = \sigma^2\left(1 - \frac{p}{n}\right)$$

Whew! That was a lot of work! But can you imagine how much more work this would
have been without all of these matrix tools? 

::: {.callout-note title="Pause to Reflect" collapse="true"}

Make sure you can justify every step in the derivation above. This is a
particularly long computation, but we will use the individual steps many more
times in this course.

:::

## Bias-Variance Decomposition

In [Report #01](../reports/repot01.html), you will show that, under MSE
loss, our expected test error can be decomposed as

$$\text{MSE} = \text{Bias}^2 + \text{Variance} + \text{Irreducible Error}$$

Let's show how we can analyze these quantities for a KNN *regression* problem. Here, we're using the 'regression' version of KNN since it
plays nicely with MSE.[^tradeoff]

[^tradeoff]: The Bias-Variance decomposition (and tradeoff) holds
approximately for other loss functions, though the math is only this
nice for MSE. 

```{r}
#| warning: false
# Don't attach the package to avoid weird aliasing issues
knn.reg <- FNN::knn.reg 
args(knn.reg)
```

We also need a 'true' function which we're trying to estimate. 
Let's use the following model:

$$\begin{align*}
X &\sim \mathcal{U}([0, 1]) \\
Y &\sim \mathcal{N}(4\sqrt{X} + 0.5 * \sin(4\pi * X), 0.25)
\end{align*}$$

That is, $X$ is uniform on the unit interval and $Y$ is a non-linear
function of $X$ plus some Gaussian noise. 

First let's plot $X$ vs $\E[X]$ - under MSE loss this is our 'best' (Bayes-optimal) possible guess. 

```{r}
yfun <- function(x) 4 * sqrt(x) + 0.5 * sinpi(4 * x)

x <- seq(0, 1, length.out=101)
y_mean <- yfun(x)

plot(x, y_mean, type="l", 
     xlab="X", ylab="E[X]", 
     main="True Regression Function", cex.lab=1.5)
```

To generate training data from this model, we simply implement the PRNG components:

```{r}
x_train <- runif(25, 0, 1) # 25 training points
y_train <- rnorm(25, mean=yfun(x_train), sd=0.5)


plot(x, y_mean, type="l", 
     xlab="X", ylab="E[X]", 
     cex.lab=1.5)

points(x_train, y_train)
```

We have some variance of course, but we can still "squint" to get the
right shape of our function. Let's see how KNN looks on this data. 

We start with $K=1$:

```{r}
X_train <- matrix(x_train, ncol=1)
plot_grid <- matrix(seq(0, 1, length.out=101), ncol=1)
Y_hat <- knn.reg(train=X_train, y=y_train, k=1, test=plot_grid)$pred

plot(x, y_mean, type="l", 
     xlab="X", ylab="E[X]", 
     cex.lab=1.5)

points(x_train, y_train)

lines(plot_grid, Y_hat, col="red4", lwd=2)
```

This is not a great fit - what happens if we repeat this 
process may times? 

```{r}
plot(x, y_mean, type="l", 
     xlab="X", ylab="E[X]", 
     cex.lab=1.5)

for(i in seq(1, 20)){
    X_train <- matrix(runif(25, 0, 1), ncol=1)
    y_train <- matrix(rnorm(25, mean=yfun(X_train), sd=0.5), ncol=1)
    test_grid <- matrix(seq(0, 1, length.out=101), ncol=1)
    Y_hat <- knn.reg(train=X_train, y=y_train, k=1, test=test_grid)$pred
    lines(test_grid, Y_hat, col="#FFAA0099", lwd=0.5)
}
```

Clearly we have some variance!

If we repeat with a higher value of $K$, we see far less variance:

```{r}
plot(x, y_mean, type="l", 
     xlab="X", ylab="E[X]", 
     cex.lab=1.5)

for(i in seq(1, 20)){
    X_train <- matrix(runif(25, 0, 1), ncol=1)
    y_train <- matrix(rnorm(25, mean=yfun(X_train), sd=0.5), ncol=1)
    plot_grid <- matrix(seq(0, 1, length.out=101), ncol=1)
    Y_hat <- knn.reg(train=X_train, y=y_train, k=10, test=test_grid)$pred
    lines(test_grid, Y_hat, col="#FFAA0099", lwd=0.5)
}
```

How well does KNN do _on average_? 

That is, if we could repeat this process (infinitely) many times, how
well would it recover the true regression function? Let's try $K=1$
and $K=10$:

```{r}
plot(x, y_mean, type="l", 
     xlab="X", ylab="E[X]", 
     cex.lab=1.5)



KNN_AVERAGE_PRED_K1 <- rowMeans(replicate(500, {
    X_train <- matrix(runif(25, 0, 1), ncol=1)
    y_train <- matrix(rnorm(25, mean=yfun(X_train), sd=0.5), ncol=1)
    plot_grid <- matrix(seq(0, 1, length.out=101), ncol=1)
    knn.reg(train=X_train, y=y_train, k=1, test=test_grid)$pred
}))

lines(test_grid, KNN_AVERAGE_PRED_K1, col="red4")


KNN_AVERAGE_PRED_K10 <- rowMeans(replicate(500, {
    X_train <- matrix(runif(25, 0, 1), ncol=1)
    y_train <- matrix(rnorm(25, mean=yfun(X_train), sd=0.5), ncol=1)
    plot_grid <- matrix(seq(0, 1, length.out=101), ncol=1)
    knn.reg(train=X_train, y=y_train, k=10, test=test_grid)$pred
}))

lines(test_grid, KNN_AVERAGE_PRED_K10, col="blue4")
```

We see here that, _on average_, KNN with $K=1$ (red) basically gets the
function just right - no bias!

On the other hand, because KNN with $K=10$ smooths out the function, we
see systematic errors (here oversmoothing). That's some bias. 

So which is better?
- $K=1$ - High variance, but low bias
- $K=10$ - Low variance, but high bias

We'll have to look at some test error to see. For now, we'll generate
our test data exactly the same way as we generate our training data:

```{r}
KNN_K1_ERROR <- replicate(500, {
    X_train <- matrix(runif(25, 0, 1), ncol=1)
    y_train <- matrix(rnorm(25, mean=yfun(X_train), sd=0.5), ncol=1)
    plot_grid <- matrix(seq(0, 1, length.out=101), ncol=1)
    
    # Generate from same model as before
    X_test <- matrix(runif(25, 0, 1), ncol=1) 
    y_test <- matrix(rnorm(25, mean=yfun(X_train), sd=0.5), ncol=1)
    
    y_hat <- knn.reg(train=X_train, y=y_train, k=1, test=X_test)$pred
    
    err = y_test - y_hat
})

KNN_K1_MSE <- mean(rowMeans(KNN_K1_ERROR^2))

KNN_K10_ERROR <- replicate(500, {
    X_train <- matrix(runif(25, 0, 1), ncol=1)
    y_train <- matrix(rnorm(25, mean=yfun(X_train), sd=0.5), ncol=1)
    plot_grid <- matrix(seq(0, 1, length.out=101), ncol=1)
    
    # Generate from same model as before
    X_test <- matrix(runif(25, 0, 1), ncol=1) 
    y_test <- matrix(rnorm(25, mean=yfun(X_train), sd=0.5), ncol=1)
    
    y_hat <- knn.reg(train=X_train, y=y_train, k=10, test=X_test)$pred
    
    err = y_test - y_hat
})

KNN_K10_MSE <- mean(rowMeans(KNN_K10_ERROR^2))

cbind(K1_MSE = KNN_K1_MSE, 
      K10_MSE = KNN_K10_MSE)
```

$K=10$ does better overall!

But does it do better _everywhere_ or are some parts of the problem
better for $K=1$?

Now we'll be systematic in our test data - spacing it equally on the
grid and computing 'pointwise' MSE: 

```{r}
KNN_K1_ERROR <- replicate(500, {
    X_train <- matrix(runif(25, 0, 1), ncol=1)
    y_train <- matrix(rnorm(25, mean=yfun(X_train), sd=0.5), ncol=1)
    
    # Generate from same model as before
    test_grid <- seq(0, 1, length.out=101)
    X_test <- matrix(runif(test_grid, 0, 1), ncol=1) 
    y_test <- matrix(rnorm(test_grid, mean=yfun(X_test), sd=0.5), ncol=1)
    
    y_hat <- knn.reg(train=X_train, y=y_train, k=1, test=X_test)$pred
    
    err = y_test - y_hat
})

KNN_K1_MSE <- rowMeans(KNN_K1_ERROR^2)

KNN_K10_ERROR <- replicate(500, {
    X_train <- matrix(runif(25, 0, 1), ncol=1)
    y_train <- matrix(rnorm(25, mean=yfun(X_train), sd=0.5), ncol=1)
    
    test_grid <- seq(0, 1, length.out=101)
    X_test <- matrix(runif(test_grid, 0, 1), ncol=1) 
    y_test <- matrix(rnorm(101, mean=yfun(X_test), sd=0.5), ncol=1)
    
    y_hat <- knn.reg(train=X_train, y=y_train, k=10, test=X_test)$pred
    
    err = y_test - y_hat
})

KNN_K10_MSE <- rowMeans(KNN_K10_ERROR^2)

plot(KNN_K1_MSE, col="blue4", pch=16, ylim=c(0, 1))
points(KNN_K10_MSE, col="red4", pch=16)
```

It looks like - for this set up at least - $K=10$ is better
_everywhere_ but that's not always the case.

Play around with the size of the training data, the noise in the 
samples, and the data generating function (`yfun`) to see if you 
can get different behavior. 

Also - is $K=10$ really the optimal choice here? What would happen
if we changed $n$?

So, now that we have a good sense of (average) test error, can we verify our MSE decomposition? 

Recall
$$\begin{align*}
\E[\text{MSE}] &= \text{Bias}^2 + \text{Variance} + \text{Irreducible Error} \\
\text{ where } \text{Bias}^2 &= \E\left[\left(\E[\hat{y}] - \E[y]\right)^2\right] \\
&= \left(\E[\hat{y}] - \E[y]\right)^2 \text{ (Why can I drop the outer expectation?)} \\
\text{Variance} &= \E\left[\left(\hat{y} - \E[\hat{y}]\right)^2\right] \\
\text{Irreducible Error} &= \E\left[\left(y - \E[y]\right)^2\right]
\end{align*}$$

(Make sure you understand these definitions and how they work together!)

Let's work these out using all the tools we built before. 

First, for the bias: 
- we already have $\E[y]$ - this is just the `yfun` we selected
- we can compute $\E[\hat{y}]$ by running KNN many times and averaging the result

```{r}
sample_grid <- matrix(seq(0, 1, length.out=101), ncol=1)
KNN_AVERAGE_PRED_K1 <- rowMeans(replicate(500, {
    X_train <- matrix(runif(25, 0, 1), ncol=1)
    y_train <- matrix(rnorm(25, mean=yfun(X_train), sd=0.5), ncol=1)
    knn.reg(train=X_train, y=y_train, k=1, test=sample_grid)$pred
}))

KNN_BIAS_K1 <- KNN_AVERAGE_PRED_K1 - yfun(sample_grid)
plot(sample_grid, KNN_BIAS_K1^2, col="red4", 
     type="l", main="Squared Bias of KNN with K=1")
```

Not too much bias, but things do go a bit off the rails near the 
end points.

Next, we can compute variance pointwise:

```{r}
sample_grid <- matrix(seq(0, 1, length.out=101), ncol=1)
KNN_VARIANCE_K1 <- rowMeans(replicate(500, {
    X_train <- matrix(runif(25, 0, 1), ncol=1)
    y_train <- matrix(rnorm(25, mean=yfun(X_train), sd=0.5), ncol=1)
    (knn.reg(train=X_train, y=y_train, k=1, test=sample_grid)$pred - KNN_AVERAGE_PRED_K1)^2
}))

plot(sample_grid, KNN_VARIANCE_K1, col="red4", 
     type="l", main="Variance of KNN with K=1")
```

For this data at least, the variance term is generally larger than the
bias term: this is what we expect with a very flexible (high variance
+ low bias) model like $1$-NN.

Finally, irreducible error is just 0.25 everywhere (recall 
$y \sim \mathcal{N}(\E[y], 0.25)$). 

```{r}
KNN_IE <- rowMeans(replicate(500, {
    sample_grid <- matrix(seq(0, 1, length.out=101), ncol=1)
    X_test <- matrix(sample_grid, ncol=1)
    y_test <- matrix(rnorm(sample_grid, mean=yfun(X_test), sd=0.5), ncol=1)
    y_best_pred <- matrix(yfun(X_test), ncol=1)
    (as.vector(y_best_pred - y_test))^2
}))

plot(KNN_IE)
```

Put these together and we see the decomposition in action:

```{r}
mean(KNN_BIAS_K1)^2 + mean(KNN_VARIANCE_K1) + mean(KNN_IE)
```

as compared to

```{r}
mean(KNN_K1_MSE)
```

Visually, 

```{r}
#| message: false
library(dplyr)
library(tidyverse)
DECOMP_DATA <- data.frame(
   sample_grid = sample_grid, 
   KNN_K1_BIAS2 = KNN_BIAS_K1^2, 
   KNN_K1_VARIANCE=KNN_VARIANCE_K1, 
   KNN_K1_IE = KNN_IE, 
   KNN_K1_MSE = KNN_K1_MSE) |> 
   pivot_longer(-sample_grid) |>
   mutate(Error=value,
          Type=case_when(
            name=="KNN_K1_BIAS2" ~ "Bias^2", 
            name=="KNN_K1_IE" ~ "Irreducible Error", 
            name=="KNN_K1_VARIANCE" ~ "Variance", 
            name=="KNN_K1_MSE" ~ "Total Error")) 


ggplot() + 
  geom_bar(data=DECOMP_DATA |> filter(Type != "Total Error"), 
           mapping=aes(x=sample_grid, y=Error, color=Type), 
           stat="identity") + 
  geom_line(data=DECOMP_DATA |> filter(Type == "Total Error"), 
            mapping=aes(x=sample_grid, y=Error), 
            color="red4", linewidth=2) + 
  xlab("X") + ylab("Test Error")
```

So a bit of weirdness at the left end point - but holds up as well 
as we might expect for $N=25$ samples. 

How do the relative magnitudes of these terms change as you 
adjust the parameters of the simulation?

## Introduction to Convex Optimization

In this course, we will frequently want to find the minimizer of
certain functions. Typically, these will arise as ERMs and the function
will take a $p$-vector of inputs, *e.g.* regression coefficients, and 
produce a scalar loss output. Let us generally call this function
$f: \R^p \to \R$.  

In some circumstances, we can take the derivative of $f$, typically
called the _gradient_ in this context, and set it equal to zero. For
example, suppose we want to minimize an expression of the form: 

$$f(\bx) = \frac{1}{2}\bx^{\top}\bA\bx + \bb^{\top}\bx + c$$

where $\bA$ is a _symmetric_ strictly positive-definite matrix, 
$\bb$ is an arbitrary $p$-vector, and $c$ is a constant. Taking the
gradient, we find

$$ \frac{\partial f}{\partial \bx} = \bA\bx + \bb$$

We set this to zero and find a crticial point at: 

$$\begin{align*}
\mathbf{0} &= \bA\bx + \bb \\
- \bb &= \bA\bx \\
\implies \bx &= -\bA^{-1}\bb
\end{align*}$$

assuming that $\bA$ is invertible. (Here, invertibility is implied
by assuming $\bA$ is **strictly** positive-definite.) As usual, we are
not quite done here, as we must also check that this is a
_minimizer_ and not a maximizer or a saddle point. To do so, we
take the *second* derivative to find 

$$ \frac{\partial}{\partial \bx}\frac{\partial f}{\partial \bx} = \bA$$

In this multivariate context, we need the second derivative, 
*a.k.a* the Hessian, to be _strictly positive-definite_ to guarantee
that we have found a minimizer and this is indeed exactly what we
assumed above. As discussed above, 'definiteness' plays the role of
sign for many applications of matrix-ness. Here, by assuming strictly
positive definite, we are essentially treating the matrix as
strictly positive ($>0$), which is exactly the condition we need to 
guarantee a minimizer in the scalar case as well. 

Hence, we have that the one minimizer of $f$ is found at 
$\bx_* = \bA^{-1}\bb$. Compare this to the scalar case of minimizing
a quadratic $\frac{1}{2}ax^2 + bx + c$ with minimizer at $x = -b/a$.[^spd]

[^spd]: Note that we can only minimize this quadratic in the case where $a$ is
strictly positive so the parabola is upward facing. This is the scalar
equivalent of the strict positive-definiteness condition we put on $\bA$. 

The above analysis works well, but it is essentially the only 
minimization we will be able to do in closed form in this course.[^ols]

For other functions, we will need to apply _optimization_: the
mathematical toolkit for finding minimizers (or maximizers) of functions.
Fortunately for us, many of the methods we begin this course with
fall in the realm of _convex optimization_, a particularly nice branch
of optimization. 

*Convex* Optimization refers to the problem of *minimizing* *convex* functions
over *convex* sets. Let's define both of these: 

- A *convex* function is one which satisfies this inequality at all points: 
  $$f(\lambda \bx + (1-\lambda)\by) \leq \lambda f(\bx) + (1-\lambda) f(\by)$$
  
  for any $\bx, \by$ and any $\lambda \in [0, 1]$. 
  
  This definition is a bit non-intuitive, but it basically implies that we have
  a "bowl-like" function. This definition captures the idea that the actual
  function value is always less than we might get from linear interpolation. 
  A picture is helpful here: 
 
```{r}
#| echo: false
x  <- seq(-2, 3, length.out=201)
f  <- function(x) x^4 / 5 + 3
fx <- f(x)

x0  <- -1
fx0 <- f(x0)

x1  <- 2.5
fx1 <- f(x1)

lambda <- 0.75
xl  <- lambda * x0 + (1-lambda) * x1
fxl <- f(xl)

plot(x, fx, type="l", 
     main="A Convex Function", 
     xlab="x", 
     ylab=expression(f(x) == x^4))

points(c(x0, x1), c(fx0, fx1), cex=2, pch=16, col="red4")
segments(x0=x0, y0=fx0, x1=x1, y1=fx1, col="red4", lty=2)

points(xl, fxl, col="blue4", cex=2, pch=16)
```
  Here, we see that the actual value of the function (blue dot) is less 
  than what we would get if we interpolated the two red points (red line). 
  
  An alternative definition is that if $f$ is twice-differentiable, its Hessian
  (2nd derivative matrix) is positive semi-definite. 
  
- A *convex* set is one that allows us to look "between" points. 
  Specifically, a set $\mathcal{C} \subseteq \R^p$ is convex if 
  $$\bx, \by \in \mathcal{C} \implies \lambda \bx + (1-\lambda)\by \in \mathcal{C}$$
  
  for all $\bx, \by \in \mathcal{C}$ and any $\lambda \in [0, 1]$. 
  
Clearly these are related by the idea of looking "between" two points: 

i) Convex functions guarantee that this will produce something lower than
   naive interpolation; and 
ii) Convex sets guarantee that the midpoint will be "allowable". 

To tie these together, note another alternative characterization of a *convex
function*: one whose epigraph (the area above the curve in the plot) is a 
convex set. 

For optimization purposes, these two properties imply a rather remarkable fact: 

If $\bx_0$ is a *local* minimum of $$\min_{\bx \in \mathcal{C}} f(\bx)$$, then it
is a *global* minimum. 

This is quite shocking: if we find a point where we can't improve by going in 
any direction, we are guaranteed to have found a global minimum and no point
could be better. (It is possible to have _multiple_ minimizers however: consider
$f(\bx) = 0$. Any choice of $\bx$ is _a_ global minimizer.) This lets us turn
the "global" search problem into a "local" one. 

Of course, this only helps us _if_ we can find a local minimizer of $f$. How
might we do so? Let's recall a basic calculus idea: the gradient of a function
is a vector that points in the direction of greatest increase (the "steepest"
uphill). So if we go in the opposite direction of the gradient, we actually
will go "downhill". In fact, this is basically all we need to start applying
*gradient descent*. 

**Gradient Descent**: Given a convex function $f$: 

- Initialize at (arbitrary) starting point $\bx_0$. 
- Initialize step counter $k=0$. 
- Repeat until convergence: 
  - Compute the gradient of $f$ at $\bx_k$: $\left.\nabla f\right.|_{\bx=\bx_k}$
  - Set $\bx_{k+1} = \bx_k - c \left.\nabla f\right.|_{\bx=\bx_k}$
  
Repeated infinitely many times, this will converge to a local, and hence global,
minimizer of $f$. There are **many** variants of this basic idea[^pt], mostly
related to selecting the optimal step size $c$, but this is the most important
algorithm in convex optimization and it is important to understand it deeply. 

[^pt]: For instance, of the [14 optimization algorithms](https://pytorch.org/docs/stable/optim.html#algorithms) included in
base `pytorch`, all but one are advanced versions of gradient descent. 
The exception is `LBFGS` which attempts to (approximately) use both the gradient
and the Hessian (second derivative); computing the Hessian is normally quite 
expensive, so `LBFGS` uses some clever tricks to _approximate_ the Hessian. 

We can apply it here to the 3D function 
$$f(\bx) = (x_1-2)^2 + (x_2-3)^2 + (x_3 - 4)^2.$$
Clearly, we can see that the minimizer has to be at $(2, 3, 4)$, but our
algorithm won't use that fact. 

Before we can implement this algorithm, we need the gradient, which is given by

$$f(\bx) = \left\|\bx - (2,3,4)^{\top}\right\|_2^2 \implies \nabla f = 2[\bx - (2, 3, 4)^{\top}]$$

```{r}
x <- matrix(c(0, 0, 0), ncol=1) # We want to work exclusively with column vecs
converged <- FALSE
c <- 0.001 # Small step size
f <- function(x) sum((x - c(2, 3, 4))^2)
grad <- function(x) 2 * (x - matrix(c(2, 3, 4), ncol=1))

while(!converged){
    g <- grad(x)
    x_new <- x - c * g
    
    if(sum(abs(x - x_new)) < 1e-5){
        converged <- TRUE
    }
    
    x <- x_new
}

x
```

We don't get the _exact_ minimizer, but we certainly get something 'close enough'
for our purposes. In [Report #01](../reports/repot01.html), you will use
gradient descent and some variants to study the least squares problem. 

Often, we will want to see how the value of $f(\bx_k)$ changes over the course
of the optimization. We expect that it will go down monotonically, but it may
not be worth continuing the optimization if we have reached a point of 
'diminishing returns.' You can do this by hand (evaluating $f$ after each update
and storing the results), but many optimizers will often track this automatically
for you: *e.g.* [TensorBoard](https://www.tensorflow.org/tensorboard).[^tools]

[^ols]: OLS and some basic variants (ridge regression) can be written
in this 'quadratic' style. Can you see why? 

[^tools]: Modern ML toolkits like `pytorch` or `TensorFlow` are (at heart) fancy
systems to do two things automatically that we did 'by hand' in this example: 

    - Compute gradients via a process known as 'back-propogation' or 'autodiff'
    - Implement (fancy) gradient descent

