---
session: "`r session <- 8; session`"
topic: "`r library(tidyverse); topic <- read_csv('../key_dates.csv', name_repair='universal') |> filter(Course.Element == 'Class Session', Item.Number==session) |> pull(Details); topic`"
date: "Last Updated `r Sys.Date()`"
---

$$\newcommand{\R}{\mathbb{R}}\newcommand{\bx}{\mathbf{x}}\newcommand{\bw}{\mathbf{w}} \newcommand{\P}{\mathbb{P}} \newcommand{\V}{\mathbb{V}} \newcommand{\C}{\mathbb{C}} \newcommand{\E}{\mathbb{E}}$$

## Discriminative Classifiers

This week we begin to examine _discriminative_ classifiers. Unlike
generative classifiers, which attempt to model $\bx$ as a function
of $y$ and then use Bayes' rule to _invert_ that model, 
discriminative classifiers take the more standard approach of 
directly modeling $y$ as a function of $\bx$. This is the
approach you have already seen in our various variants of OLS. 

### Why Not Accuracy Maximization? 

As we saw in our regression unit, we can pose many interesting
estimators as _loss minimization_ problems. For least squares, we
primarily focused on mean squared error (MSE) as a loss function,
though we also briefly touched on MAE, MAPE, and 'check' losses.
There is not a single 'canonical' loss for the classification
setting and different choices of loss will yield different
classifiers. 

Before we build our actual first loss, it is worth asking
why we can't use something like classification accuracy as our loss
function. Specifically, define
$$\text{Acc}(y, \hat{y}) = \begin{cases} 1 & y = \text{sign}(\hat{y}) \\ 0 & y \neq \text{sign}(\hat{y}) \end{cases}$$
Here we consider the case where $\hat{y}$ may be real-valued 
as a (slight) generalization of probabilistic classification and
we use the $\pm 1$ convention for the two classes. 

Fixing $y = 1$, $\text{Acc}$ has the following shape: 

```{r}
y_hat <- seq(-3, 3, length.out=201)
acc <- function(y_hat, y=1) sign(y_hat) == sign(y)

plot(y_hat, acc(y_hat), type="l",
     xlab=expression(hat(y)),
     ylab=expression(Acc(hat(y), y==1)), 
     main="Accuracy as a Loss Function")
```
This is a terrible loss function! It is both non-convex (can you see
why?) and just generally unhelpful. Because it is flat almost
everywhere, we have no access to gradient information (useful for
fitting) or to any sense of 'sensitivity' to our loss. In this
scenario, if $y=1$, both $\hat{y}=0.00001$ and $\hat{y}=0.9999$
have the same accuracy even though the latter is a more 'confident'
prediction. 

### OLS for Classification?

So if not accuracy, where else might we get a good loss function?
You might first ask if we an use OLS? It works perfectly well for 
regression and we know it is well-behaved, so what happens if we
try it for classification? The answer is ... complicated. 

OLS as a _loss_ function for classification is a bit strange, but
it turns out to be more-or-less fine. In particular, OLS can be
used to evaluate _probabilistic_ classifiers, where it is known
as the _Brier score_. OLS as a _predictor_ can be a bit more 
problematic: in particular, if we just predict with 
$\P(y=1) = \bx^{\top}\bbeta$, we have to deal with the fact that
$\hat{y}$ can be _far_ outside a $[0, 1]$ range we might want
from a probabilistic classifier. In certain limited circumstances, 
we can assume away this problem (perhaps by putting specific
bounds on $\bx$), but these fixes are fragile. Alternatively, 
we can try to 'patch' this approach and use a classifier
like 

$$\P(y=1) = \begin{cases} 1 & \bx^{\top}\bbeta > 1 \\ 0 & \bx^{\top}\bbeta < 0 \\ \bx^{\top}{\bbeta} & \text{ otherwise}\end{cases}$$

Even if this feels a bit unsophisticated, ii is not awful, 
but it still struggles from the 'long flat region' problems that
accuracy encounters. 

At this point, it's hopefully clear that it will be a bit hard to
'hack together' a suitable loss and that we might benefit from
approaching the problem with more theoretical grounding. 

## Deriving Logistic Regression

While we argued that OLS is well-grounded in MSE alone, we can
recall it has additional connections to the Gaussian. These
"Gaussian vibes" were not essential to making OLS work, but 
they were useful in _deriving_ OLS as a sensible procedure. 
Can we do something similar for classification? That is, can we 
assume a 'working model' to come up with a good loss function
and then use it, even if we doubt that the model is 'correct'? 

As with all leading questions, the answer is of course a resounding
yes. We start from the rather banal observation that $y$ must have
a Bernoulli distribution conditional on $\bx$. After all, the
Bernoulli is (essentially) the only $\{0, 1\}$ distribution we
have to work with. So we really just need a way to model
the $p$ parameter of a Bernoulli as a function of $\bx$. Because
we love linear models, we may choose to set $p = \bx^{\top}\bbeta$,
but this gets us back to the range problem we had above. 

Specifically, we require the Bernoulli parameter to take values in
$[0, 1]$ but our linear predictor $\bx^{\top}\bbeta$ takes values
in all of $\R$. We can fix this with a simple 'hack': we need a 
function that 'connects' $\R$ to $[0, 1]$. If we call this function
$\mu$, we then can specify our whole model as $$y | \bx \sim \text{Bernoulli}(\mu(\bx^{\top}\bbeta))$$ and reduce our problem
to estimating $\bbeta$. Where can we get such a function? 

The most common choice is $$\mu(z) = \frac{e^z}{1+e^z} = \frac{1}{1+e^{-z}}$$, which is known as the _logistic_ or
_sigmoid_ function due to its 's-like' shape: 

```{r}
z <- seq(-5, 5, length.out=201)
plot(z, 1/(1+exp(-z)), type="l", main="Sigmoid Mapping")
```
More generally, we can use essentially the CDF of any random
variable supported on the real line as a choice of $\mu$: since
CDFs map the support onto the $[0, 1]$ range they are perfect 
for this choice. Unfortunately, most CDFs are a bit unwieldy
so this approach, while theoretically satisfying, is not too widely
used in practice. If you see it, the most common choice is 
$\mu(z) = \Phi(z)$, the normal CDF, resulting in a method known as
"probit" regression (after 'probability integral transform,' an
old-fashioned name for the CDF) or 
$$\mu(z) = \frac{\tan^{-1}(z)}{\pi} + \frac{1}{2}$$
which gives a method known as "cauchit" regression, since
this is the CDF of the standard Cauchy distribution. These
choices are far less common than the default sigmoid we used
above. Generally, they are only _slightly_ different in practice
and far more computationally burdensome and just aren't really worth
it. 

Returning to our default sigmoid, we now have the model 
$$ y | \bx \sim \text{Bernoulli}\left(\frac{1}{1+e^{-\bx^{\top}\bbeta}}\right)$$
This model is well-posed (in the sense that the distribution 'fits'
the data and we are guaranteed never to put invalid parameters in)
but we don't yet have a way to use it as a loss function. 

### From Model to Loss Function

We can build a loss-function by relying on the _maximum likelihood_
principle. The maximum likelihood principle is a core idea of
statistics - and one you will explore in much greater detail in other
courses - but, in essence, it posits that we should use the
(negative) PMF or PDF as our loss function. Specifically, the ML
principle says that, if many models could fit our data, we should
pick the one that makes our data _most probable_ (as determined
by the PMF/PDF). 

Before we work out the ML estimator (MLE) for our classification 
model,  let's take a brief detour and look at the MLE for 
(Gaussian) regression. Specifically, if we assume a model 
$$y \sim \mathcal{N}(\bx^{\top}\bbeta, \sigma^2)$$ for known 
$$\sigma^2$$, the PDF of the training point $(\bx_1, y_1)$ is

$$p(y_1 | \bx_1) = \frac{1}{\sqrt{2\pi\sigma^2}}\exp\left\{-\frac{\|y_1 - \bx_1^{\top}\bbeta\|_2^2}{2\sigma^2}\right\}$$

If we have a set of $n$ IID training data points, the joint PDF can
be obtained by multiplying together PDFs (IID is great!) to get

$$\begin{align*}
p(\mathcal{D}_{\text{train}}) &= \prod_{i=1}^n p(y_i | \bx_i) \\
&= \prod_{i=1}^n \frac{1}{\sqrt{2\pi\sigma^2}}\exp\left\{-\frac{\|y_i - \bx_i^{\top}\bbeta\|_2^2}{2\sigma^2}\right\} \\
&= (2\pi\sigma^2)^{-n/2}\exp\left\{-\frac{1}{2\sigma^2}\sum_{i=1}^n \|y_i - \bx_i^{\top}\bbeta\|_2^2\right\} \\
&= (2\pi\sigma^2)^{-n/2}\exp\left\{-\frac{1}{2\sigma^2} \|\by - \bX\bbeta\|_2^2\right\} \\
\end{align*}$$

We could maximize this, but a few minor tweaks make the problem
much easier: 

i) Since all of the action is in an exponential, we might as well
   maximize the _log_ of the PDF instead of the raw PDF. Since 
   logarithms are a monotonic transformation, this won't change
   our minimizer. Similarly, we will also strategically drop
   some constant terms as they also do not change the minimizer. 
ii) By convention, we like to _minimize_ more than _maximize_, 
    specifically when dealing with convex approaches, so we 
    will introduce a sign flip.
    
$$\begin{align*}
\hat{\bbeta} &= \argmax_{\bbeta} p(\mathcal{D}_{\text{train}}) \\
             &= \argmax_{\bbeta} \log p(\mathcal{D}_{\text{train}}) \\
             &= \argmax_{\bbeta} \log\left( (2\pi\sigma^2)^{-n/2}\exp\left\{-\frac{1}{2\sigma^2}\sum_{i=1}^n \|\by - \bX\bbeta\|_2^2\right\}\right) \\
             &= \argmax_{\bbeta} -\frac{n}{2}\log(2\pi\sigma^2) -\frac{1}{2\sigma^2} \|\by - \bX\bbeta\|_2^2 \\
             &= \argmin_{\bbeta} \frac{n}{2}\log(2\pi\sigma^2) + \frac{1}{2\sigma^2} \|\by - \bX\bbeta\|_2^2 \\
             &= \argmin_{\bbeta} \frac{1}{2}\|\by - \bX\bbeta\|_2^2
\end{align*}$$

Why were we able to do the simplifications in the last line? (You
might also recognize the _penultimate_ line as a term used in the AIC
of a linear model.)

This is quite cool! If we assume $\by$ follows a particular 
Gaussian, we get OLS by applying the general MLE approach. Deeper
statistical theory tells us that the MLE is (essentially) optimal,
so if our data is not too 'un-Gaussian' it makes sense that the 
Gaussian MLE (OLS) will perform reasonably well. Since many things
in this world are Gaussian-ish, OLS is optimal-ish for a large
class of interesting problems. 

Returning to our classification problem, we see that if our 
training data is IID, the MLE can be obtained by minimizing
the negative sum of the log PDFs. For space, we often start from 
this point instead of doing the full set of manipulations from scratch.

So what is the Bernoulli log PMF? Recall that if $B \sim \text{Bernoulli}(p)$, its PMF is given by: 

$$\P(B=k) = \begin{cases} p & k=1 \\ 1-p & k = 0 \end{cases}$$

or more compactly, 

$$\P(B = k) = p^k(1-p)^(1-k)$$

Taking logarithms, we have

$$\log \P(B = k) = k \log p + (1-k) \log(1-p)$$

and hence the negative log-likelihood: 

$$-\log \P(B = k) = -k \log p -(1-k) \log(1-p)$$


From our working model, we take $p = 1/(1+e^{-\bx^{\top}\bbeta})$ so our joint
MLE is given by: 

$$\hat{\bbeta} = \argmin_{\bbeta} \sum_{i=1}^n -y_i \log\left(1/(1+e^{-\bx_i^{\top}\bbeta})\right) - (1-y_i)\log\left(1 - 1/(1+e^{-\bx_i^{\top}\bbeta})\right)$$

This is still quite hairy, so let's simplify it. In the first term, we note
that $-\log(1/x) = \log x$ to get: 

$$\hat{\bbeta} = \argmin_{\bbeta} \sum_{i=1}^n y_i \log\left(1+e^{-\bx_i^{\top}\bbeta}\right) - (1-y_i)\log\left(1 - 1/(1+e^{-\bx_i^{\top}\bbeta})\right)$$

For the second term, note that

$$1-\frac{1}{1+e^{-z}} = \frac{1+e^{-z} - 1}{1+e^{-z}} = \frac{e^{-z}}{1+e^{-z}} \implies \log\left(1-\frac{1}{1+e^{-z}}\right) = -z - \log(1 + e^{-z})$$

so we get: 

$$\begin{align*}
\hat{\bbeta} &=  \argmin_{\bbeta} \sum_{i=1}^n y_i \log\left(1+e^{-\bx_i^{\top}\bbeta}\right) - (1-y_i)\log\left(1 - 1/(1+e^{-\bx_i^{\top}\bbeta})\right) \\
&= \argmin_{\bbeta} \sum_{i=1}^n y_i \log\left(1+e^{-\bx_i^{\top}\bbeta}\right) - (1-y_i)\left[-\bx_i^{\top}\bbeta + \log\left(1+e^{-\bx_i^{\top}\bbeta}\right)\right] \\
&= \argmin_{\bbeta} \sum_{i=1}^n y_i \log\left(1+e^{-\bx_i^{\top}\bbeta}\right) + (1-y_i)\left[\bx_i^{\top}\bbeta + \log\left(1+e^{-\bx_i^{\top}\bbeta}\right)\right] \\
&= \argmin_{\bbeta} \sum_{i=1}^n y_i \log\left(1+e^{-\bx_i^{\top}\bbeta}\right) + \left[\bx_i^{\top}\bbeta + \log\left(1+e^{-\bx_i^{\top}\bbeta}\right)\right] -y_i\left[\bx_i^{\top}\bbeta + \log\left(1+e^{-\bx_i^{\top}\bbeta}\right)\right]\\
&= \argmin_{\bbeta} \sum_{i=1}^n-y_i \bx_i^{\top}\bbeta + \bx_i^{\top}\bbeta + \log\left(1-e^{-\bx_i^{\top}\bbeta}\right)
\end{align*}$$

We could work in this form, but it turns out to actually be a bit nicer to 
'invert' some of the work we did above to get rid of a term. In particular, 
note

$$z + \log(1-e^{-z}) = \log(e^z) + \log(1-e^{-z}) = \log(e^z + e^{z-z}) = \log(1 + e^z)$$

which gives us: 

$$\hat{\bbeta} = \argmin_{\bbeta} \sum_{i=1}^n-y_i \bx_i^{\top}\bbeta  + \log\left(1+e^{\bx_i^{\top}\bbeta}\right)$$

Wow! Long derivation. And we still can't actually solve this!

Unlike OLS, where we could obtain a closed-form solution, we *have to use* 
an iterative algorithm here. While we could use _gradient_ descent, we can
get to an answer far more rapidly if we use more advanced approaches. Since
this is not an optimization course, we'll instead use some software to solve
for us: 

```{r}
# Generate data from the logistic model
n <- 500
p <- 5
X    <- matrix(rnorm(n * p), ncol=p)
beta <- c(1, 2, 3, 0, 0)

P <- 1/(1+exp(-X %*% beta))
y <- rbinom(n, size=1, prob=P)

# Use optimization software
# See https://cvxr.rbind.io/cvxr_examples/cvxr_logistic-regression/ for details
library(CVXR)

beta <- Variable(p)
eta  <- X %*% beta

objective <- sum(- y * eta + logistic(eta))
problem   <- Problem(Minimize(objective))

beta_hat <- solve(problem)$getValue(beta)
```

```{r}
#| echo: false
knitr::kable(beta_hat, col.names="beta_hat")
```

Not too bad, if we compare this to `R`'s built-in logistic regression function, 
we see our results basically match: 

```{r}
summary(glm(y ~ X + 0, family=binomial))
```
Good match! 

Note that, in the above, we had to use the `logistic` function built into `CVX`.
If we used `log(1+exp(eta))`, the solver would not be able to prove that it is
convex and would refuse to try to solve. See the `CVX` documentation for more 
details.

We can compute accuracy after defining a 'decision rule'. Here, let's
just round the predicted probability

```{r}
p_hat <- 1/(1 + exp(-X %*% beta_hat))
y_hat <- round(p_hat)

mean(y_hat == y)
```

Now that we have this toolkit built up, we can also easily apply ridge and lasso
penalization: 

```{r}
penalty <- sum(abs(beta))

lasso_problem <- Problem(Minimize(objective + 9 * penalty))
beta_hat_lasso <- solve(lasso_problem)$getValue(beta)
```

```{r}
#| echo: false
knitr::kable(beta_hat_lasso)
```

We got some sparsity, as we would expect with $\ell_1$ penalization. 

### Generalized Linear Models

The approach we took here is a special case of a _generalized linear model_: 
generalized linear models (GLMs) consist of three parts: 

- A linear model for the 'core predictor' $\eta = \bX\bbeta$; 
- A sampling distribution for the data $y \sim \text{Dist}$; and
- An 'adaptor' function[^link], $\mu$, that maps $\eta \in \R$ to the parameter 
  space of $y$: specifically, we need $\E[y | \bx] = \mu(\bx^{\top}\bbeta)$.
  
The adaptor function needs to be smooth (for nice optimization properties) and
monotonic (or else the interpretation gets too weird), but we otherwise have some
flexibility. As noted above, if we pick $\mu$ to be the normal or Cauchy CDFs, we
get alternative 'Bernoulli Regressions' (probit and cauchit). 

We can generalize the model for $\eta$ by allowing it to be an additive (spline)
model or a kernel method. To fit splines, you can use the `gam` function
from the `mgcv` function. 

Finally, we also have choices in the sampling distribution. While normal and
Bernoulli are the most common, you will also sometimes see Poisson and Gamma
in the wild. For both of these, the mean is a positive value, so we require
an adaptor that maps $\R$ to $\R \geq 0$ and we typically take $\mu(z) = e^z$. 

A good exercise is to repeat the above analysis for Poisson regression and compare
your result (obtained with `CVXR`) to the Poisson regression built into `R`. 

## Loss Functions for Classification

We now have a workable loss function for discriminative classification.
Returning to our example from above 
(changed to $\{0, 1\}$ convention briefly). Let us set 
$\alpha = y\hat{y}$ and investigate our loss functions as a property
of $\alpha$: 

```{r}
alpha <- seq(-3, 3, length.out=201)
acc     <- function(alpha) ifelse(alpha < 0, 1, 0)
lr_loss <- function(alpha) log(1 + exp(-2*alpha))

plot(alpha, acc(alpha), type="l",
     xlab=expression(alpha),
     ylab="Loss", 
     main="Classification Losses", 
     ylim=c(0, 5))
lines(alpha, lr_loss(alpha), col="red4")
legend("topright", 
       col=c("black", "red4"), 
       lwd=2, 
       legend=c("Accuracy", "Logistic Regression"))
```

We see here that the logistic loss defines a _convex surrogate_ of the 
underlying accuracy. Just like we used $\ell_1$ as a tractable
approximation for best squares, logistic regression can be considered
a tractable approximation for accuracy minimization.[^persp]

Other useful loss functions can be created using this perspective, 
perhaps none less important than the *hinge loss*, which gives rise
to a classifier known as the *support vector machine* (SVM). 

### Support Vector Machines

In this section, we go back to $\{\pm 1\}$ convention. 

Instead of using a logistic loss, consider a hinge loss of the form

$$H(y, \hat{y}) = (1 - y\hat{y})_+$$

When $y = \hat{y}=1$ or $y = \hat{y} = 0$, this is clearly 0 as we
would expect from a good loss. What happens for cases where our 
prediction is wrong or not 'full force', say $\hat{y} = 1/2$. 

Looking ahead, we will use a linear combination of features to create
$\hat{y} = \bx^{\top}\bbeta$ yielding:

$$H(\bbeta) = (1 - y \bx^{\top}\bbeta)_+$$

If $y=1$, we get zero loss so long as $\bx^{\top}\bbeta > 1$ and a 
small loss for $0 < \bx^{\top}\bbeta < 1$. As $\bx^{\top}\bbeta$
crosses zero and goes negative, the loss grows linearly without
bound. (Reverse all of this for the case $y = -1$.) This is an
interesting loss function: we cannot make our loss decrease
as our predictions become 'more right', but our loss continues
to increase as our prediction becomes 'more wrong'.

Visually, we can draw this on our plot from above: 

```{r}
alpha <- seq(-3, 3, length.out=201)
acc     <- function(alpha) ifelse(alpha < 0, 1, 0)
lr_loss <- function(alpha) log(1 + exp(-2*alpha))
hinge_loss <- function(alpha) pmax(0, 1-alpha)

plot(alpha, acc(alpha), type="l",
     xlab=expression(alpha),
     ylab="Loss", 
     main="Classification Losses", 
     ylim=c(0, 5))
lines(alpha, lr_loss(alpha), col="red4")
lines(alpha, hinge_loss(alpha), col="green4")
legend("topright", 
       col=c("black", "red4", "green4"), 
       lwd=2, 
       legend=c("Accuracy", "Logistic Regression", "Hinge Loss"))
```

This is also a pretty nice surrogate loss. In this case, let's just
go 'straight at it' to construct a classifier: 

$$\hat{\bbeta} = \argmin \sum_{i=1}^n (1-y_i \bx_i^{\top}\bbeta)_+$$

This classifier is not uniquely defined for many problems (because of
the long flat part of the loss), so it is conventional to add a 
_regularization_ term to the SVM: 

$$\hat{\bbeta} = \argmin \sum_{i=1}^n (1-y_i \bx_i^{\top}\bbeta)_+ + \lambda \|\bbeta\|_2^2$$

We can solve this directly using CVX as follows: 

```{r}
# Generate data from the logistic model and fit an SVM
n <- 500
p <- 5
X    <- matrix(rnorm(n * p), ncol=p)
beta <- c(1, 2, 3, 0, 0)

P <- 1/(1+exp(-X %*% beta))
y <- rbinom(n, size=1, prob=P)
## Convert to +/-1 convention
y <- 2 * y - 1

library(CVXR)

beta <- Variable(p)
eta  <- X %*% beta

objective <- sum(pos(1 - y * eta)) + 0.5 * norm2(beta)^2
problem   <- Problem(Minimize(objective))

beta_hat <- solve(problem)$getValue(beta)
```

Using the decision rule $\hat{y} = \text{sign}(\bx^{\top}\bbeta)$, this
gives us good accuracy: 

```{r}
mean(sign(X %*% beta_hat) == y)
```

This is comparable to what we got for logistic regression even
though: 

i) the loss function is 'wrong' (doesn't model the DGP)
ii) we did no tuning of the ridge parameter

So, all in all, pretty impressive. 

SVMs are a remarkably powerful and general classification technology.
Before we move past them, we will now take a second perspective
focusing on the _geometry_ of SVMs. 


  
[^link]: For historical reasons, this is known as the _inverse link_, but I prefer
to simply think of it as an adaptor and to never use the (forward) link. 

[^persp]: This 'surrogate loss' perspective was first discussed in: 
P.L. Barlett, M.I. Jordan, and J.D. McAuliffe. "Convexity, Classification, and Risk Bounds". *Journal of the American Statistical
Association* **101(473)**, pp.138-156. 2006. DOI:[10.1198/016214505000000907](https://dx.doi.org/10.1198/016214505000000907)

### SVM: The Geometric Viewpoint

```{r}
#| include: false
TRUE # To ensure knitr activates
```


## Non-Linear Regression

We have now developed a variety of tools for linear regression and
for linear classification. These work well - particularly in
low SNR (high dimension / low sample size) scenarios - but they
have a necessarily limited capacity. When we have lots of training
data, how can we go further? By using _non-linear_ methods of course!

In this section we introduce several methods for non-linear regression. 
We also pause to consider _when_ non-linear methods can actually be
expected to outperform linear methods. 

Also note that we are arguably being a bit 'sloppy' about 
what we mean by 'linear' methods. For purposes of these notes,
we define a 'linear' method as one where the fitted model has a
linear response to changes in the inputs; formally, one where

$$\frac{\partial y}{\partial x_i} = \beta_i \text{ (a constant) for all } i \in \{1, \dots, p\}$$

This definition includes things like best-subsets and lasso
regression, which are not linear functions of $\by$, but
excludes things like polynomial regression.[^poly] In the
above definition, each $\beta_i$ is essentially a regression
coefficient, and values of $i$ where $\beta_i = 0$ are
non-selected variables if we are using a sparse method.

[^poly]: In other classes, you may have been taught that
polynomial regression is a linear model. It is definitely linear
in the sense that that class would have used (linear in $\by$)
but we're using a different sense here. The core insight of 
that analysis - that a non-linear function can be expressed as
a linear combination of non-linear parts - will appear several
times in these notes.

## Non-Linearity via Linear Combination of Non-Linear Parts

We have spent the previous few weeks developing a set of tools
for building _linear_ models of $\by$. Specifically, we have
sought ways to approximate $\E[y | \bx]$ as a weighted sum
of each $x_i$: 

$$ \E[y | \bx] \approx \langle \bx, \hat{\bbeta} \rangle = \sum_{i=1}^p x_i \hat{\beta}_i$$

For situations where the relationship is nearly (or even truly)
linear, this approximation can be quite useful and good
estimates of the $\beta_i$ coefficients can lead to accurate
estimates of $\E[y | \bx]$, and ultimately accurate prediction 
of test set response $\by$. Using the bias-variance
decomposition, we have seen that it is often worthwhile to accept
a bit of bias in estimation of each $\beta_i$ if it is compensated by
a worthwhile reduction in the variance of $\beta_i$. 

We can further refine this equality by decomposing bias a bit further: 

$$\text{MSE} = \text{Irreducible Error} + \text{Variance} + \text{Model Bias}^2 + \text{Estimation Bias}^2$$

Here, we have decomposed $\text{Bias}^2$ into two terms: 

- $\text{Model Bias}^2$: a measure of the _systematic_ error arising
  from use of a linear model to predict a non-linear DGP
- $\text{Estimation Bias}^2$: a measure of the _systematic_ error arising
  from use of a regularized estimation procedure which exhibits shrinkage
  
Put another way, "Model Bias" results from use of a linear model when
something non-linear should be used, while "Estimation Bias" arises from
the use of a biased method (like ridge regression) instead of something
nominally unbiased like OLS. For the previous two weeks, we have mainly
focused on linear models for linear DGPs, so model bias has been zero. If
we expand our gaze to non-linear DGPs, we have to deal with model bias
a bit more directly. As with estimation bias, it is frequently worthwhile
to accept model bias to reduce variance; see more discussion below. 

### Polynomial Expansion

You have likely already seen polynomial expansion (or polynomial regression)
in previous courses. Essentially, PE fits a low(-ish)-order polynomial
instead of a line to data. More formally, let $f(\bx) \equiv \E[y|\bx]$
be the regression function (best possible predictor) that week seek to
approximate from noisy observations. Standard linear regression essentially
seeks to fit a first-order Taylor approximation of $f$ around some point $\overline{\bx}$:

$$\begin{align*}
f(\bx) &\approx f(\overline{\bx}) + \nabla_f(\bx_0)^{\top}(\bx - \overline{\bx}) \\
       &= f(\overline{\bx}) + \sum_{i=1}^p \left. \frac{\partial f}{\partial x_i}\right|_{x = \overline{x}_i}(x_i - \overline{x}_i) 
\end{align*}$$

If we rearrange this a bit, we get

$$\begin{align*}
f(\bx) &\approx f(\overline{\bx}) + \sum_{i=1}^p \left. \frac{\partial f}{\partial x_i}\right|_{x = \overline{x}_i}(x_i - \overline{x}_i) \\
&= \underbrace{\left(f(\overline{\bx}) - \sum_{i=1}^p \left.\frac{\partial f}{\partial x_i}\right|_{x_i=\overline{x}_i}\overline{x}_i\right)}_{=\beta_0} + \sum_{i=1}^p \underbrace{\left.\frac{\partial f}{\partial x_i}\right|_{x_i=\overline{x}_i}}_{=\beta_i} x_i\\
&= \beta_0 + \sum_{i=1}^p x_i \beta_i
\end{align*}$$

so we see that the regression coefficients are more-or-less the
(suitably-averaged) partial derivatives of $f$ while the intercept
is the value of $f$ at the center of our expansion ($f(\overline{\bx})$)
plus some differential adjustments. Note that, if the variables $\bX$ are
centered so that $\E[\bX] = \mathbf{0}$, the intercept is exactly the
average value of $f$, as we would expect.

Clearly, this linear approximation will do better in the situations
where Taylor expansions generally perform better: when i) the derivative
of $f$ is roughly constant and ii) the higher order terms are small. This 
makes sense: if $f$ is very-close-to-linear, there is minimal loss from
fitting a linear approximation to $f$; if $f$ is very non-linear, however,
a linear approximation can only be be so good, and we are saddled with
significant model bias. 

If we want to mitigate some of that model bias, we can choose to use
a higher-order Taylor series. If we examine a second order Taylor series, 
we get a similar approximation as before, where $\mathcal{H}$ denotes the _Hessian_
matrix of second derivatives: 

$$\begin{align*}
f(\bx) &\approx f(\overline{\bx}) + \nabla_f(\bx_0)^{\top}(\bx - \overline{\bx}) + \frac{1}{2}(\bx - \overline{\bx})^{\top}\mathcal{H}_f(\overline{\bx})(\bx - \overline{\bx})^{\top}
\end{align*}$$

After some rearrangement this becomes: 

$$f(\bx) \approx \beta_0 + \sum_{i=1}^p \frac{\partial f}{\partial x_i} x_i +\sum_{i=1}^p \frac{\partial^2 f}{\partial x_i^2} x_i^2 +  \sum_{\substack{i, j = 1 \\i \neq j}}^p \frac{\partial^2 f}{\partial x_i \partial x_j} x_ix_j$$

We recognize the constant term (intercept) and first
order terms (linear slopes) from before, but now we
have added two types: 

- Second order terms ($x_i^2$): these capture higher
  order non-linearities in a single variable. For
  example, if $y = x^2$, then the 'true fit' really
  lies in the second order term 
  $\partial^2f / \partial x^2 = 2$ instead of any 
  linear approximation
- Cross terms ($x_ix_j$): these capture non-linear
  (non-additive) multi-variate relationships between
  features. You may have seen these before as
  _interaction_ terms. 
  
Rules of thumb vary as to which set of terms are more
important. Historically, statisticians have tended
to put in the interaction (cross) terms first, though
this is far from universal practice. In particular, 
when dealing with 'binary' variables, the higher
order term is unhelpful: if a feature is $0/1$ (did the
patient receive the drug or not?), adding a squared
term has no effect since $0^2 = 0$ and $1^2 = 1$. 

Regardless, this is why we sometimes use 'polynomial
expansion' of our original variables. We are trying
to capture the higher-order (non-linear) terms of the
Taylor approximation. 

It is worth thinking about what happens when we add
higher-order terms to our model. In particular, note
that we i) pick up some correlation among features; and
ii) we will generally have more variance since we have
more features. In practice, the correlation isn't too much
of a problem and the actual polynomials fit by, *e.g.*, `poly`
in `R` are modified to remove correlation. The variance however
can be a more significant problem. 

Let's see this in action. Let's first consider fitting polynomial regression to a simple (univariate) non-linear function:  $f(x) = \sqrt{|x|^3 + 5} * \cos(x)$

```{r}
curve(sqrt(abs(x)^3 + 5) * cos(x), from=1, to=10)
```

This is not *too* non-linear. It's still very smooth (*analytic* in 
the jargon) and should be relatively easy to fit. Let's also generate
some noisy observations of this function. 

```{r}
n <- 25
# Linear regression doesn't care about the order of our data, but plotting does
# so sorting gives better visuals
x <- sort(runif(n, 2, 10)) 
Ey <- sqrt(abs(x)^3 + 5) * cos(x)
y <- Ey + rnorm(n, sd=3)


x_grid <- seq(min(x), max(x), length.out=501)
Ey_grid <- sqrt(abs(x_grid)^3 + 5) * cos(x_grid)

plot(x, y)
lines(x_grid, Ey_grid)
```

We can now fit some linear models to this data (shown in color): we use
R's built-in `poly` function to automatically create the polynomials we
seek. You can think of `poly(x, k)` as creating `k` features $x^1, x^2,
\dots, x^k$ but it actually does something a bit more subtle under the
hood to make model fitting a bit more numerically stable.

```{r}
plot(x, y)
lines(x_grid, Ey_grid)

for(k in 1:5){
    m <- lm(y ~ poly(x, k))
    
    yhat <- predict(m, data.frame(x = x_grid))
    lines(x_grid, yhat, col=k+1)
}

legend("bottomleft",
    c("True Regression Function", 
         "First Order (Linear) Fit",
         "Second Order (Quadratic) Fit", 
         "Third Order (Cubic) Fit", 
         "Fourth Order (Quartic) Fit",
         "Fifth Order (Quintic) Fit"), 
       col=1:6, lty=1)
```

Clearly, as we increase the order of the polynomial, we get a better
(in-sample) fit. This makes sense: we know that more features gives us a
better fit all else being equal, so what's the issue? 

As usual, the issue occurs for extrapolation. You can actually see some
issues already beginning to manifest at the ends of our prediction
intervals: the higher-order polynomials go radically different directions
as we extrapolate and even the 'best fit' (quintic) looks like it's going
to be too steep as we go further. 

Let's expand the prediction region for these fits: 

```{r}
x_pred <- seq(-5, 15, length.out=501)
Ey_pred <- sqrt(abs(x_pred)^3 + 5) * cos(x_pred)
plot(x_pred, Ey_pred, type="l")
points(x, y)

for(k in 1:5){
    m <- lm(y ~ poly(x, k))
    yhat <- predict(m, data.frame(x=x_pred))
    lines(x_pred, yhat, col=k+1)
}

legend("bottomleft",
    c("True Regression Function", 
         "First Order (Linear) Fit",
         "Second Order (Quadratic) Fit", 
         "Third Order (Cubic) Fit", 
         "Fourth Order (Quartic) Fit",
         "Fifth Order (Quintic) Fit"), 
       col=1:6, lty=1)
```

There's some problems!

To be fair, it is a bit unreasonable to expect these models to perform too well outside of our original sampling area. What's more worrying is that the extrapolations don't just miss the curves of the true regression function, they actually create their own even more intense wiggles. 

We can also manifest this without extrapolation by fitting very high-order polynomials:

```{r}
#| warning: false
k <- 9
m <- lm(y ~ poly(x, degree=10 + k, raw=TRUE))
yhat <- predict(m, data.frame(x = x_grid))
```

```{r}
#| warning: false
plot(x, y)
lines(x_grid, Ey_grid)

for(k in 1:5){
    m <- lm(y ~ poly(x, degree=14 + k, raw=TRUE))
    yhat <- predict(m, data.frame(x = x_grid))
    lines(x_grid, yhat, col=k)
}
```

Note that we have to use `raw=TRUE` here to avoid `R` complaining about
having too-high a degree in the polynomial. (`R` gives an error about
"unique points" but the real issue is one about matrix rank, not ties in
`x`.) I'm also surpressing some warnings here about a sub-standard fit.

So how can we address this 'over-wiggliness'? 

As we discussed last time, we need to use a different set of functions: one that is smooth (like polynomials), but not too-high order. It would also be really nice if we could get something 'adaptive' - allowing for more wiggliness where we have more data (and we need more wiggliness) and less wiggliness where we don't have enough data (fall-back to linearity). 
### Local Linear Models

One way to do this is the idea of "local linear (polynomial) models."
Instead of fitting a single (global) line, we can fit different lines in
different parts of our data: that way, we can get an upward line when the
true (non-linear) relationship is increasing and a downward line when the
true relationship is decreasing. Or at least that's the hope!

```{r}
plot(x, y)
lines(x_grid, Ey_grid)

# Divide our data into five 'buckets' and fit sub-models
# This works because we've already sorted our data
# (What would happen if we hadn't?)
for(k in 1:5){
    xk <- x[(5 * k - 4):(5*k)]
    yk <- y[(5 * k - 4):(5*k)]
    mk <- lm(yk ~ xk)
    y_hat <- predict(mk)
    lines(xk, y_hat, col="red4")
}
```

Definitely some rough edges here - particularly in the areas between our buckets, but we're on a good path. The `locfit` package will help us with the details:

```{r}
library(locfit)

m <- (locfit(y ~ lp(x)))
summary(m)

plot(x, y)
lines(x_grid, Ey_grid)
y_hat <- predict(m, data.frame(x = x_grid))
lines(x_grid, y_hat, col="red4")
```

Not perfect, but pretty nice compared to what we had before. 

You can also see a discussion of the "degrees of freedom" in the model
output. This is not exactly the DoF you learned in earlier classes, but
it's sort of "morally equivalent." This local fit is about as flexible as
a 4th degree polynomial would be for this problem. Even though this model
is made out of quadratics, it's more flexible than a single quadratic.
But we avoid the extreme variability associated with a polynomial of that
high order. Win-win!

We can tweak some parameters of the local fit to get different responses:
the big ones are the degree (`deg`) and the number of neighbors to use. 

```{r}
#| eval: false
?lp
```

```{r}
m <- (locfit(y ~ lp(x, deg=4)))
summary(m)

plot(x, y)
lines(x_grid, Ey_grid)
y_hat <- predict(m, data.frame(x = x_grid))
lines(x_grid, y_hat, col="red4")
```

Pretty nice. If we 'turn down' the number of neighbors used: 

```{r}
m <- (locfit(y ~ lp(x, nn=0.2, deg=4)))
summary(m)

plot(x, y)
lines(x_grid, Ey_grid)
y_hat <- predict(m, data.frame(x = x_grid))
lines(x_grid, y_hat, col="red4")
```

Too far - super crazy! But not entirely unexpected - we know that
$K$-Nearest Neighbors for small $K$ has a huge variance. 

Too far!

The `loess` function in base `R` does this very nicely as well without 
requiring additional packages. For practical work, it's a very nice
tool for univariate modeling.[^pyloess]

[^pyloess]: In Python, the `statsmodels` package also implements [LOESS
fits](https://www.statsmodels.org/dev/examples/notebooks/generated/lowess.html).

```{r}
m <- loess(y ~ x)
summary(m)

plot(x, y)
lines(x_grid, Ey_grid)
y_hat <- predict(m, data.frame(x = x_grid))
lines(x_grid, y_hat, col="red4")
```

### Spline Regression (Additive Models)

### Splines

We can generalize this idea a bit using splines. Splines are functions that solve a penalized approximation problem: 

$$\hat{f} = \text{arg min}_{f} \frac{1}{2n} \|y - f(x)\|_2^2 + \lambda \int |f''(x)|^2 \text{d}x$$

Here we are saying we'll take _any function_ (not just a polynomial) 
that achieves the optimal trade-off between data fit (first / loss term)
and not being too rough (second / penalty term). 

Why is this the right penalty to use? In essence, we are trying to
penalize 'deviation from linearity' and since linear functions have
second derivative 0 (by definition) the integral of the second
derivative gives us a measure of non-linearity. 

In some remarkable work, [Grace Wahba](https://en.wikipedia.org/wiki/Grace_Wahba) 
and co-authors showed that the solutions to that optimization problem are
piecewise polynomials with a few additional constraints - these functions are
_splines_. In addition to piecewise polynomialness, splines also guarantee:

- continuity of the function and its derivatives
- linearity outside the range of data fit ('natural splines')

Because splines not only match the value of the function at the knots
(places where the 'pieces' match up) but also the derivatives, they are very
smooth indeed. 

Stepping back, splines are just a different sort of feature engineering.
Instead of using polynomial basis functions ($x^1, x^2, \dots, x^k$), splines
use a much smoother basis and hence give smoother results, avoiding the 'wiggles'
problem we saw above.

`R` provides the basic tools for spline fitting in the `splines` package. For
real work, you almost surely want to use the advanced functionality of the `mgcv`
package. For more on splines and models using them ('additive models'), see 
[this online course](https://noamross.github.io/gams-in-r-course/).

We can see that individual splines are quite nice little functions and usually
are only non-zero on small regions: 

```{r}
library(splines)

nsx <- ns(x_grid, df=5) # The natural spline basis
plot(x_grid, apply(nsx, 1, max), ylim=range(nsx), type="n")

for(i in 1:5){
    lines(x_grid, nsx[,i], col=i)
}
```

We can use linear combinations of these 'bumps' to fit non-linear functions, 
including our example from above: 

```{r}
m <- lm(y ~ ns(x, df=5))
plot(x, y)
lines(x_grid, Ey_grid)
y_hat <- predict(m, data.frame(x = x_grid))
lines(x_grid, y_hat, col="red4")
```

(This basically matches what we had before with $\text{DoF} \approx 4.7$, but
you can get different answers by messing with the `df` parameter  here.)

Our predicted response is actually the sum of the individual spline effects: 

```{r}
m <- lm(y ~ ns(x, df=5))
y_hat <- predict(m, data.frame(x = x_grid))

plot(x, y, lwd=2)
lines(x_grid, Ey_grid)
for(i in 1:5){
    lines(x_grid, ns(x_grid, df=5)[,i] * coef(m)[-1][i], col=i+1)
}
```

If we were able to sum up the colored lines, we would get the black line back. 

We can also see that natural splines give smooth predictions outside the range
of the original data: 

```{r}
y_pred <- predict(m, data.frame(x = x_pred))

plot(x_pred, Ey_pred, type="l")
lines(x_pred, y_pred, col="red4")
```

Admittedly, not ideal, but it will do. In particular, note that outside of the
'data region', we fall back on a nice comfortable 

(You can repeat this process with the slightly more common $b$-splines, but the differences aren't huge.)

```{r}
m <- lm(y ~ bs(x, df=5))
y_hat <- predict(m, data.frame(x = x_grid))

plot(x, y, xlim=range(x_grid))
lines(x_grid, Ey_grid)
for(i in 1:5){
    lines(x_grid, bs(x_grid, df=5)[,i] * coef(m)[-1][i], col=i+1)
}

y_pred <- predict(m, data.frame(x = x_grid))

lines(x_grid, y_pred, col="red4", lwd=2)
```

For a 'full-strength' version of spline fitting, you can use the `mgcv` package:

```{r}
library(mgcv)
m <- gam(y ~ s(x))

summary(m)
```


```{r}
m <- gam(y ~ s(x))
y_hat <- predict(m, data.frame(x = x_grid))

plot(x, y, xlim=range(x_grid))
lines(x_grid, Ey_grid)
lines(x_grid, y_hat, col="red4", lwd=2)
```

Run `?s` to see the _many_ features of the `mgcv` package. `mgcv` also includes
several specialized splines (*e.g.* periodic) which you may find useful in some
problems. 

Splines are not normally consider a 'machine learning' tool but they are
_incredibly_ useful in applied statistics work. (The two methods you will rely
on most from this class are spline regression and random forests.) 

### $\ell_1$-Filtering

## Kernel Methods

### Manual feature expansion

We can think of feature expansion as mapping our data to a higher-dimensional space $\Phi: \R^p \to \R^P$ and fitting a linear model there. As we have seen above,
splines and their kin provide a useful way of constructing the mapping $\Phi$, 
but we are still constrained to work with a fairly restricted type of problem. 

Can we generalize this formally to more interesting maps $\Phi(\cdot)$? 
Yes - via kernel methods!

### Ridge without coefficients

Before introducing kernel methods formally, let's look back to ridge regression.
We showed that the ridge solution is given by 

$$\hat{\beta}_{\text{Ridge}} = (\bX^T\bX + \lambda \bI)^{-1}(\bX^T\by)$$

Some nifty linear algebra will let us rewrite this as 

$$\hat{\beta}_{\text{Ridge}} = \lambda^{-1}\bX^T(\bX\bX^{\top}/\lambda + \bI)^{-1}\by = \bX^{\top}(\bX\bX^{\top}+\lambda \bI)^{-1}\by$$

You can prove this using the [Woodbury Matrix Identity](https://en.wikipedia.org/wiki/Woodbury_matrix_identity)
or we can just check it for a single data set and trust that it generalizes: 

```{r}
n <- 100
p <- 5 

X <- matrix(rnorm(n * p), nrow=n, ncol=p) 

Ey <- X[,1] + sqrt(X[,2]^2 + 5) + 0.1 * X[,3]^3 + cos(abs(X[,4])) + 1/(abs(X[,5]) + 3)
y <- Ey + rnorm(n, sd=sqrt(0.25))

lambda <- 2
eye <- function(p) diag(1, p, p)
beta_hat_ridge <- solve(crossprod(X) + lambda * eye(p), crossprod(X, y))

beta_hat_ridge_alt <- t(X) %*% solve(tcrossprod(X) + lambda * eye(n)) %*% y

cbind(beta_hat_ridge, beta_hat_ridge_alt)
```


So they are the same! But why did we do this? 

It is sometimes a bit more computationally efficient if we have to invert an 
$n \times n$ matrix instead of a $p \times p$ matrix, but there are faster 
methods if we're really interested in speed. 

Note that, if we want to make a ridge regression _prediction_ now for a new
$\tilde{\bx}$, we just need to compute: 

$$\newcommand{\bx}{\mathbf{x}}\begin{align*}
\hat{\beta}_{\text{Ridge}}^T\tilde{\bx} &= \left[\bX^{\top}(\bX\bX^{\top}+\lambda \bI)^{-1}\by\right]^T\tilde{\bx} \\
&= \left[\by^T(\bX\bX^{\top}+\lambda \bI)^{-T}\bX\right]\tilde{\bx} \\
&= \by^T(\bX\bX^{\top}+\lambda \bI)^{-1} \bX\tilde{\bx}
\end{align*}$$


Intuitively, recall that the inner product of two vectors measures the angle
between them and can, up to some scaling, be used as a 'similarity' measure. 
This result essentially says that our prediction on a new data set is a weighted
average of the training data, with weights based on similarity of the new point
with the training data. This is not a crazy structure...

If we look closer, we see that we **only** need to compute products of the form
$\bx_1\bx_2$ to make this work. This is where the magic of kernels lies.

### Kernel Trick

Earlier, we considered non-linear regression by _feature expansion_, where we
replaced $\bX$ by $\Phi(\bX)$, where $\Phi: \R^p \to \R^P$ maps to a
higher-dimensional space (applied row-wise). We can use this mapping in our new
ridge formula to get our non-linear predictions as 

$$\begin{align*}
\hat{y}(\tilde{\bx}) &= \by^T(\Phi(\bX)\Phi(\bX)^{\top}+\lambda \bI)^{-1} \Phi(\bX)\Phi(\tilde{\bx})
\end{align*}$$

Longer, but not necessarily better. If $P$ is very large, computing with 
$\Phi(X)$ can be *incredibly* expensive.

It turns out, however, that we never need to actually form $\Phi(X)$, we only 
need $\kappa(\bx_1, \bx_2) = \Phi(\bx_1)^T\Phi(\bx_2)$. If we can compute 
$\kappa$ directly, we never need to work with $\Phi$. 

Functions that allow this are called (Mercer) kernels and they are just a 
little bit magical.

If we let $\K = \Phi(X)\Phi(X)^{\top}$ be defined by
$\K_{ij} = \kappa(\bx_i, \bx_j)$, then we can write our feature-augmented ridge regression as: 

$$\hat{y}(\tilde{\bx}) = \by^T(\K+\lambda \bI)^{-1} \kappa(\bX, \tilde{\bx})$$

If we break this apart, we see that our predictions at the new point $\tilde{\bx}$ are essentially just weighted averages of our original observations $\by$, weighted by the similarity between the new point and our training points $\kappa(\bX, \cdot)$. This intuition is super important and we'll dig into it further below. 

Let's try out _Kernel(ized) Ridge Regression_:

```{r}
library(kernlab)

## Define a kernel function - we'll spell this out below.
rbf <- rbfdot(sigma = 0.05)

## calculate kernel matrix
K <- kernelMatrix(rbf, X)
dim(K)
```

Note that this is a $n \times n$ matrix - not a $p \times p$ matrix!

We can now use this to make predictions:

```{r}
krr <- function(x, lambda=1) {
    crossprod(y, solve(K + lambda * eye(n), kernelMatrix(rbf, X, x)))
}

krr_MSE <- mean((y - krr(X))^2)
print(krr_MSE)
```

This is _better_ than the OLS MSE: 

```{r}
mean(resid(lm(y ~ X))^2)
```

Why? How is this not a contradiction of everything said before. 

But how do we actually _do_ kernel multiplication? 

There are _many_ kernel functions in the world. The most common are defined by: 

- $\kappa(\bx_1, \bx_2) = \bx_1^{\top}\bx_2$. This is the linear kernel, 
  equivalent to non-kernel methods
- $\kappa(\bx_1, \bx_2) = (\bx_1^{\top}\bx_2 + c)^d$. This is the _polynomial
  kernel_, equivalent to fitting polynomial regression up to degree $d$ with all
  cross products. (The role of $c$ is tricky, but it essentially controls the
  relative weight of the higher and lower order terms)
- $\kappa(\bx_1, \bx_2) = e^{-\sigma^2\|\bx_1 - \bx_2\|^2}$. This is the 
  _squared exponential_ or _radial basis function_ (RBF) kernel; it is very 
  popular in spatial statistics, where it is closely related to a method known
  as [kriging](https://en.wikipedia.org/wiki/Kriging).

While the first two kernels give us a natural $\Phi(\cdot)$ mapping, the
$\Phi(\cdot)$ for the RBF kernel is actually _infinite dimensional_: it lets 
us fit function classes we could not fit without the kernel trick.

For even more kernels, see: 

- [David Duvenaud's Kernel Cookbook](https://www.cs.toronto.edu/~duvenaud/cookbook/)
- [GPML Section 4.2.1](http://gaussianprocess.org/gpml/chapters/RW.pdf)

Much of our current understanding of deep learning is based on specialized
kernel methods: 

- The [Neural Tangent Kernel](https://en.wikipedia.org/wiki/Neural_tangent_kernel)
- [Kernel Methods for Deep Learning](https://papers.nips.cc/paper_files/paper/2009/hash/5751ec3e9a4feab575962e78e006250d-Abstract.html)
- [Gradient Descent as a Kernel Machine](https://arxiv.org/abs/2012.00152) (This paper is a bit controversial)

[This blog](https://rajatvd.github.io/NTK/) is a particularly nice 
(but still somewhat technical) introduction to the Neural Tangent Kernel and 
its relationship to neural networks.

Spline theory has also found use to explain ReLU (piecewise linear activation)
neural networks: 

- [Mad Max: Spline Insights into Deep Learning](https://arxiv.org/abs/1805.06576)
- [Representer Theorems for Neural Networks and Ridge Splines](https://jmlr.csail.mit.edu/papers/v22/20-583.html)

We can also _kernelize_ some of the other methods we've studied in this course. 

Notably, we can kernelize $K$-Nearest Neighbors to get "KKNN" if we modify the
definition of distance used to be "kernel-ish." We first recall that we can
write distances solely in terms of inner products:  

$$\begin{align*}
\|\bx_1 - \bx_2\|_2^2 &= (\bx_1 - \bx_2)^T(\bx_1 - \bx_2) \\
&= \bx_1^T\bx_1 - \bx_2^T\bx_1 - \bx_1^T\bx_2 + \bx_2^T\bx_2 \\
&= \bx_1^T\bx_1  - 2\bx_1^T\bx_2 + \bx_2^T\bx_2 \\
\implies \text{Kernel Distances} &= \kappa(\bx_1, \bx_1) - 2\kappa(\bx_1, \bx_2) + \kappa(\bx_2, \bx_2)
\end{align*}$$

If we compute distances this way for a given $\kappa$, we get kernel KNN. This
is particularly nice when we have kernels that capture specific behaviors 
(*e.g.*, periodic or embedded categorical) that we can't really treat as 
Euclidean.

## Trade-Offs between Linear and Non-Linear Methods

Having developed some tools for non-linear regression, let's step back and ask
whether they are worthwhile. Recall our error decomposition: 

$$\text{MSE} = \text{Irreducible Error} + \text{Variance} + \text{Model Bias}^2 + \text{Estimation Bias}^2$$

We have already discussed how tools like ridge and lasso let us remove variance
at the cost of (estimation) bias. 

We argued that this was a worthwhile trade
when the problem had a lot of 'innate' variance, either from having
large noise variance ($\V[\epsilon] \gg 0$) or from having many features. 
In particular, use of some sort of shrinkage (ridge or lasso penalization) was
essential in the 'high-dimensional' case ($p > n$) where OLS had so much
variance that it was not even uniquely defined. We also argued that, as we 
got more and more data ($n \to \infty$), the variance took care of itself in 
the usual statistical way. 

These lessons generalize to the non-linear vs linear debate as well. Choosing
to use a linear model is itself a variance reducing choice - there are 'more'
curves than lines in some sense. If we restrict our attention to linear models
only, we are potentially accepting some Model Bias, again with a hope of reducing
variance to get overall better performance. As such, the same intuition as above
applies: linear models are preferred when variance is the primary concern, either
from noisy data or from small data; as we get more data, variance decreases
naturally, so going to non-linear models reduces bias, giving overall 
smaller error. 

In the non-linear context, the 'estimation bias / variance' trade-off remains, 
but modern tools like `mgcv` essentially handle this automatically for us. It
is of course still there, but `mgcv` has some [nifty
auto-tuning](https://doi.org/10.1080/01621459.2016.1180986) built-in.[^gambook]

[^gambook]: If you want to go much deeper into the practical use of additive
(spline) models, Simon Wood's book [Generalized Additive
Models](https://www.routledge.com/Generalized-Additive-Models-An-Introduction-with-R-Second-Edition/Wood/p/book/9781498728331)
is fantastic. 

## Other Topics in Regression

### Alternative Loss Functions

For most of this class, we have used MSE as our loss function, building on a foundation of least squares. 

Squared ($\ell_2^2$) error is a bit interesting: as an error gets _larger_, it counts even more. This makes OLS/MSE methods sensitive to outliers (in the same way the mean is), but it also implies that OLS won't work too hard to "over-minimize" small errors. 

We can consider some other loss functions: 
- $\ell_1$ error (absolute difference): robust to outliers, fits to the conditional _median_ instead of of the conditional _mean_. 

  The resulting _Median Absolute Deviation_ (MAD) regression no longer has a closed form, but can be solved quickly
  using tools like `CVX.`
  
- Huber loss: a blend of $\ell_2$ and $\ell_1$ error, which uses $\ell_2$ for small errors and $\ell_1$ for big errors. 
$$\text{HL}_{\delta}(x) = \begin{cases} \frac{1}{2}x^2 & |x| < \delta \\ \delta * (|x| - \delta/2) & |x| \geq \delta \end{cases}$$

  This is still convex, so we can solve it with `CVX.`
  
- $\epsilon$-insensitive: 
  $$L_{\epsilon}(x) = (|x| - \epsilon)_+ = \begin{cases} 0 & |x| < \epsilon \\ |x| - \epsilon & |x| \geq \epsilon \end{cases}$$
  This captures the idea of "close enough is good enough" with "close enough" being defined by $\epsilon$. It's a bit unnatural statistically, but relates to an important classification method we'll cover in a few weeks. 
  
  
### Multi-Task Regression

In some contexts, we may want to perform _multiple regressions_ at the same time. That is, we have a _vector_ of responses for each observation. The OLS generalization is straightforward: 
$$\newcommand{\bbeta}{\mathbf{\beta}}\newcommand{\bB}{\mathbf{B}}\newcommand{\bY}{\mathbf{Y}}\argmin_{\bB} \frac{1}{2}\|\bY - \bX\bB\|_F^2 \implies \hat{\bB} = (\bX^T\bX)^{-1}\bX^T\bY$$
so we essentially just fit different OLS models for each element of the response. This is not super interesting. 

Multi-task regression becomes interesting if we want to add structure to $\bB$: a particularly common requirement is to select the _same set of features_ for each of the regression targets. We do this using a group lasso on _each row_ of $\bB$: under this structure, if we select a feature for one part of the response, we'll use it for every part. (Can you see why?)

$$\argmin_{\bB} \frac{1}{2}\|\bY - \bX\bB\|_F^2 + \lambda \sum_{j=1}^p \|\bB_{j\cdot}\|_2$$

Another common requirement is for $\bB$ to be "low-rank" but we won't consider that here. If you want to look into it, keywords are "nuclear-norm regularized."

#### Weights and generalization

- WLS loss: $\newcommand{\bW}{\mathbf{W}}(\by - \bX\bbeta)^T\bW(\by - \bX\bbeta)$ for known diagonal $\bW$. 
  Weight samples based on their variance (higher variance gets smaller weights)
- Weighted lasso: $+\lambda\sum_{j=1}^p w_j|\beta_j|$. 
  Higher weights are less likely to be selected
  
  Basis of 'adaptive' lasso methods (not discussed)
- Generalized ridge: $+\frac{\lambda}{2}\bbeta^T\Omega\beta$

  Used to 'smooth' $\hat{\beta}$ - useful for ridge analogue of fused lasso

