---
session: "`r session <- 6; session`"
topic: "`r library(tidyverse); topic <- read_csv('../key_dates.csv', name_repair='universal') |> filter(Course.Element == 'Class Session', Item.Number==session) |> pull(Details); topic`"
date: "Last Updated `r Sys.Date()`"
---

{{< include _latex_defines.qmd >}}

## Ridge Regression

In the [previous set of notes](./notes05.qmd), we saw that
$(\overline{X}_n)_{[-\beta, \beta]}$ could outperform the naive estimator
$\overline{X}_n$ if the true parameter is 'not too big'. 

Can we extend this idea to regression? 

Recall that we formulated OLS as: 

$$\hat{\bbeta}_{\text{OLS}} = \argmin_{\bbeta \in \R} \frac{1}{2}\|\by - \bX\bbeta\|$$

We can apply a 'clamp' as: 

$$\hat{\bbeta}_{\tau-\text{Clamped}} = \argmin_{\bbeta \in \R: \|\bbeta\| < \tau} \frac{1}{2}\|\by - \bX\bbeta\| = \argmin_{\bbeta \in \R} \frac{1}{2}\|\by - \bX\bbeta\| \text{ such that } \|\bbeta\| \leq {\tau}$$

Here, we have some choice in measuring the 'size' of $\hat{\bbeta}$: in fact, we
can *theoretically* use any of our $\ell_p$-norms. As with the squared loss, 
it turns out to be mathematically easiest to start with the $\ell_2$ (Euclidean)
norm. This gives us the so-called *ridge regression* problem: 

$$\hat{\bbeta}_{\text{Ridge}} = \argmin_{\bbeta \in \R} \frac{1}{2}\|\by - \bX\bbeta\| \text{ such that } \|\bbeta\|_2 \leq {\tau} = \argmin_{\bbeta \in \R} \frac{1}{2}\|\by - \bX\bbeta\| \text{ such that } \|\bbeta\|_2^2 \leq {\tau}^2$$

Before we discuss solving this, let's confirm it is convex. Recall that for a
problem to be convex, it needs to have a convex *objective* and a convex *feasible
set*. Clearly the objective here is convex - we still have the MSE loss from OLS
unchanged. So now we need to convince ourselves that $\|\bbeta\|_2^2 \leq \tau^2$
defines a convex set.  We can look at this in 2D first: 

$$\|\bbeta\|_2^2 \leq \tau^2 \implies \beta_1^2 + \beta_2^2 \leq \tau$$

But this is just the equation defining a circle and its interior (in math speak
a 2D 'ball') so it has to be convex!

In fact, constraints of the form 

$$\|\bbeta\|_p \leq \tau \Leftrightarrow \|\bbeta\|_p^2 \leq \tau^p$$

are convex for all $\ell_p$ norms ($p \geq 1$). We will use this fact many times
in this course. The sets that this constraint defines are called $\ell_p$ 'balls'
by analogy with the $\ell_2$ figure, even if they aren't actually round.

```{r}
#| echo: false
#| fig.asp: 0.3
library(patchwork)
plot_ball <- function(p){
    title <- bquote(l[.(p)]~"Unit Ball")
    expand_grid(x=seq(-2, 2, length.out=201), 
                y=seq(-2, 2, length.out=201)) |>
        mutate(interior = abs(x)^p + abs(y)^p <= 1) |>
        filter(interior) |>
        ggplot(aes(x=x, y=y)) + 
        geom_point() + 
        coord_fixed() + 
        ggtitle(title) + 
        theme_bw() + 
        xlab("") + ylab("")
}
plot_ball(1) + plot_ball(2) + plot_ball(Inf)
```

So our ridge problem 

$$\hat{\bbeta}_{\text{Ridge}} = \argmin_{\bbeta \in \R} \frac{1}{2}\|\by - \bX\bbeta\| \text{ such that } \|\bbeta\|_2^2 \leq {\tau}^2$$
is indeed convex. 

While we can solve this without too much trouble (though we need to adjust
our basic gradient descent method to avoid going 'out of bounds'), it turns
out to be easier to use an optimization trick known as *Lagrange Multipliers* 
to change this to something easier to solve. 

The [Method of Lagrange Multipliers](https://en.wikipedia.org/wiki/Lagrange_multiplier)
lets us turn *constrained* optimization problems that look like: 

$$\argmin_x f(x) \text{ such that } g(x) \leq \tau$$

into *unconstrained* problems like: 

$$\argmin_x f(x) + \lambda g(x)$$

Here, the constant $\lambda$ is known as the Lagrange multiplier. Instead of 
_forcing_ $g(x)$ to be bounded, we simply _penalize_ any $x$ that makes $g(\cdot)$
large. For large enough penalties (large $\lambda$), we will eventually force
$g(x)$ to be small enough that we satisfy the original constraint. In fact, 
there is a one-to-one relationship between $\tau$ and $\lambda$, but it's usually
not possible to work it out in any meaningful useful manner; all we can say is that
larger $\lambda$ correspond to smaller $\tau$. 

::: {.callout-note title="Pause to Reflect" collapse="true"}

Why is this true? Why does small $\tau$ imply large $\lambda$ and *vice versa*?

:::

In this form, it's straightforward to pose ridge regression as: 

$$\argmin_{\beta} \frac{1}{2}\|\by - \bX\bbeta\|_2^2 + \frac{\lambda}{2}\|\bbeta\|_2^2$$

::: {.callout-note title="Pause to Reflect" collapse="true"}

Why were we allowed to put an extra $\frac{1}{2}$ in front of the penalty term?

:::

Since both terms of this expression are differentiable, it is easy to show 
that the solution to this problem is given by: 

$$\hat{\bbeta}_{\lambda} = (\bX^{\top}\bX + \lambda \bI)^{-1}\bX^{\top}\by$$

To wit: 

$$\begin{align*}
\bzero &= \frac{\partial}{\partial \bbeta}\left(\frac{1}{2}\|\by - \bX\bbeta\|_2^2 + \frac{\lambda}{2}\|\bbeta\|_2^2\right) \\
       &= -\bX^{\top}(\by - \bX\bbeta) + \lambda \bbeta \\
\bX^{\top}\by &= (\bX^{\top}\bX + \lambda \bI)\bbeta \\
\implies \hat{\beta}_{\text{Ridge}} &= (\bX^{\top}\bX + \lambda \bI)^{-1}\bX^{\top}\by
\end{align*}$$

There are several important things to note about this expression: 

- There are actually *many* ridge regression solutions, one for each value of
  $\lambda$. It is a bit improper to refer to "the" ridge regression solution. 
  We will discuss selecting $\lambda$ below. 
- If you are a bit sloppy, this looks something like: 

  $$\hat{\beta} \approx \frac{SS_{XY}}{SS_{XX} + \lambda}$$
  
  so we 'shrink' the standard OLS solution towards zero, by an amount depending 
  on $\lambda$. Higher $\lambda$ 
- The penalized form (as opposed to the constraint form) allows for $\hat{\beta}$
  to be arbitrarily large, **if the data supports it.**
  
  More than that, $\lambda$ gives us an explicit *trade-off*. When comparing
  two potential solutions ($\hat{\bbeta}_1, \hat{\bbeta}_2$), we will prefer
  $\hat{\bbeta}_1$ if: 
  
  $$\begin{align*}
  \frac{1}{2}\|\by - \bX\bbeta_1\|_2^2 + \frac{\lambda}{2}\|\bbeta_1\|_2^2 & \leq \frac{1}{2}\|\by - \bX\bbeta_2\|_2^2 + \frac{\lambda}{2}\|\bbeta_2\|_2^2 \\
  \|\by - \bX\bbeta_1\|_2^2 - \|\by - \bX\bbeta_2\|_2^2 & \leq \lambda(\|\bbeta_2\|_2^2 - \|\bbeta_1\|_2^2) \\
  \Delta \MSE &\leq \lambda \Delta \text{$\ell_2^2$-Norm}
  \end{align*}$$
  
  That is, we will take a solution with a MSE that is $\lambda$ lower if that lowers
  the squared-norm of the solution by 1. Here $\lambda$ measures the "price" we
  are willing to pay (in terms of MSE) for shrinking the norm of $\hat{\bbeta}$:
  at larger $\lambda$, we demand a *larger* MSE reduction *ceteribus paribus*
  to accept a larger $\hat{\bbeta}$, while at small $\lambda$, we are willing
  to take a larger $\hat{\bbeta}$ in exchange for a much smaller decrease in MSE.
  
- If we set $\lambda = 0$, we recover standard OLS. Continuing the 'price'
  discussion above, this is the limit in which we will take a larger $\hat{\bbeta}$
  for *any* decrease in training MSE. This makes sense as we originally built
  OLS as the ERM for MSE, by definition caring about *only* MSE reduction and
  nothing else. 
- Unlike the OLS solution, the ridge regression exists even when $p > n$ because
  the $+\lambda \bI$ term ensures the 'denominator' prefactor is always 
  invertible.
  
While this isn't as simple to analyze as some of the expressions we considered 
above, this looks similar enough to our 'shrunk' mean estimator that it's 
plausible it will improve on OLS. In fact, you can show that there is always 
_some_ value of $\lambda$ that guarantees improvement over OLS in expectation. 
(This is a result known as the "MSE Existence Theorem")

To wit, 

```{r}
#| warning: false
n <- 100
p <- 80
Z <- matrix(rnorm(n * p), nrow=n)
L <- chol(toeplitz(0.6^(1:p))) # Add 'AR(1)' correlation
X <- Z %*% L
beta <- runif(p, 2, 3)

eye <- function(n) diag(1, n, n)
calculate_ridge_mse <- function(lambda, nreps = 1000){
    MSE <- mean(replicate(nreps, {
        y <- X %*% beta + rnorm(n, sd=1)
        beta_hat <- solve(crossprod(X) + lambda * eye(p), crossprod(X, y))
        sum((beta - beta_hat)^2)
    }))
    
    data.frame(lambda=lambda, MSE=MSE)
}

lambda_grid <- 10^(seq(-2, 2, length.out=41))

RIDGE_MSE <- map(lambda_grid, calculate_ridge_mse) |> list_rbind()
OLS_MSE <- calculate_ridge_mse(0)$MSE # Ridge at lambda = 0 => OLS

ggplot(RIDGE_MSE, aes(x=lambda, y=MSE)) + 
    geom_point() + 
    geom_line() + 
    geom_abline(intercept=OLS_MSE, slope=0, lwd=2, lty=2) + 
    xlab(expression(lambda)) + 
    ylab("Estimation Error (MSE)") + 
    ggtitle("Ridge Regularization Improves on OLS") + 
    theme_bw() + 
    scale_x_log10()
```

Clearly, for smallish $\lambda$, we are actually outperforming OLS, sometimes
by a significant amount! For this problem, the minimum MSE (best estimate) actually
seems to occur near $\lambda \approx$ `r RIDGE_MSE |> filter(MSE == min(MSE)) |> pull(lambda) |> round(3)`

### Model Selection

While theory and our experiment above tell us that there is _some_ value of
$\lambda$ that beats OLS, how can we actually find it? We can't do a simulation
like above as we had to know $\bbeta_*$ to compute the MSE and the theory is
non-constructive. (In particular, we have to know $\bbeta_*$ to compute the 
bias.) 

Well, we can go back to our overarching goal: we want results that _generalize_
to unseen data. Here, in particular, we want results that _predict well_ on
unseen data. So let's just do that. 

We can split our data into a _training_ set and a _validation_ set. The _training_
set is like we have already seen, used to estimate the regression coefficients
$\bbeta_{\text{Ridge}}$; the _validation_ set may be new. It is used to compute
the MSE on 'pseudo-new' data and we then can select the best predicting value of
$\lambda$. In this case, this looks something like the following: 

```{r}
# Continuing with X, beta from above
y <- X %*% beta + rnorm(n, sd = 1)

TRAIN_IX <- sample(n, 0.8 * n)
VALID_IX <- setdiff(seq(n), TRAIN_IX) # Other elements

TRAIN_X <- X[TRAIN_IX, ]
TRAIN_Y <- y[TRAIN_IX]

VALID_X <- X[VALID_IX, ]
VALID_Y <- y[VALID_IX]

compute_validation_error <- function(lambda){
    beta_hat_rr <- solve(crossprod(TRAIN_X) + lambda * eye(p), 
                         crossprod(TRAIN_X, TRAIN_Y))
    
    y_pred <- VALID_X %*% beta_hat_rr
    
    data.frame(lambda = lambda, 
               validation_mse = mean((y_pred - VALID_Y)^2))
}

validation_error <- map(lambda_grid, compute_validation_error) |> list_rbind()

ggplot(validation_error, 
       aes(x = lambda, 
           y = validation_mse)) + 
    geom_point() + 
    xlab(expression(lambda)) + 
    ylab("MSE on Validation Set") + 
    ggtitle("Hold-Out Tuning of Ridge Regression Regularization Parameter") + 
    theme_bw() + 
    scale_x_log10()
```

Our results here aren't quite as definitive as when we knew the true $\bbeta_*$, 
but they suggest we want to take $\lambda \approx$ 
`r validation_error |> filter(validation_mse == min(validation_mse)) |> pull(lambda) |> round(3)`
which isn't too far from what we found above. 

::: {.callout-note title="Pause to Reflect" collapse="true"}

Does this procedure give an unbiased estimate of the true test error? Why or 
why not? 

:::

There are some limitations to this basic procedure - randomness and (arguably)
inefficient data usage - that we will address later, but this gets us started. 

## Lasso Regression

Ridge regression is a rather remarkable solution to the problem of 'inflated' 
regression coefficients. But let's remind ourselves why we find inflated
coefficients in the first place. 

A common cause of overly large regression coefficients is correlation among the
columns of $\bX$. For example, 

```{r}
max_beta_sim <- function(rho, n, p){
    R <- replicate(100, {
        Z <- matrix(rnorm(n * p), nrow=n)
        L <- chol(toeplitz(rho^(1:p)))
        X <- Z %*% L
        beta <- runif(p, 2, 3)
        y <- X %*% beta + rnorm(n)
        
        beta_hat <- coef(lm(y ~ X))
        max(abs(beta_hat))
    })
    data.frame(beta_max = max(R),
               n = n, 
               p = p, 
               rho = rho)
}

RHO_GRID <- seq(0.5, 0.999, length.out=101)

BETAS <- map(RHO_GRID, max_beta_sim, n = 40, p = 35) |> list_rbind()

ggplot(BETAS, 
       aes(x = rho, 
           y = beta_max))+ 
    geom_point() + 
    scale_y_log10() + 
    theme_bw() + 
    xlab("Feature Correlation") + 
    ylab("Maximum Observed Regression Coefficient") + 
    geom_abline(intercept = 3, color="red4", lwd=2, lty=2)
```

This simulation is maybe a bit unfair to OLS since we are taking maxima instead
of averages or similar, but the message is certainly clear. As the columns of
$\bX$ become highly correlated, we get larger values of $\hat{\beta}$. Because
OLS is unbiased here, this is fundamentally a story of *variance*: as we have 
more features, 

$$\V[\hat{\beta}] \propto (\bX^{\top}\bX)^{-1}$$

increases, making it likely that we see increasingly wild values of $\hat{\beta}$. 
(Recall that we generated our data so that $\beta_*$ was never larger than 3, as 
noted in red above.)

Why does this happen? It's maybe a bit tricky to explain, but with highly
correlated features the model can't really distinguish them well, so it 
'overreacts' to noise and produces very large swings (variance) in response
to small changes in the training data. Like above, we can hope to tame some of
this variance if we take on a bit of bias. 

Another way we might want to 'calm things down' and improve on OLS is by 
having fewer features in our model. This reduces the chance of 'accidental 
correlation' (more features leads to more 'just because' correlations) and gives
us a more interpretable model overall. (More about interpretability below)

### Best Subsets

You have likely already seen some sort of 'variable selection' procedures in 
prior courses: things like _forward stepwise_, _backwards stepwise_, or even 
_hybrid stepwise_. All of these are trying to get at the idea of only including
variables that are 'worth it'; in other settings, you may also see them used as
finding the 'most valuable' variables. Let's use our newfound knowledge of 
optimization to formalize this. 

The _best-subsets_ problem is defined by: 

$$\argmin_{\beta} \frac{1}{2}\|\by - \bX\bbeta\|_2^2 \text{ such that } \|\bbeta\|_0 \leq k$$

This looks similar to our ridge regression problem, but now we've replaced the
$\ell_2$-norm constraint with the $\ell_0$-'norm'. Recall that the 
$\ell_0$-'norm' is the number of non-zero elements in $\bbeta$ and that it's not
a real norm. So this says, find the minimum MSE $\beta$ with at most $k$ non-zero
elements. Since more features always reduces (or at least never increases) training
MSE, we can assume that this problem will essentially always pick the best 
_combination_ of $k$ variables. 

Alternatively, in penalized form, we might write: 

$$\argmin_{\beta} \|\by - \bX\bbeta\|_2^2 + \lambda \|\bbeta\|_0$$

Here, we only introduce a new variable if including it lowers our training MSE
by $\lambda$: we've implicitly set $\lambda$ as our 'worth it' threshold for
variables. 

Note that these problems require finding the best _combination_ of variables. 
The most individually useful variables might not be useful in combination: 
*e.g.*, someone's right shoe size might be very useful for predicting their height, 
and their left shoe size might be very useful for predicting their height, and their
gender might be a little bit useful for predicting their height, but you'll do better
with right shoe and gender than with right and left shoe. 

Because we're checking all combinations of variables, this is a so-called 
'combinatorial' optimization problem. These are generally quite **hard** to solve
and require specialized algorithms (unlike our general purpose convex optimization
algorithms). A naive approach of 'try all models' becomes impractical quickly. 
If we have $p=30$ potential features, that's $2^{30}$, or just over a billion,
possible models we would need to check. Even if we can check one model every
minute, that takes about 2040 **years** to check them all; if we have a computer
do the work for us and it's faster, say 1000 models per minute, we still need
over 2 years. And that's just for 30 features! In realistic problems where $p$ 
can be in the hundreds or thousands, we have no chance. 

Thankfully, smart people have worked on this problem for us and found that for
$p \approx 100 - 1000$, very fancy software can solve this problem.[^bertsimas]
But that still leaves us very far we want to go...

[^bertsimas]: "Best subset selection via a modern optimization lens" by 
Dimitris Bertsimas, Angela King, Rahul Mazumder
*Annals of Statististics* **44(2)**: 813-852 (April 2016). DOI: 
[10.1214/15-AOS1388](https://dx.doi.org/10.1214/15-AOS1388)

### Algorithmic Approximations

### Convex Approximations

## The Lasso

### The Value of Sparsity

### Algorithms

### glmnet

### Model Selection

TODO
